---
title: "Chapter 4: Results"
author: "Elaf"
format:
  html: default
  docx: default
bibliography: references.bib
execute:
  echo: false
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)

# The solutions folder is one level below the data
students  <- read.csv("../students.csv")
semesters <- read.csv("../semesters.csv")
first_semester <- semesters |> filter(semester == 1)
```

## Sample

The sample included `r nrow(students)` graduate students from
`r n_distinct(students$faculty)` faculties. Their average wellbeing in the first
semester was `r sprintf("%.1f", mean(first_semester$wellbeing, na.rm = TRUE))` on the
wellbeing index (0 to 100). @tbl-faculty shows wellbeing by faculty.

```{r}
#| label: tbl-faculty
#| tbl-cap: "Wellbeing in the first semester, by faculty."
first_semester |>
  left_join(students, join_by(student_id)) |>
  summarise(Students = n(),
            Mean = mean(wellbeing, na.rm = TRUE),
            SD = sd(wellbeing, na.rm = TRUE),
            .by = faculty) |>
  arrange(faculty) |>
  knitr::kable(digits = 1, col.names = c("Faculty", "Students", "Mean", "SD"))
```

## Change over the programme

Wellbeing changed over the four semesters of the programme (@fig-wellbeing).
Differences are described with standardised effect sizes [@cohen1988], and all
results can be reproduced from the data and this document [@peng2011].

```{r}
#| label: fig-wellbeing
#| fig-cap: "Average wellbeing by semester."
#| fig-alt: "Line chart of average wellbeing over semesters 1 to 4."
semesters |>
  summarise(wellbeing = mean(wellbeing, na.rm = TRUE), .by = semester) |>
  ggplot(aes(x = semester, y = wellbeing)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Semester", y = "Average wellbeing (0 to 100)") +
  theme_minimal()
```

## References
