# Installation and Setup {#sec-app-installation}

```{r}
#| label: setup
#| include: false
library(data2thesis)
```

This appendix explains how to install everything the book uses, and how to fix the problems that most often get in the way. If you only want to start, the short version is: install **R**, then **RStudio**, then run the package installation command in @sec-install-packages. If you are not ready to install anything yet, every chapter's exercises also run in your web browser, in the [playground](../playground/index.html).

Installers and websites change their appearance from year to year, so this appendix describes each step in words rather than with screenshots. The choices that matter are the same in every version.

## What you need

| Software | What it is | Required? |
|---|---|---|
| **R** | The programming language and the engine that runs your code | Yes |
| **RStudio Desktop** | The program you work in: editor, console, plots, and help in one window | Yes (or Positron) |
| **Positron** | A newer editor from the makers of RStudio, an alternative to it | Optional |
| **Quarto** | Turns documents with code into reports and books (Chapter 17) | Comes with RStudio |
| **R packages** | Add-ons for specific tasks, installed from within R | Yes, as needed |
| **Rtools** (Windows) or **Xcode command line tools** (macOS) | Compilers for building packages from source | Only if asked |

: The software used in this book {#tbl-software}

All of it is free. R must be installed first, because RStudio and Positron are programs *for* R: they need R to run your code. Install R, then the editor.

## Installing R

R is downloaded from **CRAN**, the Comprehensive R Archive Network, at [cran.r-project.org](https://cran.r-project.org). Always download the latest release. This book was built with R `r paste(R.version$major, R.version$minor, sep = ".")`; any later version works.

### Windows

1. On the CRAN page, choose *Download R for Windows*, then *base*, then the link to download the latest version (a file such as `R-4.x.y-win.exe`).
2. Run the downloaded file. If Windows asks whether to allow the app to make changes, choose *Yes*. If you have no administrator rights on a university computer, the installer offers to install for your user only, which works just as well.
3. Accept the default settings on every screen. The defaults install R in `C:\Program Files\R\` and register it so that RStudio finds it automatically.

### macOS

1. On the CRAN page, choose *Download R for macOS*.
2. Choose the right package for your Mac's processor. Newer Macs (from late 2020) have **Apple silicon** (M1, M2, and later): use the file marked *arm64*. Older Macs have **Intel** processors: use the file marked *x86_64*. The Apple menu, *About This Mac*, shows which you have.
3. Open the downloaded `.pkg` file and follow the installer, accepting the defaults.

### Linux

R is available in the package manager of every major Linux distribution, but the version there is often out of date. CRAN's *Download R for Linux* page gives up-to-date instructions for Ubuntu, Debian, Fedora, and others. On Ubuntu, for example, the instructions add CRAN's repository and then install R with `sudo apt install r-base r-base-dev`. Follow the page for your distribution, because the exact commands change with each release.

### Checking the installation

Open R (from the Start menu on Windows or the Applications folder on macOS) and type:

```{r}
R.version.string
```

If it prints a version number, R is working. You will rarely open R on its own again; from now on you will work in RStudio.

## Installing RStudio

1. Go to [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/). The page detects your operating system; step 1 on the page (installing R) is already done.
2. Download **RStudio Desktop** (the free version) and install it like any other program, accepting the defaults. On macOS, drag RStudio into the Applications folder.
3. Open RStudio. It finds R automatically and shows the R version in the Console pane, in the lower left (Chapter 1 takes you on a tour of the panes).

RStudio includes **Quarto**, so the reports of Chapter 17 work without installing anything else. For PDF output, Quarto also needs LaTeX; install a small version once by typing `quarto install tinytex` in RStudio's **Terminal** tab.

### Positron, an alternative

**Positron** ([positron.posit.co](https://positron.posit.co)) is a newer editor from Posit, the company behind RStudio, built for both R and Python. Everything in this book works in Positron as well. RStudio is used in the book's instructions because it is still the most widely used, and most tutorials and university courses assume it.

### No installation at all

If you cannot install software, for example on a locked-down computer, two options need only a web browser:

- The book's [playground](../playground/index.html) runs R in your browser, with exercises for every chapter.
- **Posit Cloud** ([posit.cloud](https://posit.cloud)) offers RStudio in the browser, with a free plan that is limited but enough to follow most chapters.

## Recommended RStudio settings

A few settings make RStudio work the way this book recommends. Open *Tools > Global Options*:

- **General > Basic**: untick *Restore .RData into workspace at startup*, and set *Save workspace to .RData on exit* to **Never**. Every session then starts clean, and your scripts, not a hidden saved workspace, hold your work. This is the single most important setting for reproducible work (Chapter 17).
- **Code > Editing**: tick *Use native pipe operator, \|>*, so that the pipe shortcut (Ctrl+Shift+M, or Cmd+Shift+M on a Mac) inserts the pipe used in this book (Chapter 3).
- **Code > Display**: tick *Show margin* and *Highlight R function calls*, which make code easier to read.
- **Appearance**: choose a font size and theme that are comfortable for you; dark themes are popular for long sessions.

## Installing the book's packages {#sec-install-packages}

Packages are installed once, with `install.packages()`, and loaded in every session with `library()` (Chapter 1). Each chapter lists the packages it uses at the start; @tbl-packages collects them all.

| Chapters | Packages |
|---|---|
| 1-3 | here, readxl, haven, writexl, dplyr, tidyr, stringr, readr (or the whole tidyverse) |
| 4, 6-7 | ggplot2, psych |
| 8 | broom, car |
| 9 | psych, GPArotation, corrplot, factoextra |
| 10 | lme4, lmerTest |
| 11-13 | tidymodels, kknn, ranger, kernlab, themis, glmnet, xgboost |
| 14 | mclust, dbscan, factoextra, cluster |
| 15 | tidymodels (nnet comes with R) |
| 16 | tsibble, fable, feasts, urca |
| 17 | rmarkdown, knitr, shiny, bslib, renv, usethis |
| 18 | ellmer, stringr, yardstick |
| 19 | patchwork, broom, lme4 |

: Packages used in each chapter {#tbl-packages}

To install everything at once, copy this command into the Console. It downloads a few hundred megabytes and can take ten minutes or more, so it is best done on a good connection:

```r
install.packages(c(
  "tidyverse", "here", "readxl", "haven", "writexl",
  "psych", "GPArotation", "corrplot", "factoextra", "broom", "car",
  "lme4", "lmerTest",
  "tidymodels", "kknn", "ranger", "kernlab", "themis", "glmnet", "xgboost",
  "mclust", "dbscan", "patchwork",
  "tsibble", "fable", "feasts", "urca",
  "rmarkdown", "shiny", "bslib", "renv", "usethis",
  "ellmer"
))
```

The **tidyverse** package installs dplyr, tidyr, stringr, readr, ggplot2, and several others in one go. The first time you install packages, R may ask you to choose a **CRAN mirror** (choose *0-Cloud*, which is fast everywhere) and whether to use a personal library (answer *Yes*).

### The student wellbeing data: the data2thesis package

Elaf's data is in the **data2thesis** package, which is installed from this book's website rather than from CRAN:

```r
install.packages("https://polla-fattah.github.io/data2thesis_r/downloads/data2thesis_1.1.0.tar.gz",
                 repos = NULL, type = "source")
```

The package contains only data, so it installs on every system without compiling anything. Check that it works:

```{r}
library(data2thesis)
nrow(students)
```

The same data is also available as ordinary files (CSV, Excel, and SPSS) on the book's website, for readers who prefer to import files, as Chapter 2 does.

### The versions used in this book

The results in this book were produced with these versions. Newer versions usually give the same results, but if a number in your output differs slightly from the book, a different package version is a likely reason.

```{r}
#| echo: false
key_packages <- c("dplyr", "tidyr", "ggplot2", "lme4", "tidymodels", "fable", "ellmer", "data2thesis")
tibble::tibble(
  Software = c("R", key_packages),
  Version  = c(paste(R.version$major, R.version$minor, sep = "."),
               vapply(key_packages, \(p) as.character(packageVersion(p)), character(1)))
) |>
  knitr::kable()
```

## Common problems and solutions

**"There is no package called ..."**
: The package is not installed, or the name is misspelled (names are case-sensitive: `GPArotation`, not `gparotation`). Install it with `install.packages("name")`, then load it with `library(name)`.

**"Package ... is not available for this version of R"**
: Either the name is misspelled, or your R is too old for the current version of the package. Update R (see below). A few packages are not on CRAN at all; their documentation says how to install them.

**"Do you want to install from sources the package which needs compilation?"**
: A newer version exists as source code than as a ready-made (binary) package. Answer *No* to install the slightly older binary version, which is fine for this book. Answering *Yes* needs **Rtools** on Windows (from CRAN's *Download R for Windows* page, choosing the Rtools version that matches your R) or the **Xcode command line tools** on macOS (run `xcode-select --install` in the macOS Terminal).

**"Lib ... is not writable" or permission errors**
: R cannot write to the system's package folder, which is common on university computers. When R offers to create a personal library, answer *Yes*.

**Problems with folder names on Windows**
: If your Windows user name or project path contains non-English characters, or your files are in a synchronised folder such as OneDrive, installation and file reading can fail in confusing ways. Keeping R projects in a simple local folder, such as `C:\research\`, avoids most of these problems.

**Downloads fail at the university**
: Some networks block or slow down downloads. Try the *0-Cloud* mirror (`chooseCRANmirror()`), another network, or ask your IT service whether a proxy must be set.

**RStudio cannot find R**
: Install R before RStudio, or reinstall R with the default settings. On Windows, *Tools > Global Options > General* lets you choose the R version RStudio uses.

**Everything worked yesterday, and today nothing does**
: Restart R (*Session > Restart R*), then run your script from the top. Many problems come from objects left over from earlier work, which a clean start removes (see the settings above).

## Keeping R up to date

Update your packages every few months with `update.packages()`, or the *Update* button in RStudio's Packages pane. Update R itself once or twice a year by installing the new version exactly as the first time; your scripts keep working, but packages must be reinstalled for the new version, which the command in @sec-install-packages does in one go. Tools such as **rig** ([github.com/r-lib/rig](https://github.com/r-lib/rig)) can install and switch between several R versions, useful if an old project needs an old version. For projects that must keep exactly the same package versions, use **renv** (Chapter 17).

Two things do not belong in any script. API keys for AI services (Chapter 18) go in your personal `.Renviron` file, opened with `usethis::edit_r_environ()`; and passwords never go into code at all.
