---
number-sections: false
---

# Appendix B: Glossary {#sec-app-glossary}

```{r}
#| label: setup
#| include: false
# The glossary is built from two plain-text files next to this one:
#   glossary.txt   term | aliases :: definition
#   functions.txt  group | function | package | description
# Chapter numbers are found automatically: for terms, from each chapter's
# "Key terms" list; for functions, from the code chunks of each chapter.

chapter_files <- sort(list.files(pattern = "^[0-9]{2}-.*\\.qmd$"))
chapter_number <- as.integer(substr(chapter_files, 1, 2))
chapter_text <- lapply(chapter_files, readLines, encoding = "UTF-8", warn = FALSE)

# Key terms of every chapter
key_terms <- lapply(chapter_text, function(x) {
  start <- which(x == "### Key terms")
  if (length(start) == 0) return(character(0))
  line <- x[start + 2]
  terms <- trimws(strsplit(sub("\\.$", "", line), ",")[[1]])
  tolower(terms)
})

# Glossary entries
lines <- readLines("glossary.txt", encoding = "UTF-8", warn = FALSE)
lines <- lines[nzchar(trimws(lines)) & !startsWith(lines, "#")]
parts <- strsplit(lines, " :: ", fixed = TRUE)
names_part <- vapply(parts, `[`, character(1), 1)
definitions <- vapply(parts, `[`, character(1), 2)
all_names <- lapply(strsplit(names_part, " | ", fixed = TRUE), trimws)
display <- vapply(all_names, `[`, character(1), 1)

chapters_for <- function(names) {
  hit <- vapply(key_terms, function(k) any(tolower(names) %in% k), logical(1))
  chapter_number[hit]
}
entry_chapters <- lapply(all_names, chapters_for)

# Every key term in the book must have a glossary entry
covered <- tolower(unlist(all_names))
missing <- setdiff(unique(unlist(key_terms)), covered)
if (length(missing) > 0) stop("Key terms without a glossary entry: ", paste(missing, collapse = "; "))

sort_key <- tolower(gsub("[^A-Za-z0-9 ]", "", display))
first_letter <- toupper(substr(sort_key, 1, 1))
ord <- order(sort_key)
```

This appendix collects the key terms from every chapter's review, with a short definition of each and the chapters where it is introduced or used, followed by a table of the R functions used most often in the book. Definitions are written in plain language; the chapters give the full explanations and examples.

## Key terms

```{r}
#| echo: false
#| results: asis
current <- ""
for (i in ord) {
  if (first_letter[i] != current) {
    current <- first_letter[i]
    cat("\n### ", current, " {.unnumbered}\n\n", sep = "")
  }
  ch <- entry_chapters[[i]]
  where <- if (length(ch) == 0) "" else
    paste0(" (", if (length(ch) == 1) "Chapter " else "Chapters ", paste(ch, collapse = ", "), ")")
  aliases <- all_names[[i]][-1]
  aliases <- aliases[tolower(gsub("`", "", aliases)) != tolower(gsub("`", "", display[i]))]
  also <- if (length(aliases) > 0) paste0(" Also: ", paste(aliases, collapse = ", "), ".") else ""
  cat("**", display[i], "**\n: ", definitions[i], also, where, "\n\n", sep = "")
}
```

## R functions

@tbl-functions lists the functions used most often in the book, grouped by task, with the package they come from and the chapters that use them. `base R` functions are available without loading any package. For any function, `?name` in the Console opens its help page.

```{r}
#| label: tbl-functions
#| tbl-cap: "Frequently used R functions."
#| echo: false
functions <- read.delim("functions.txt", sep = "|", header = FALSE, comment.char = "#",
                        strip.white = TRUE, col.names = c("Task", "Function", "Package", "What it does"),
                        check.names = FALSE)

code_lines <- lapply(chapter_text, function(x) {
  # keep only the lines inside R code blocks: run (```{r}) and shown only (```r)
  in_chunk <- FALSE
  keep <- logical(length(x))
  for (j in seq_along(x)) {
    if (!in_chunk && grepl("^```(\\{r|r\\s*$)", x[j])) { in_chunk <- TRUE; next }
    if (in_chunk && grepl("^```\\s*$", x[j])) { in_chunk <- FALSE; next }
    keep[j] <- in_chunk
  }
  x[keep]
})

functions$Chapters <- vapply(functions$Function, function(f) {
  pattern <- paste0("(^|[^A-Za-z0-9_.])", gsub(".", "\\.", f, fixed = TRUE), "\\(")
  hit <- vapply(code_lines, function(x) any(grepl(pattern, x)), logical(1))
  ch <- chapter_number[hit]
  if (length(ch) > 4) paste0(paste(ch[1:3], collapse = ", "), ", and later") else paste(ch, collapse = ", ")
}, character(1))

functions$Function <- paste0("`", functions$Function, "()`")
knitr::kable(functions)
```
