---
title: "Data Visualization"
description: "Chapter 4: what graphs are for, how people read them, choosing a graph, the grammar of graphics with ggplot2, publication-ready and honest graphs, and exporting figures."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_deck <- theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 15))
theme_set(theme_deck)
update_geom_defaults("point", list(colour = navy))
update_geom_defaults("bar", list(fill = navy))
update_geom_defaults("col", list(fill = navy))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      fig.width = 6, fig.height = 4, message = FALSE, warning = FALSE)

first_sem <- semesters |> filter(semester == 1) |> left_join(students, join_by(student_id))
wellbeing_by_sem <- semesters |>
  left_join(students, join_by(student_id)) |>
  summarise(mean_wellbeing = mean(wellbeing), .by = c(semester, workshop))
faculty_means <- first_sem |> summarise(wellbeing = mean(wellbeing), .by = faculty)
anscombe_long <- tibble(
  set = rep(paste("Set", 1:4), each = 11),
  x   = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y   = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
)
```

# Data Visualization

A graph is an instrument and an argument

**Chapter 4**

Polla Fattah

---

## By the end of today you can

- explain why graphs are needed alongside numerical summaries;
- tell exploratory graphs from explanatory ones;
- explain why position and length are read more accurately than angle, area, or colour;
- choose a graph from the question and the variables;
- build graphs with ggplot2 from data, mappings, and layers;
- judge whether a graph is honest, and improve a misleading one;
- export figures at the size and resolution a thesis needs.

---

## Summaries can hide almost anything

```r
anscombe_long |>
  summarise(mean_x = mean(x), mean_y = mean(y),
            sd_y = sd(y), correlation = cor(x, y), .by = set)
```

| Set | mean x | mean y | sd y | correlation |
|---|---|---|---|---|
| 1 to 4 | 9 | 7.5 | 2.03 | 0.82 |

Anscombe's four datasets share every summary.

---

## Four identical summaries, four different stories

```{r}
#| fig-width: 9
#| fig-height: 3.6
ggplot(anscombe_long, aes(x = x, y = y)) +
  geom_point(size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, colour = copper, linewidth = 0.9) +
  facet_wrap(~ set, nrow = 1)
```

A line, a curve, an outlier, and one point making the whole trend.

---

## Look first

> **Anyone who analysed these data without looking would draw the same, wrong, conclusion from all four.**

Many errors in data, and in reasoning about data, are found only by looking.

---

## Exploring and explaining

| | Exploratory graph | Explanatory graph |
|---|---|---|
| for | the researcher | the reader |
| purpose | understand the data | show a known finding |
| number | many, quickly | a few, carefully |
| polish | none needed | clear message and labels |

A thesis contains a few explanatory graphs, chosen from many exploratory ones.

---

## A graph turns numbers into visual properties

```{mermaid}
flowchart LR
    A["Position on a<br>common scale"] --> B["Length"] --> C["Angle<br>and slope"] --> D["Area"] --> E["Colour<br>shade"]
```

Cleveland and McGill's experiments: judged most accurately on the left, least on the right.

Put the most important comparison into position or length.

---

## The same five numbers, two ways

```{r}
#| fig-width: 9
#| fig-height: 3.6
shares <- tibble(group = LETTERS[1:5], percent = c(23, 21, 20, 19, 17))
pie_chart <- ggplot(shares, aes(x = "", y = percent, fill = group)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  scale_fill_viridis_d(end = 0.9) +
  theme_void(base_size = 15)
bar_chart <- ggplot(shares, aes(x = group, y = percent)) +
  geom_col() +
  labs(x = NULL, y = "Percent")
pie_chart + bar_chart
```

In the pie the slices look equal. In the bars the decline from A to E is obvious.

---

## Two more rules of thumb

- put values to be compared **next to each other on the same scale**, within one panel;
- use colour to **separate a few categories**, not to carry the main message.

Prefer bar charts and dot plots to pie charts, and scatter plots to bubble charts.

---

## Choosing a graph

| Question | Variables | Graph |
|---|---|---|
| how are values distributed? | one numeric | histogram, density |
| how many in each category? | one categorical | bar chart |
| do groups differ? | numeric by categorical | box plot, points |
| how are two variables related? | two numeric | scatter plot |
| how does it change over time? | numeric over time | line chart |
| how do many variables relate? | several numeric | heat map |

---

## The grammar of graphics

```{mermaid}
flowchart LR
    D["Data<br>a data frame"] --> M["Mappings<br>aes(): variable → property"] --> G["Geoms<br>points, bars, lines"]
```

Every ggplot2 graph combines these three parts, joined with `+`.

---

## Data and mappings give an empty canvas

:::: {.columns}
::: {.column}
```r
pilot <- tibble(
  sleep  = c(6.5, 7.5, 5.5, 8, 6),
  stress = c(3.2, 2.1, 4.5, 1.8, 3.9)
)

ggplot(pilot, aes(x = sleep, y = stress))
```

The axes exist, but nothing is drawn yet.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
pilot <- tibble(sleep = c(6.5, 7.5, 5.5, 8, 6), stress = c(3.2, 2.1, 4.5, 1.8, 3.9))
ggplot(pilot, aes(x = sleep, y = stress))
```
:::
::::

---

## A geom draws the data

:::: {.columns}
::: {.column}
```r
ggplot(pilot, aes(x = sleep, y = stress)) +
  geom_point(size = 3)
```

One point per row. Every further detail is another `+`.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(pilot, aes(x = sleep, y = stress)) + geom_point(size = 3)
```
:::
::::

---

## One row per student

```r
first_sem <- semesters |>
  filter(semester == 1) |>
  left_join(students, join_by(student_id))
```

Each student's first-semester record, with their background information.

Most graphs are simplest with one row per case.

---

## Histograms show a distribution

:::: {.columns}
::: {.column}
```r
ggplot(first_sem, aes(x = sleep_hours)) +
  geom_histogram(binwidth = 0.5)
```

Bell-shaped, centred a little above 6 hours.

`r round(100 * mean(first_sem$sleep_hours < 7, na.rm = TRUE))`% of students sleep less than the recommended 7.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(first_sem, aes(x = sleep_hours)) + geom_histogram(binwidth = 0.5, colour = "white")
```
:::
::::

---

## The binwidth changes the picture

```{r}
#| fig-width: 9
#| fig-height: 3.4
bw <- function(b) ggplot(first_sem, aes(x = sleep_hours)) +
  geom_histogram(binwidth = b, colour = "white") +
  labs(title = paste("binwidth =", b), x = NULL, y = NULL)
bw(0.1) + bw(0.5) + bw(2)
```

Too narrow is noisy, too wide hides the shape. Try a few widths.

---

## Bar charts count categories

:::: {.columns}
::: {.column}
```r
ggplot(students, aes(x = faculty)) +
  geom_bar()
```

`geom_bar()` does the counting.

For heights already calculated, such as averages, use `geom_col()` with `y`.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(students, aes(y = faculty)) + geom_bar() + labs(y = NULL)
```
:::
::::

---

## Box plots compare groups

:::: {.columns}
::: {.column}
```r
ggplot(first_sem,
       aes(x = study_mode, y = wellbeing)) +
  geom_boxplot()
```

Part-time students sit a little lower, but the boxes overlap considerably.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(first_sem, aes(x = study_mode, y = wellbeing)) +
  geom_boxplot(fill = "#eaf0f7", colour = navy) + labs(x = NULL)
```
:::
::::

---

## Reading a box plot

| Part | Shows |
|---|---|
| box | the middle half of the values |
| line inside | the median |
| whiskers | the typical range |
| points beyond | unusual values |

Whether the study-mode difference is larger than chance is a question for Chapter 8.

---

## Scatter plots show relationships

:::: {.columns}
::: {.column}
```r
ggplot(first_sem,
       aes(x = sleep_hours, y = gpa)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm")
```

`alpha` makes points transparent, against **overplotting**.

The line rises, but the points scatter widely.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(first_sem, aes(x = sleep_hours, y = gpa)) +
  geom_point(alpha = 0.4) + geom_smooth(method = "lm", colour = copper, fill = "grey80")
```
:::
::::

---

## Line charts show change over time

:::: {.columns}
::: {.column}
```r
wellbeing_by_sem <- semesters |>
  left_join(students, join_by(student_id)) |>
  summarise(mean_wellbeing = mean(wellbeing),
            .by = c(semester, workshop))

ggplot(wellbeing_by_sem,
       aes(x = semester, y = mean_wellbeing,
           colour = workshop)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5)
```
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(wellbeing_by_sem, aes(x = semester, y = mean_wellbeing, colour = workshop)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  theme(legend.position = "bottom")
```
:::
::::

---

## Mapping and setting

```r
aes(fill = workshop)                          # mapped: depends on the data
geom_histogram(fill = "steelblue")            # set: the same everywhere
```

| | Inside `aes()` | Outside `aes()` |
|---|---|---|
| meaning | depends on a variable | fixed for the whole layer |
| legend | added | none |

---

## The most common ggplot2 mistake

:::: {.columns}
::: {.column}
```r
ggplot(first_sem,
       aes(x = sleep_hours,
           fill = "steelblue")) +
  geom_histogram(binwidth = 0.5)
```

A fixed colour inside `aes()` is treated as data: salmon bars and a pointless legend.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(first_sem, aes(x = sleep_hours, fill = "steelblue")) +
  geom_histogram(binwidth = 0.5) + theme(legend.position = "bottom")
```
:::
::::

---

## `fill` and `colour`

| Aesthetic | Controls |
|---|---|
| `fill` | the inside of a shape: a bar, a box |
| `colour` | the outline, or points and lines |

A white `colour` on histogram bars separates them clearly.

---

## From exploratory to explanatory

```{mermaid}
flowchart LR
    A["Labels"] --> B["Scales and<br>colours"] --> C["Theme"] --> D["Annotations"]
```

Each addition is a design decision with a reason behind it.

---

## Labels say what was measured

```r
p <- ggplot(wellbeing_by_sem,
            aes(x = semester, y = mean_wellbeing, colour = workshop)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(title    = "Wellbeing over the two years",
       subtitle = "Average score per semester, by workshop group",
       x = "Semester", y = "Average wellbeing (0 to 100)",
       colour = "Workshop")
```

Units in the axis labels, the message in the title. Storing the graph as `p` lets you add to it.

---

## Scales and colour-blind-friendly palettes

```r
p <- p +
  scale_x_continuous(breaks = 1:4) +
  scale_colour_viridis_d(end = 0.8)
```

Whole semesters only: there is no semester 2.5.

Viridis palettes work for colour vision deficiency, about one man in twelve, and print in greyscale.

---

## Colour should never be the only cue

In a printed figure, groups should also differ by:

- shape or line type;
- labels placed directly on the lines.

Manual colours: `scale_colour_manual(values = c("Invited" = "#1b9e77", "Not invited" = "#7570b3"))`.

---

## Themes control everything that is not data

:::: {.columns}
::: {.column}
```r
p <- p +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

`base_size` sets the font size; `theme()` adjusts details.

The same data, now ready for a thesis.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 4
ggplot(wellbeing_by_sem, aes(x = semester, y = mean_wellbeing, colour = workshop)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  labs(title = "Wellbeing over the two years", x = "Semester",
       y = "Average wellbeing (0 to 100)", colour = "Workshop") +
  scale_x_continuous(breaks = 1:4) + scale_colour_viridis_d(end = 0.8) +
  theme(legend.position = "bottom")
```
:::
::::

---

## Annotations give a reference

:::: {.columns}
::: {.column}
```r
ggplot(first_sem, aes(x = sleep_hours)) +
  geom_histogram(binwidth = 0.5) +
  geom_vline(xintercept = 7,
             linetype = "dashed") +
  annotate("text", x = 7.1, y = Inf,
           label = "Recommended: 7 hours",
           hjust = 0, vjust = 1.5)
```

`y = Inf` puts the text at the top, whatever the bar heights.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(first_sem, aes(x = sleep_hours)) +
  geom_histogram(binwidth = 0.5, colour = "white") +
  geom_vline(xintercept = 7, linetype = "dashed", colour = copper, linewidth = 0.9) +
  annotate("text", x = 7.1, y = Inf, label = "Recommended: 7 hours",
           hjust = 0, vjust = 1.5, colour = copper, size = 4.5) +
  labs(x = "Hours of sleep per night", y = "Students")
```
:::
::::

---

## A graph can be accurate and still mislead

Most misleading graphs are not deliberate. They come from defaults nobody questioned.

| Choice | The risk |
|---|---|
| axis range | lengths no longer mean their values |
| aspect ratio | the same line looks steep or flat |
| clutter | decoration competes with the data |
| colour | bright bands suggest boundaries |

---

## Bar axes start at zero

A bar represents its value by its **length**.

If the axis starts elsewhere, a bar twice as long no longer means twice as much.

Points and lines use **position**, so their axes may be zoomed, as long as the reader is told.

---

## Zooming a line chart

The wellbeing line chart's axis runs only from about `r floor(min(wellbeing_by_sem$mean_wellbeing))` to `r ceiling(max(wellbeing_by_sem$mean_wellbeing))`, which makes a 5-point gap look large.

```r
p + coord_cartesian(ylim = c(0, 100))
```

Either say so in the caption, or show the full scale and let readers judge the size.

---

## Aspect ratio, clutter, and colour

- a graph somewhat wider than tall; the same shape for graphs that will be compared;
- no heavy grid lines, repeated legends, or three-dimensional effects;
- **qualitative** palettes for categories;
- **sequential** palettes for quantities in one direction;
- **diverging** palettes around a meaningful midpoint such as zero.

A rainbow palette is poor for all of these.

---

## A misleading graph

```{r}
#| fig-width: 9
#| fig-height: 3.6
ggplot(faculty_means, aes(x = faculty, y = wellbeing, fill = faculty)) +
  geom_col() +
  coord_cartesian(ylim = c(58, 62.5)) +
  theme_gray(base_size = 13)
```

The Education bar looks several times taller than Social Sciences. The averages differ by about `r round(diff(range(faculty_means$wellbeing)), 1)` points out of 100.

---

## What is wrong with it

- the axis starts at 58, so the bar lengths mean nothing;
- the colours repeat the axis labels, and the legend repeats them again;
- the faculties are alphabetical, not in a meaningful order;
- the axis titles are variable names;
- only averages are shown, hiding how much students differ.

---

## The same data, redesigned

```{r}
#| fig-width: 9
#| fig-height: 3.4
ggplot(first_sem, aes(x = wellbeing, y = reorder(faculty, wellbeing))) +
  geom_jitter(height = 0.15, alpha = 0.25, colour = "grey45") +
  stat_summary(fun = mean, geom = "point", size = 4, colour = copper) +
  labs(x = "Wellbeing in semester 1 (0 to 100)", y = NULL)
```

Differences between faculties are small compared with the differences within them.

---

## How the redesign was built

```r
ggplot(first_sem, aes(x = wellbeing, y = reorder(faculty, wellbeing))) +
  geom_jitter(height = 0.15, alpha = 0.25, colour = "grey45") +
  stat_summary(fun = mean, geom = "point", size = 3.5, colour = "#b2182b") +
  labs(x = "Wellbeing in semester 1 (0 to 100)", y = NULL)
```

| Function | Job |
|---|---|
| `reorder()` | order faculties by their average |
| `geom_jitter()` | spread points so they do not hide each other |
| `stat_summary()` | calculate and draw each mean |

---

## Facets: small multiples

:::: {.columns}
::: {.column}
```r
ggplot(first_sem,
       aes(x = sleep_hours, y = gpa)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  facet_wrap(~ programme)
```

One panel per group, with shared axes, so comparison is by position.

`facet_grid(rows ~ columns)` for two grouping variables.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(first_sem, aes(x = sleep_hours, y = gpa)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "lm", colour = copper, fill = "grey80") +
  facet_wrap(~ programme) + labs(x = "Sleep (hours)", y = "GPA")
```
:::
::::

---

## Density plots compare distributions

:::: {.columns}
::: {.column}
```r
semesters |>
  filter(semester == 2) |>
  left_join(students, join_by(student_id)) |>
  ggplot(aes(x = wellbeing, fill = workshop)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(end = 0.8)
```

The whole distribution moved, not just a few students.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
semesters |> filter(semester == 2) |> left_join(students, join_by(student_id)) |>
  ggplot(aes(x = wellbeing, fill = workshop)) +
  geom_density(alpha = 0.5, colour = NA) + scale_fill_viridis_d(end = 0.8) +
  labs(x = "Wellbeing (0 to 100)", y = "Density", fill = NULL) +
  theme(legend.position = "bottom")
```
:::
::::

---

## Heat maps show a table of numbers

```r
correlations |>
  as.data.frame() |>
  mutate(var1 = rownames(correlations)) |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(x = var1, y = var2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = round(r, 2))) +
  scale_fill_gradient2(limits = c(-1, 1))
```

The matrix becomes a long table with one row per pair (Chapter 3).

---

## Reading the heat map

```{r}
#| fig-width: 6.5
#| fig-height: 4.2
correlations <- first_sem |> select(gpa, sleep_hours, study_hours, wellbeing) |> cor(use = "complete.obs")
correlations |> as.data.frame() |> mutate(var1 = rownames(correlations)) |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(x = var1, y = var2, fill = r)) +
  geom_tile(colour = "white") + geom_text(aes(label = round(r, 2)), size = 5) +
  scale_fill_gradient2(limits = c(-1, 1)) + labs(x = NULL, y = NULL, fill = "r")
```

A diverging palette, white at zero. The numbers matter: colour alone is read imprecisely.

---

## Exporting figures

```r
ggsave("figures/wellbeing-lines.png", plot = p,
       width = 16, height = 10, units = "cm", dpi = 300)
ggsave("figures/wellbeing-lines.pdf", plot = p,
       width = 16, height = 10, units = "cm")
```

The file type follows from the file name.

---

## Save at the printed size

| Decision | Rule |
|---|---|
| width | about 8 cm for one journal column, 16 cm for a full page |
| resolution | at least 300 dpi for PNG |
| format | PDF or SVG when accepted: sharp at any size |
| text | check it at the final size; adjust `base_size` |

The figure code lives in the script: a supervisor's request is one changed line.

---

## In your field: economics and business

:::: {.columns}
::: {.column}
```r
ggplot(economics,
       aes(x = date, y = unemploy)) +
  geom_line()
```

US unemployment, 1967 to 2015, with each recession a peak.

It shows the **number** unemployed, not the rate, while the population grew greatly.
:::
::: {.column}
```{r}
#| fig-width: 4.3
#| fig-height: 3.4
ggplot(economics, aes(x = date, y = unemploy)) + geom_line(colour = navy) +
  labs(x = NULL, y = "Unemployed (thousands)")
```
:::
::::

---

## Practical lab: the Chapter 4 playground

Work through the playground exercises in your browser, with hints and solutions.

Every exercise also runs in RStudio, from the downloadable chapter project.

---

## Practical exercises 1–3: the basic graphs

1. A histogram of study hours with binwidths 1, 5, and 10: which shows the shape best?
2. A bar chart of employment in one colour, ordered from no job to full-time.
3. Box plots of GPA by faculty, with clear labels.

---

## Practical exercises 4–6: relationships and honesty

4. Caffeine against sleep, with a trend line: what does it suggest?
5. Save a graph as a PNG, 16 by 10 cm, at 300 dpi.
6. Redraw the misleading graph from zero, and choose one of the three for a thesis.

---

## Try this yourself

Pick one variable from your own research.

- make three exploratory graphs of it in five minutes;
- choose the one that answers your question;
- turn it into an explanatory graph: labels, scale, theme, one annotation;
- save it at 16 cm wide.

Then show it to someone for five seconds and ask what they saw.

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| salmon colour and a strange legend | a fixed colour inside `aes()` |
| points hide each other | overplotting; use `alpha` or `geom_jitter()` |
| categories in alphabetical order | factor levels not set, or no `reorder()` |
| "semester 2.5" on the axis | a numeric axis without `breaks` |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| small differences look dramatic | a zoomed axis on bars |
| text unreadable in the thesis | saved too large, then shrunk |
| groups vanish in print | colour was the only cue |
| a blurry figure | a PNG below 300 dpi |

---

## Completion checklist

- [ ] every variable was graphed before it was analysed;
- [ ] the graph matches the question and the variable types;
- [ ] the key comparison uses position or length in one panel;
- [ ] axes are labelled with units;
- [ ] bar axes start at zero, and zoomed axes are stated;
- [ ] colours are colour-blind-friendly and not the only cue;
- [ ] the figure is saved at its printed size.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| summary statistics tell the whole story | Anscombe's sets share every summary |
| more elements make a graph more informative | decoration competes with the data |
| pie charts are good for proportions | bars or dots are almost always clearer |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| zooming an axis is always wrong | wrong for bars, acceptable for lines if stated |
| colour makes a graph clearer | colour separates a few categories only |
| a graph is decoration at the end | graphs are one of the main instruments |

---

## The chapter in one sentence

> **Look at the data first, then choose a graph that puts the key comparison in position or length and tells the truth about its size.**

---

## Next: Chapter 5

The next chapter moves from a topic to a plan:

- what data analysis is for;
- from a topic to a research question and a hypothesis;
- variables, measurement, reliability, and validity;
- study designs and randomisation;
- an analysis plan written before the data is seen.

---

## Questions

Think of a graph that convinced you of something.

Would it still convince you with the axis starting at zero?
