---
title: "ANOVA and Regression"
description: "Chapter 8: partitioning variation, one-way and two-way ANOVA, simple and multiple regression, confounders, curves, interactions, diagnostics, and logistic regression."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 14)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x)

scores <- questionnaire |>
  mutate(stress_4 = 6 - stress_4,
         stress  = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
         support = rowMeans(pick(support_1:support_6), na.rm = TRUE)) |>
  select(student_id, stress, support)
study <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1), join_by(student_id))

group_scores <- tibble(
  group  = rep(c("A", "B", "C"), each = 4),
  stress = c(2.5, 3.0, 2.8, 3.1, 3.4, 3.8, 3.5, 3.9, 2.9, 3.3, 3.0, 3.4))
group_means <- group_scores |> summarise(mean = mean(stress), .by = group)
offsets   <- rep(c(-0.3, -0.1, 0.1, 0.3), 3)
group_avg <- rep(group_means$mean, each = 4)
Fval <- function(d) summary(aov(stress ~ group, data = d))[[1]][["F value"]][1]
F_tight <- Fval(group_scores |> mutate(stress = group_avg + 0.5 * offsets))
F_wide  <- Fval(group_scores |> mutate(stress = group_avg + 4 * offsets))

stress_anova <- aov(stress ~ faculty, data = study)
anova_table  <- summary(stress_anova)[[1]]
eta_sq <- anova_table[["Sum Sq"]][1] / sum(anova_table[["Sum Sq"]])
tukey  <- TukeyHSD(stress_anova)$faculty
hs_hum <- tukey["Humanities-Health Sciences", ]

five <- tibble(sleep = c(5, 6, 6.5, 7, 8), gpa = c(2.8, 3.0, 3.2, 3.1, 3.4))
gpa_simple <- lm(gpa ~ sleep_hours, data = study)
gpa_model  <- lm(gpa ~ sleep_hours + study_hours + stress + support, data = study)
caf1 <- coef(summary(lm(gpa ~ caffeine_mg, data = study)))
caf2 <- coef(summary(lm(gpa ~ caffeine_mg + sleep_hours, data = study)))
gpa_curve  <- lm(gpa ~ study_hours + I(study_hours^2), data = study)
peak_hours <- -coef(gpa_curve)[2] / (2 * coef(gpa_curve)[3])
study <- study |> mutate(support_c = support - mean(support),
                         dropout = as.integer(considering_dropout == "Yes"))
b <- coef(lm(gpa ~ support_c * programme + sleep_hours + study_hours, data = study))
slope_masters <- b["support_c"]; slope_phd <- b["support_c"] + b["support_c:programmePhD"]
g  <- broom::glance(gpa_model)
tm <- broom::tidy(gpa_model, conf.int = TRUE); sl <- tm[tm$term == "sleep_hours", ]

jobs <- table(students$employment, students$considering_dropout)
odds_full <- jobs["Full-time job", "Yes"] / jobs["Full-time job", "No"]
odds_none <- jobs["None", "Yes"] / jobs["None", "No"]
dropout_model <- glm(dropout ~ stress + support + financial_worry + employment + study_mode,
                     data = study, family = binomial)
or <- exp(coef(dropout_model)); or_ci <- exp(confint.default(dropout_model))
examples <- tibble(stress = c(2.5, 4.0), support = c(4.0, 2.0), financial_worry = 3,
                   employment = "None", study_mode = "Full-time")
p_ex <- predict(dropout_model, newdata = examples, type = "response")
tree_model <- lm(Volume ~ Girth + Height, data = datasets::trees)
```

# ANOVA and Regression

How much of the variation can other variables explain?

**Chapter 8**

Polla Fattah

---

## By the end of today you can

- explain how ANOVA splits variation, and what F compares;
- compare three or more groups, and find which differ;
- analyse two factors and their interaction;
- fit and interpret simple and multiple linear regression;
- separate a real effect from a confounder;
- model curves and interactions;
- read diagnostic plots;
- interpret logistic regression as odds ratios and probabilities;
- report ANOVA and regression in a thesis.

---

## One idea behind both methods

Students differ in stress, grades, and wellbeing.

Which part of those differences is connected with faculty, sleep, or support, and which part remains unexplained?

| Method | Models |
|---|---|
| ANOVA | the averages of three or more groups |
| linear regression | a numeric outcome from predictors |
| logistic regression | a yes-or-no outcome from predictors |

---

## The data for this chapter

```r
scores <- questionnaire |>
  mutate(stress_4 = 6 - stress_4,
         stress   = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
         support  = rowMeans(pick(support_1:support_6), na.rm = TRUE)) |>
  select(student_id, stress, support)

study <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1), join_by(student_id))
```

One row per student: background, first semester, and questionnaire scores.

---

## Why not ten t-tests?

Five faculties give 10 pairs.

Each test has a 5% false alarm risk, and across 10 tests the chance of at least one is about **40%**.

ANOVA asks one question first: do the faculty averages differ **at all**?

---

## Variation between and within groups

```{r}
#| fig-width: 7
#| fig-height: 3.6
ggplot(group_scores, aes(x = group, y = stress)) +
  geom_hline(yintercept = mean(group_scores$stress), linetype = "dashed", colour = soft) +
  geom_point(size = 3, colour = navy,
             position = position_nudge(x = rep(c(-0.09, -0.03, 0.03, 0.09), 3))) +
  geom_errorbar(data = group_means, aes(y = mean, ymin = mean, ymax = mean),
                width = 0.4, linewidth = 1.3, colour = copper) +
  labs(x = "Group", y = "Stress score")
```

Group averages (copper) vs the overall average: **between**. Points vs their own average: **within**.

---

## The F statistic

$$
F = \frac{\text{variation between groups}}{\text{variation within groups}}
$$

```r
summary(aov(stress ~ group, data = group_scores))
```

F near 1: averages differ no more than chance produces. Large F: they differ more.

---

## The ratio matters, not only the gaps

```r
tight <- group_scores |> mutate(stress = group_avg + 0.5 * offsets)
wide  <- group_scores |> mutate(stress = group_avg + 4 * offsets)
```

| Version | Group averages | F |
|---|---|---|
| tight | identical | `r round(F_tight, 1)` |
| wide | identical | `r round(F_wide, 1)` |

Half a point is convincing when students within a group hardly differ, and unremarkable when they differ by two.

---

## Stress by faculty

```{r}
#| fig-width: 9
#| fig-height: 3.6
ggplot(study, aes(x = faculty, y = stress)) +
  geom_boxplot(fill = "#eaf0f7", colour = navy) +
  labs(x = NULL, y = "Stress (1 to 5)")
```

Much overlap. Health Sciences sits a little higher, Humanities a little lower.

---

## One-way ANOVA

```r
stress_anova <- aov(stress ~ faculty, data = study)
summary(stress_anova)
```

*p* = `r round(anova_table[["Pr(>F)"]][1], 3)`: the faculties differ in stress.

The next question is how much.

---

## Eta squared: the size of the effect

```r
ss <- summary(stress_anova)[[1]][["Sum Sq"]]
ss[1] / sum(ss)
```

$\eta^2$ ≈ `r f2(eta_sq)`: only about `r round(100 * eta_sq)`% of the differences in stress lie between faculties.

| $\eta^2$ | 0.01 | 0.06 | 0.14 |
|---|---|---|---|
| label | small | medium | large |

Faculty matters, but says very little about one student's stress.

---

## Post-hoc tests: which groups differ?

```r
TukeyHSD(stress_anova)
```

Tukey's test compares every pair while keeping the overall false alarm rate at 5%.

Only one pair differs: Health Sciences is `r f2(abs(hs_hum[["diff"]]))` points more stressed than Humanities.

---

## Assumptions of ANOVA

| Assumption | Check |
|---|---|
| independent observations | the design |
| roughly normal in each group | graphs, or large groups |
| similar spread | Levene's test: `car::leveneTest()` |

```r
kruskal.test(stress ~ faculty, data = study)
```

Kruskal-Wallis compares ranks when assumptions fail. Here it agrees.

---

## Two-way ANOVA

| Term | Asks |
|---|---|
| main effect of programme | Master's vs PhD, on average |
| main effect of study mode | full-time vs part-time, on average |
| interaction | does the study-mode effect **depend on** programme? |

---

## An interaction plot

```{r}
#| fig-width: 7
#| fig-height: 3.6
study |>
  summarise(wellbeing = mean(wellbeing), .by = c(programme, study_mode)) |>
  ggplot(aes(x = programme, y = wellbeing, colour = study_mode, group = study_mode)) +
  geom_line(linewidth = 1.2) + geom_point(size = 3.5) +
  scale_colour_manual(values = c(navy, copper)) +
  labs(x = NULL, y = "Average wellbeing", colour = NULL)
```

Parallel lines: no interaction. Part-time is lower in both programmes.

---

## Testing main effects and interaction

```r
summary(aov(wellbeing ~ programme * study_mode, data = study))
```

`programme * study_mode` asks for both main effects and their interaction.

Study mode matters; programme and the interaction do not.

A significant interaction is interpreted **before** the main effects.

---

## Regression models the average outcome

ANOVA compares groups. Regression models how an outcome changes with predictors, numeric or categorical.

ANOVA is a special case of regression.

---

## Average GPA at each level of sleep

```r
sleep_bands <- study |>
  filter(!is.na(sleep_hours)) |>
  mutate(sleep_band = round(sleep_hours * 2) / 2) |>
  summarise(mean_gpa = mean(gpa), students = n(), .by = sleep_band)
```

The average GPA of students who sleep 5 hours, 5.5 hours, 6 hours, and so on.

---

## The line summarises conditional averages

```{r}
#| fig-width: 9
#| fig-height: 3.8
sleep_bands <- study |> filter(!is.na(sleep_hours)) |>
  mutate(sleep_band = round(sleep_hours * 2) / 2) |>
  summarise(mean_gpa = mean(gpa), students = n(), .by = sleep_band)
ggplot(study, aes(x = sleep_hours, y = gpa)) +
  geom_point(alpha = 0.2, colour = "grey45") +
  geom_smooth(method = "lm", se = FALSE, colour = navy, linewidth = 1.2) +
  geom_point(data = sleep_bands, aes(x = sleep_band, y = mean_gpa, size = students), colour = copper) +
  labs(x = "Hours of sleep per night", y = "GPA", size = "Students")
```

The spread of grey points around the line is what the model does not explain.

---

## Simple linear regression

$$
\text{GPA} = b_0 + b_1 \times \text{sleep} + \text{error}
$$

| Term | Meaning |
|---|---|
| $b_0$, intercept | predicted GPA at zero sleep; places the line |
| $b_1$, slope | change in predicted GPA per extra hour |
| residual | vertical distance from point to line |

**Least squares** makes the squared residuals as small as possible.

---

## Fitting a line to five students

```r
five <- tibble(sleep = c(5, 6, 6.5, 7, 8),
               gpa   = c(2.8, 3.0, 3.2, 3.1, 3.4))
lm(gpa ~ sleep, data = five)
```

Each extra hour goes with a GPA about `r f2(coef(lm(gpa ~ sleep, data = five))[2])` higher.

Nobody sleeps zero hours, so the intercept rarely means anything on its own.

---

## Three results in `summary()`

```r
gpa_simple <- lm(gpa ~ sleep_hours, data = study)
summary(gpa_simple)
```

| Result | Value | Meaning |
|---|---|---|
| slope | `r f2(coef(gpa_simple)[2])` | GPA per extra hour of sleep |
| R-squared | `r f2(summary(gpa_simple)$r.squared)` | share of GPA variation explained |
| residual SE | `r f2(sigma(gpa_simple))` | typical distance from the line |

Sleep matters, but it is far from the whole story.

---

## Multiple regression

```r
gpa_model <- lm(gpa ~ sleep_hours + study_hours + stress + support,
                data = study)
```

| Predictor | Change in GPA per unit |
|---|---|
| sleep (hour) | `r f2(coef(gpa_model)["sleep_hours"])` |
| study (hour a week) | `r sprintf("%.3f", coef(gpa_model)["study_hours"])` |
| stress (point) | `r f2(coef(gpa_model)["stress"])` |
| support (point) | `r f2(coef(gpa_model)["support"])` |

Each coefficient **holds the other predictors constant**.

---

## How much the model explains

R-squared ≈ `r round(100 * summary(gpa_model)$r.squared)`%: modest, and normal for grades, which depend on much no survey measures.

- **adjusted R-squared** corrects for the number of predictors;
- `lm()` drops students with any missing value: `r length(gpa_model$na.action)` here.

Always report the number of cases each model is based on.

---

## Categorical predictors

```r
lm(gpa ~ sleep_hours + study_hours + stress + support + gender,
   data = study)
```

R compares each category with a **reference category**, alphabetically first: `Female`.

`genderMale` is tiny and far from significant: no evidence that gender affects GPA. That is a finding too.

`relevel(factor(faculty), ref = "Humanities")` chooses another reference.

---

## Two explanations for caffeine and GPA

```{mermaid}
flowchart LR
  subgraph A["Caffeine as a cause"]
    C1[Caffeine] --> G1[GPA]
  end
  subgraph B["Sleep as a confounder"]
    S2[Sleep] --> C2[Caffeine]
    S2 --> G2[GPA]
    C2 -. "apparent link" .- G2
  end
```

Regression can tell them apart by comparing students who sleep the same amount.

---

## Adjusting for the confounder

```r
lm(gpa ~ caffeine_mg, data = study)
lm(gpa ~ caffeine_mg + sleep_hours, data = study)
```

| Model | Caffeine coefficient | p |
|---|---|---|
| caffeine alone | `r sprintf("%.5f", caf1["caffeine_mg", 1])` | `r format.pval(caf1["caffeine_mg", 4], eps = 0.001, digits = 2)` |
| with sleep | `r sprintf("%.5f", caf2["caffeine_mg", 1])` | `r sprintf("%.2f", caf2["caffeine_mg", 4])` |

Among students who sleep the same, caffeine makes no difference to grades.

---

## Regression adjusts only for what you include

A confounder is removed only if it was **measured** and **put in the model**.

Outside a randomised experiment, regression shows associations adjusted for the variables in the model, not proof of cause.

---

## Curved relationships

```r
gpa_curve <- lm(gpa ~ study_hours + I(study_hours^2), data = study)
```

`I()` calculates `study_hours^2` before fitting.

The squared term is negative and significant: the line bends downwards.

---

## Diminishing returns

```{r}
#| fig-width: 9
#| fig-height: 3.6
ggplot(study, aes(x = study_hours, y = gpa)) +
  geom_point(alpha = 0.3, colour = navy) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2), colour = copper, fill = "grey80") +
  labs(x = "Study hours per week", y = "GPA")
```

The first hours make a real difference. The curve peaks at about `r round(peak_hours)` hours a week.

---

## Interactions in regression

```r
study <- study |> mutate(support_c = support - mean(support))
lm(gpa ~ support_c * programme + sleep_hours + study_hours, data = study)
```

**Centring**: zero now means average support.

| Programme | GPA per point of support |
|---|---|
| Master's | `r f2(slope_masters)` |
| PhD | `r f2(slope_phd)` |

Support matters about `r round(slope_phd / slope_masters, 1)` times as much for PhD students.

---

## What violated assumptions look like

```{r}
#| fig-width: 9
#| fig-height: 3.3
set.seed(10)
x <- runif(200, 0, 10)
simulated <- list("Assumptions met" = 2 + 0.5 * x + rnorm(200),
                  "Curved" = 2 + 0.25 * x^2 + rnorm(200),
                  "Spread increases" = 2 + 0.5 * x + rnorm(200, sd = 0.2 + 0.3 * x))
rd <- bind_rows(lapply(names(simulated), function(nm) {
  fit <- lm(simulated[[nm]] ~ x)
  tibble(case = nm, fitted = fitted(fit), residual = resid(fit))}))
rd$case <- factor(rd$case, levels = names(simulated))
ggplot(rd, aes(fitted, residual)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = soft) +
  geom_point(alpha = 0.5, size = 1.2, colour = navy) +
  facet_wrap(~ case, scales = "free") + labs(x = "Fitted values", y = "Residuals")
```

A shapeless band is good. A U shape means a missed curve. A funnel makes p-values unreliable.

---

## Checking the GPA model

```r
par(mfrow = c(1, 2))
plot(gpa_model, which = 1:2)
```

Residuals against fitted: a shapeless band, no curve, no funnel.

Q-Q plot of residuals: close to the line, so roughly normal. The model looks good.

---

## Tidy results with broom

```r
library(broom)
tidy(gpa_model, conf.int = TRUE)   # coefficients with intervals
glance(gpa_model)                  # overall fit
```

Model results as data frames, ready for a thesis table.

---

## Writing it up

> A multiple regression of first-semester GPA on sleep, study hours, stress, and supervisor support explained `r round(100 * g$r.squared)`% of the variance, *F*(`r g$df`, `r g$df.residual`) = `r f2(g$statistic)`, *p* < .001, *N* = `r g$nobs`. Holding the other predictors constant, each additional hour of sleep was associated with a GPA `r f2(sl$estimate)` points higher (95% CI [`r f2(sl$conf.low)`, `r f2(sl$conf.high)`]).

The null hypothesis of a zero sleep coefficient was rejected.

---

## Yes-or-no outcomes need logistic regression

Linear regression could predict probabilities below 0 or above 1.

**Logistic regression** models the probability of "yes" in a way that stays between 0 and 1.

---

## Odds and odds ratios

```r
jobs <- table(students$employment, students$considering_dropout)
odds_full <- jobs["Full-time job", "Yes"] / jobs["Full-time job", "No"]
odds_none <- jobs["None", "Yes"] / jobs["None", "No"]
```

| Group | Odds of considering dropout |
|---|---|
| full-time job | `r f2(odds_full)` |
| no job | `r f2(odds_none)` |
| odds ratio | `r round(odds_full / odds_none, 1)` |

1 means no difference; above 1 higher odds; below 1 lower.

---

## A model for dropout

```r
study <- study |> mutate(dropout = as.integer(considering_dropout == "Yes"))

dropout_model <- glm(
  dropout ~ stress + support + financial_worry + employment + study_mode,
  data = study, family = binomial)

exp(cbind(odds_ratio = coef(dropout_model),
          confint.default(dropout_model)))
```

The outcome is 0 or 1. `exp()` turns log-odds into odds ratios.

---

## Reading the odds ratios

| Predictor (per point) | Odds ratio | 95% CI |
|---|---|---|
| stress | `r f2(or["stress"])` | `r f2(or_ci["stress", 1])` to `r f2(or_ci["stress", 2])` |
| financial worry | `r f2(or["financial_worry"])` | `r f2(or_ci["financial_worry", 1])` to `r f2(or_ci["financial_worry", 2])` |
| support | `r f2(or["support"])` | `r f2(or_ci["support", 1])` to `r f2(or_ci["support", 2])` |

Stress raises the odds; support cuts them by about `r round(100 * (1 - or["support"]))`%. The stress hypothesis is supported.

---

## What happened to employment?

Alone, a full-time job about doubled the odds of considering dropout.

In the model, both employment intervals include 1.

A job likely matters **because** of the stress and money pressure it brings. Multiple regression makes that visible.

---

## Predicted probabilities are easier to read

```r
examples <- tibble(stress = c(2.5, 4.0), support = c(4.0, 2.0),
                   financial_worry = 3, employment = "None",
                   study_mode = "Full-time")
predict(dropout_model, newdata = examples, type = "response")
```

| Student | Chance of considering dropout |
|---|---|
| low stress, good support | `r round(100 * p_ex[1])`% |
| high stress, little support | `r round(100 * p_ex[2])`% |

---

## Explaining is not predicting

This model explains who considers dropping out.

Whether it can **predict** that for new students, and how to test that fairly, is a different question.

It is the subject of Chapters 11 and 12.

---

## In your field: forestry and agriculture

```r
tree_model <- lm(Volume ~ Girth + Height, data = trees)
summary(tree_model)$r.squared
```

Timber volume of 31 black cherry trees from measurements taken without felling them.

Girth and height explain `r round(100 * summary(tree_model)$r.squared)`% of volume, against `r round(100 * summary(gpa_model)$r.squared)`% for GPA: physics is more predictable than people.

---

## Practical lab: the Chapter 8 playground

Work through the playground exercises in your browser, with hints and solutions.

Every exercise also runs in RStudio, from the downloadable chapter project.

---

## Practical exercises 1–3: ANOVA and regression

1. Wellbeing by faculty: ANOVA, $\eta^2$, and Tukey if significant.
2. Wellbeing on sleep: interpret the slope and R-squared.
3. Add stress and support: what happens to the sleep slope and R-squared?

---

## Practical exercises 4–6: categories, odds, confounders

4. Add study mode: name the reference and interpret its coefficient.
5. Dropout on burnout alone: report and explain the odds ratio.
6. Draw a confounder from your own field, and say how regression would separate it.

---

## Try this yourself

Choose one outcome in the student wellbeing data.

- plot it against one predictor;
- fit a simple regression, then add two plausible confounders;
- watch how the first coefficient changes;
- check the residual plot;
- write one sentence reporting the adjusted effect with its interval.

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| ten t-tests, one "significant" | multiple testing; use ANOVA and post-hoc tests |
| significant ANOVA, unclear which group | no post-hoc test |
| a coefficient changes sign when adding a predictor | a confounder, or correlated predictors |
| fewer cases than expected | `lm()` dropped rows with missing values |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| a U shape in the residuals | a missed curve |
| a funnel in the residuals | unequal spread; p-values unreliable |
| predicted probabilities above 1 | linear regression on a yes-or-no outcome |
| coefficients hard to explain to readers | log-odds not turned into odds ratios or probabilities |

---

## Completion checklist

- [ ] groups are compared with one ANOVA, then post-hoc tests;
- [ ] every ANOVA reports $\eta^2$;
- [ ] interactions are checked and interpreted first;
- [ ] each coefficient is read as "holding the others constant";
- [ ] plausible confounders are in the model;
- [ ] residual plots are checked;
- [ ] logistic results are reported as odds ratios or probabilities;
- [ ] the number of cases is reported for each model.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| a significant coefficient shows a cause | an association adjusted for the model's variables |
| a low R-squared makes a model useless | a real effect can explain little variation |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| each coefficient describes its predictor alone | each holds the others constant |
| a significant ANOVA shows all groups differ | at least one differs; post-hoc tests say which |

---

## The chapter in one sentence

> **ANOVA and regression split variation into what the predictors explain and what remains, and each coefficient holds the others constant, but only the ones you measured.**

---

## Next: Chapter 9

The next chapter looks at many variables at once:

- constructs and latent variables;
- principal component analysis;
- factor analysis of the questionnaire;
- reporting a questionnaire's structure;
- cluster analysis for student profiles.

---

## Questions

Which confounder could explain the main association in your own research?

Did you measure it?
