---
title: "Multivariate Statistical Methods"
description: "Chapter 9: latent variables, correlation structure, principal component analysis, exploratory factor analysis, Cronbach's alpha, and k-means and hierarchical clustering."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(psych)
library(factoextra)

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x)

set.seed(42)
true_stress <- rnorm(600)
simulated_items <- sapply(1:6, function(i) true_stress + rnorm(600, sd = 1))
item_r <- sapply(1:6, function(k) cor(rowMeans(simulated_items[, 1:k, drop = FALSE]), true_stress))

items <- questionnaire |> select(-student_id)
items_complete <- na.omit(items)
pca <- prcomp(items_complete, scale. = TRUE)
imp <- summary(pca)$importance
eigen <- pca$sdev^2
set.seed(1)
parallel <- fa.parallel(items, fa = "fa", plot = FALSE)
efa <- fa(items, nfactors = 4, rotate = "oblimin", fm = "ml")
main_loadings <- apply(abs(unclass(efa$loadings)), 1, max)
phi <- efa$Phi
max_phi <- max(abs(phi[upper.tri(phi)]))

q <- questionnaire; q$stress_4 <- 6 - q$stress_4
alphas <- sapply(list(stress = paste0("stress_", 1:6), burnout = paste0("burnout_", 1:6),
                      support = paste0("support_", 1:6), satisfaction = paste0("satisfaction_", 1:4)),
                 function(v) psych::alpha(q[, v])$total$raw_alpha)

scores <- q |>
  mutate(stress = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
         support = rowMeans(pick(support_1:support_6), na.rm = TRUE),
         satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)) |>
  select(student_id, stress, support, satisfaction)
profiles <- semesters |>
  summarise(across(c(sleep_hours, study_hours, caffeine_mg, exercise_days), ~ mean(.x, na.rm = TRUE)),
            .by = student_id) |>
  left_join(scores, join_by(student_id)) |>
  na.omit()
profile_data <- scale(profiles |> select(-student_id))
set.seed(123)
clusters <- kmeans(profile_data, centers = 4, nstart = 25)
tree <- hclust(dist(profile_data), method = "ward.D2")
arrests_pca <- prcomp(USArrests, scale. = TRUE)
```

# Multivariate Statistical Methods

Many variables at once: items, components, factors, and profiles

**Chapter 9**

Polla Fattah

---

## By the end of today you can

- explain why a construct is measured with several items;
- explore the correlations among many variables at once;
- run a PCA and decide how many components to keep;
- tell PCA from factor analysis;
- run an exploratory factor analysis and read its loadings;
- check a scale's reliability with Cronbach's alpha;
- group cases with k-means and hierarchical clustering;
- describe clusters as summaries, not natural groups.

---

## Three multivariate methods

| Method | Groups | Answers |
|---|---|---|
| principal component analysis | variables into a few summaries | how much can be kept in fewer numbers? |
| factor analysis and alpha | items into latent constructs | does the questionnaire measure what it should? (RQ6) |
| cluster analysis | people into profiles | are there distinct kinds of student? (RQ7) |

---

## One item is a poor measure

The answer to "I feel unable to control important things in my studies" depends on:

- the student's stress: the **signal**;
- how they read the question that day, their mood, how they use the scale: **noise**.

A **latent variable** is the construct thought to cause part of every answer.

---

## Averaging keeps the signal

```r
set.seed(42)
true_stress <- rnorm(600)
simulated_items <- sapply(1:6, function(i) true_stress + rnorm(600, sd = 1))
```

| Items averaged | 1 | 2 | 3 | 4 | 5 | 6 |
|---|---|---|---|---|---|---|
| correlation with true stress | `r f2(item_r[1])` | `r f2(item_r[2])` | `r f2(item_r[3])` | `r f2(item_r[4])` | `r f2(item_r[5])` | `r f2(item_r[6])` |

The shared signal adds up; the separate noise partly cancels.

---

## The assumption to check

Averaging works only if the items reflect **one** construct.

If some stress items measured tiredness, the score would mix two things.

| Evidence | Method |
|---|---|
| construct validity: items group as intended | factor analysis |
| internal consistency: items agree | Cronbach's alpha |

---

## 22 items, 231 correlations

```r
items <- questionnaire |> select(-student_id)
corrplot(cor(items, use = "pairwise.complete.obs"),
         method = "color", order = "hclust")
```

Far too many to read one by one.

`order = "hclust"` sorts the items so that those correlating strongly sit together.

---

## Blocks along the diagonal

```{r}
#| fig-width: 6
#| fig-height: 6
library(corrplot)
par(bg = NA)
corrplot(cor(items, use = "pairwise.complete.obs"), method = "color", order = "hclust",
         tl.col = "#172a42", tl.cex = 0.6, cl.cex = 0.6, mar = c(0, 0, 0, 0))
```

Four blocks, one per scale. `stress_4` is red against the other stress items: it is worded the other way.

---

## Principal component analysis

PCA replaces many correlated variables with a few **principal components**.

- the first captures as much of the variation as possible;
- the second as much as possible of what is left;
- and so on.

Two items that almost always agree can be kept as one combined score.

---

## Running PCA

```r
items_complete <- na.omit(items)
pca <- prcomp(items_complete, scale. = TRUE)
summary(pca)$importance[, 1:6]
```

`scale. = TRUE` puts every item on the same scale. PCA needs complete data.

Only `r nrow(items_complete)` of 600 students answered all 22 items: small gaps add up.

The first component captures `r round(100 * imp[2, 1])`%; the first four together `r round(100 * imp[3, 4])`%.

---

## How many components?

| Rule | Keeps | Here |
|---|---|---|
| Kaiser | eigenvalue > 1 | `r sum(eigen > 1)` components |
| scree plot | components before the elbow | 4 |
| parallel analysis | components that beat random data | `r parallel$nfact` |

An **eigenvalue** is the variation a component captures. Parallel analysis is the most reliable.

---

## The scree plot

```{r}
#| fig-width: 8
#| fig-height: 3.6
fviz_eig(pca, addlabels = TRUE, ncp = 10, barfill = navy, barcolor = navy,
         linecolor = copper, ggtheme = theme_minimal(base_size = 14)) +
  labs(title = NULL) + theme(plot.background = element_rect(fill = "transparent", colour = NA))
```

Steep for four components, flat afterwards: four dimensions, matching the four scales.

---

## Parallel analysis

```r
library(psych)
set.seed(1)
parallel <- fa.parallel(items, fa = "fa", plot = FALSE)
parallel$nfact
#> [1] 4
```

Components are kept only if they capture more than the same analysis of random data.

---

## PCA and factor analysis are different

| | PCA | Factor analysis |
|---|---|---|
| assumes | nothing about why items relate | latent factors **cause** the items |
| produces | combinations of items | loadings of items on factors |
| used for | reducing many variables to few | checking a questionnaire |

Many theses use one when they mean the other.

---

## Exploratory factor analysis

```r
efa <- fa(items, nfactors = 4, rotate = "oblimin", fm = "ml")
print(efa$loadings, cutoff = 0.3, sort = TRUE)
```

| Argument | Choice |
|---|---|
| `nfactors = 4` | from parallel analysis |
| `rotate = "oblimin"` | lets factors correlate, as stress and burnout surely do |
| `fm = "ml"` | maximum likelihood estimation |

`fa()` also uses students who skipped an item.

---

## Loadings

A **loading** is how strongly an item relates to a factor.

| Loading (size) | Meaning |
|---|---|
| near 0 | unrelated to the factor |
| above about 0.4 | a clear relationship |

`cutoff = 0.3` hides small loadings; `sort = TRUE` groups items by their main factor.

---

## Four clear factors

- the six support items on one factor;
- the stress items on another;
- the burnout items on a third;
- the four satisfaction items on the fourth.

Main loadings range from `r f2(min(main_loadings))` to `r f2(max(main_loadings))`. The labels ML1 to ML4 are arbitrary; naming factors is the researcher's job.

---

## Reversed and cross-loading items

| Item | Pattern | Meaning |
|---|---|---|
| `stress_4` "I feel confident handling problems" | loads **negatively** on stress | exactly right for a reversed item |
| `burnout_3` "Deadlines make me feel overwhelmed" | loads on burnout **and** stress | a cross-loading, worth discussing |

A cross-loading item may be reworded or removed in future studies.

---

## How the factors correlate

```r
round(efa$Phi, 2)
```

Stress and burnout correlate strongly, about `r f2(max_phi)`, but are not the same thing.

Support and satisfaction correlate positively with each other, negatively with stress and burnout.

Relationships that make sense are themselves evidence of validity.

---

## Cronbach's alpha

Internal consistency: do the items of one scale give consistent results?

- rises when the items correlate strongly;
- rises when there are more items;
- 0.7 or above is acceptable, 0.8 or above good.

Reverse reversed items first.

---

## Alpha for the four scales

```r
q <- questionnaire
q$stress_4 <- 6 - q$stress_4
psych::alpha(q[, paste0("stress_", 1:6)])$total$raw_alpha
```

| Scale | stress | burnout | support | satisfaction |
|---|---|---|---|---|
| alpha | `r f2(alphas["stress"])` | `r f2(alphas["burnout"])` | `r f2(alphas["support"])` | `r f2(alphas["satisfaction"])` |

The full output shows alpha if each item were dropped, which spots a weak item.

---

## Why `psych::alpha()`

ggplot2 also has an `alpha()`, for transparent colours.

Whichever package is loaded last wins, so a plain `alpha()` can call the wrong one.

`package::function()` always calls the intended function.

---

## Reliability is not validity

A high alpha shows consistency.

A scale can be highly reliable and still measure the wrong thing: the target from Chapter 5.

The factor structure and relationships with other constructs are the evidence for validity.

---

## Reporting the questionnaire

> An exploratory factor analysis (maximum likelihood, oblimin rotation) supported four factors, as indicated by parallel analysis. All items loaded on their intended factor (loadings `r f2(min(main_loadings))` to `r f2(max(main_loadings))`), with one item (*burnout_3*) also loading on stress. Internal consistency was acceptable to good (α = `r f2(min(alphas))` to `r f2(max(alphas))`).

**Confirmatory factor analysis**, with the lavaan package, tests a structure decided in advance.

---

## Cluster analysis groups people

Factor analysis groups **variables**. Cluster analysis groups **people** with similar profiles.

It is descriptive: it tests no hypothesis.

Its groups are summaries of the data, not discoveries of natural kinds of student. RQ7 is exploratory.

---

## Each student's profile

```r
profiles <- semesters |>
  summarise(across(c(sleep_hours, study_hours, caffeine_mg, exercise_days),
                   ~ mean(.x, na.rm = TRUE)), .by = student_id) |>
  left_join(scores, join_by(student_id)) |>
  na.omit()

profile_data <- scale(profiles |> select(-student_id))
```

Average habits across the semesters, plus stress, support, and satisfaction.

---

## Scale before measuring distance

Clustering works with **distances** between students' profiles.

Caffeine is in hundreds of milligrams, sleep in single hours: unscaled, caffeine would dominate every distance.

`scale()` turns every variable into z-scores, so each counts equally.

---

## k-means

```{mermaid}
flowchart LR
    A["Place k points<br>at random"] --> B["Assign each student<br>to the nearest"] --> C["Move each point<br>to its students' mean"]
    C -->|repeat until stable| B
```

The result depends on the random start, so `nstart = 25` runs 25 starts and keeps the best.

---

## Choosing the number of clusters

| Guide | Looks at |
|---|---|
| elbow method | where more clusters stop reducing within-cluster distance |
| silhouette | how much closer each student is to their own cluster than the next, −1 to 1 |

```r
fviz_nbclust(profile_data, kmeans, method = "wss", nstart = 25)
fviz_nbclust(profile_data, kmeans, method = "silhouette", nstart = 25)
```

---

## The data does not decide clearly

```{r}
#| fig-width: 9
#| fig-height: 3.4
library(patchwork)
set.seed(123)
th <- theme_minimal(base_size = 13) + theme(plot.background = element_rect(fill = "transparent", colour = NA))
p1 <- fviz_nbclust(profile_data, kmeans, method = "wss", nstart = 25, linecolor = navy) + th + labs(title = "Elbow")
p2 <- fviz_nbclust(profile_data, kmeans, method = "silhouette", nstart = 25, linecolor = navy) + th + labs(title = "Silhouette")
p1 + p2
```

A gentle elbow; silhouette highest at 2, only a little lower at 3 or 4. Theory expects about four profiles.

---

## Four clusters

```r
set.seed(123)
clusters <- kmeans(profile_data, centers = 4, nstart = 25)
profiles |>
  mutate(cluster = clusters$cluster) |>
  summarise(students = n(),
            across(sleep_hours:satisfaction, ~ round(mean(.x), 1)),
            .by = cluster)
```

A cluster is useful only if its meaning can be stated: average each variable on the original scale.

---

## The cluster profiles

```{r}
#| results: asis
profiles |>
  mutate(cluster = clusters$cluster) |>
  summarise(students = n(), across(sleep_hours:satisfaction, ~ round(mean(.x), 1)), .by = cluster) |>
  arrange(cluster) |>
  rename(sleep = sleep_hours, study = study_hours, caffeine = caffeine_mg, exercise = exercise_days) |>
  knitr::kable()
```

---

## Naming the profiles

| Profile | Pattern |
|---|---|
| balanced | most sleep, moderate study, most exercise, low stress, good support |
| overloaded | long study weeks, little sleep and exercise, high stress |
| overloaded, high caffeine | the same, with very high caffeine |
| disengaged or isolated | low study, low support, low satisfaction |

Cluster numbers are arbitrary and can change with the seed.

---

## The clusters overlap

```{r}
#| fig-width: 8
#| fig-height: 3.9
fviz_cluster(clusters, data = profile_data, geom = "point", ellipse.type = "convex",
             palette = c(navy, copper, "#5b8c5a", "#8a5a9c"),
             ggtheme = theme_minimal(base_size = 13)) +
  labs(title = NULL) + theme(plot.background = element_rect(fill = "transparent", colour = NA))
```

Shown on the first two principal components. Real students do not fall into neat boxes.

---

## Hierarchical clustering

```r
tree <- hclust(dist(profile_data), method = "ward.D2")
fviz_dend(tree, k = 4, show_labels = FALSE, rect = TRUE)
```

Start with every student alone, merge the two most similar clusters, repeat until one remains.

The **dendrogram** can be cut at any height, without choosing k in advance. Ward's method keeps clusters compact.

---

## Two methods, compared

```r
hier_clusters <- cutree(tree, k = 4)
table(hierarchical = hier_clusters, kmeans = clusters$cluster)
```

They agree on the broad picture, most clearly on the balanced students.

Many students switch groups: they lie near a boundary. Trust the groups both methods find.

---

## Clusters in random data

```{r}
#| fig-width: 5.5
#| fig-height: 3.8
set.seed(7)
random_points <- tibble(x = rnorm(300), y = rnorm(300))
random_points$cluster <- factor(kmeans(random_points, centers = 4, nstart = 25)$cluster)
ggplot(random_points, aes(x, y, colour = cluster)) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(navy, copper, "#5b8c5a", "#8a5a9c")) +
  coord_equal() + theme(legend.position = "none")
```

300 random points with no groups: k-means still cuts four neat wedges.

---

## Clusters prove nothing on their own

Before reporting clusters as meaningful, check that:

- they make sense;
- different methods broadly agree;
- the groups differ in ways that matter for the research.

Chapter 14 allows overlapping groups, and students who fit no group.

---

## In your field: criminology and social policy

```r
arrests_pca <- prcomp(USArrests, scale. = TRUE)
summary(arrests_pca)$importance[, 1:2]
```

Assault, murder, and rape rates, and urban population, for 50 US states.

The first component, `r round(100 * summary(arrests_pca)$importance[2, 1])`%, is an overall violent crime rate; the second, `r round(100 * summary(arrests_pca)$importance[2, 2])`%, how urban a state is.

---

## Practical lab: the Chapter 9 playground

Work through the playground exercises in your browser, with hints and solutions.

Clustering and factor analysis run in the browser; every exercise also runs in RStudio.

---

## Practical exercises 1–3: components, factors, alpha

1. A PCA of the six support items: what does the first component capture?
2. An EFA with three factors: which scales merge, and why?
3. Alpha for stress without reversing `stress_4`: what happens?

---

## Practical exercises 4–6: clusters and noise

4. k-means with 3 clusters: which profiles merged?
5. Justify your number of clusters in two sentences for a thesis.
6. With noisier items (`sd = 2`), how many are needed for a correlation of 0.8?

---

## Try this yourself

Take a questionnaire from your own field.

- group its items by the construct each should measure;
- list the reversed items;
- plan the EFA: number of factors, rotation, and why;
- decide what alpha you would accept, and what you would do below it.

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| far fewer students in PCA | `na.omit()` on items with scattered gaps |
| a scale's alpha is unexpectedly low | a reversed item not recoded |
| `alpha()` returns a colour function | ggplot2's `alpha()`; use `psych::alpha()` |
| an item loads on two factors | a cross-loading; read its wording |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| caffeine dominates the clusters | variables not scaled |
| different clusters on every run | too few random starts; set `nstart` and a seed |
| the methods disagree on many students | overlapping groups; trust shared ones |
| clusters found in noise | clustering always finds groups |

---

## Completion checklist

- [ ] each construct has several items;
- [ ] the number of factors comes from parallel analysis;
- [ ] PCA and factor analysis are used for their own purposes;
- [ ] reversed items load negatively and are recoded;
- [ ] cross-loadings are discussed;
- [ ] alpha is reported for every scale;
- [ ] clustering variables are scaled;
- [ ] clusters are described and checked with a second method.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| PCA and factor analysis are the same | PCA summarises; FA models latent causes |
| a high alpha proves a scale is valid | alpha measures consistency only |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| clusters are natural groups | clustering finds groups even in random data |
| the software decides the number | criteria disagree; the choice is a justified judgement |

---

## The chapter in one sentence

> **Several items measure a construct better than one; factor analysis and alpha show whether they do, and clusters summarise people without proving natural groups.**

---

## Next: Chapter 10

The next chapter models data with a structure:

- repeated measurements of the same students;
- random intercepts and random slopes;
- change in wellbeing over two years;
- students within supervisors;
- missing data and mixed models.

---

## Questions

Which construct in your research is measured with a single item?

What would a second and third item add?
