---
title: "Introduction to Machine Learning in R"
description: "Chapter 11: prediction versus explanation, generalisation and overfitting, a fair test with tidymodels, recipes and workflows, evaluating classifiers, cross-validation, and tuning."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(tidymodels)
tidymodels_prefer()

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 14)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x)

true_curve  <- function(x) sin(2 * x)
make_points <- function(n) { x <- runif(n, 0, 3); tibble(x = x, y = true_curve(x) + rnorm(n, sd = 0.3)) }
set.seed(3)
train_points <- make_points(15)
new_points   <- make_points(300) |> filter(x >= min(train_points$x), x <= max(train_points$x))
grid <- tibble(x = seq(min(train_points$x), max(train_points$x), length.out = 300))
curves <- bind_rows(lapply(c(1, 3, 12), function(d) {
  fit <- lm(y ~ poly(x, d), data = train_points)
  tibble(grid, y = predict(fit, grid), model = paste("Degree", d))}))
curves$model <- factor(curves$model, levels = paste("Degree", c(1, 3, 12)))
rmse <- function(o, p) sqrt(mean((o - p)^2))
errors <- bind_rows(lapply(1:12, function(d) {
  fit <- lm(y ~ poly(x, d), data = train_points)
  tibble(degree = d, training = rmse(train_points$y, predict(fit, train_points)),
         new_data = rmse(new_points$y, predict(fit, new_points)))}))

scores <- questionnaire |>
  mutate(stress_4 = 6 - stress_4,
         stress = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
         burnout = rowMeans(pick(burnout_1:burnout_6), na.rm = TRUE),
         support = rowMeans(pick(support_1:support_6), na.rm = TRUE),
         satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)) |>
  select(student_id, stress, burnout, support, satisfaction)
dropout_data <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1) |> select(-semester), join_by(student_id)) |>
  mutate(considering_dropout = factor(considering_dropout, levels = c("Yes", "No"))) |>
  select(-student_id, -supervisor_id, -workshop, -workshop_sessions)
set.seed(2026)
dropout_split <- initial_split(dropout_data, prop = 0.75, strata = considering_dropout)
dropout_train <- training(dropout_split); dropout_test <- testing(dropout_split)
dropout_recipe <- recipe(considering_dropout ~ ., data = dropout_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
logistic_wf <- workflow() |> add_recipe(dropout_recipe) |> add_model(logistic_reg())
logistic_fit <- fit(logistic_wf, data = dropout_train)
test_results <- augment(logistic_fit, new_data = dropout_test)
acc <- accuracy(test_results, considering_dropout, .pred_class)$.estimate
always_no <- mean(dropout_test$considering_dropout == "No")
auc_test <- roc_auc(test_results, considering_dropout, .pred_Yes)$.estimate
knn1_fit <- fit(workflow() |> add_recipe(dropout_recipe) |>
                  add_model(nearest_neighbor(neighbors = 1) |> set_mode("classification")), dropout_train)
knn_train <- roc_auc(augment(knn1_fit, new_data = dropout_train), considering_dropout, .pred_Yes)$.estimate
knn_test  <- roc_auc(augment(knn1_fit, new_data = dropout_test), considering_dropout, .pred_Yes)$.estimate
set.seed(2026)
dropout_folds <- vfold_cv(dropout_train, v = 10, strata = considering_dropout)
logistic_cv <- fit_resamples(logistic_wf, resamples = dropout_folds, metrics = metric_set(roc_auc, accuracy))
cv_auc <- collect_metrics(logistic_cv) |> filter(.metric == "roc_auc")
tree_wf <- workflow() |> add_recipe(dropout_recipe) |>
  add_model(decision_tree(tree_depth = tune(), min_n = 10, cost_complexity = 0) |> set_mode("classification"))
set.seed(2026)
tree_tuning <- tune_grid(tree_wf, resamples = dropout_folds, grid = tibble(tree_depth = 1:10),
                         metrics = metric_set(roc_auc))
best_depth <- select_best(tree_tuning, metric = "roc_auc")
final_tree <- tree_wf |> finalize_workflow(best_depth) |>
  last_fit(dropout_split, metrics = metric_set(roc_auc, accuracy))
final_logistic <- last_fit(logistic_wf, dropout_split, metrics = metric_set(roc_auc, accuracy))
auc_tree <- collect_metrics(final_tree) |> filter(.metric == "roc_auc") |> pull(.estimate)
auc_lr   <- collect_metrics(final_logistic) |> filter(.metric == "roc_auc") |> pull(.estimate)
```

# Introduction to Machine Learning in R

Judged by one thing: how well it works on new cases

**Chapter 11**

Polla Fattah

---

## By the end of today you can

- tell supervised from unsupervised learning, and prediction from explanation;
- explain generalisation, overfitting, and the bias-variance trade-off;
- split data into training and test sets, and say why;
- prepare data with a recipe, and combine it with a model in a workflow;
- evaluate a classifier, and explain why accuracy can mislead;
- use cross-validation, tune hyperparameters, and run a final test.

---

## Some questions are about what will happen

- which patients are likely to be readmitted?
- which loans are likely to fail?
- which students are likely to struggle, early enough to help?

The university wants to identify at-risk students at the end of their first semester.

---

## Kinds of machine learning

| Kind | Data | Examples |
|---|---|---|
| supervised: classification | outcome is a category | considering dropout: yes or no |
| supervised: regression | outcome is a number | next semester's GPA |
| unsupervised | no outcome; find structure | clustering (Chapters 9, 14) |

Many methods are familiar statistics: logistic regression is also a classifier.

---

## Explanation and prediction

| | Explanation (Chapters 7–10) | Prediction (Chapters 11–16) |
|---|---|---|
| question | why does it happen? | what will happen for a new case? |
| judged by | sensible, well-estimated effects | accuracy on new data |
| typical models | simple, interpretable | anything that predicts well |
| main danger | confounding | overfitting |

A good predictor need not be a cause; a good explanation may predict poorly.

---

## Generalisation

Any dataset holds two things:

- the **pattern** that would appear again in new data;
- the **noise** that belongs to this sample only.

A model that learns the pattern **generalises**. One that also learns the noise **overfits**.

---

## Simulating overfitting

```r
true_curve  <- function(x) sin(2 * x)
make_points <- function(n) {
  x <- runif(n, 0, 3)
  tibble(x = x, y = true_curve(x) + rnorm(n, sd = 0.3))
}
train_points <- make_points(15)
new_points   <- make_points(300)

lm(y ~ poly(x, d), data = train_points)   # d = 1, 3, 12
```

15 noisy points from a wave, and three polynomials of increasing flexibility.

---

## Too simple, about right, too flexible

```{r}
#| fig-width: 9
#| fig-height: 3.5
ggplot() +
  geom_point(data = new_points, aes(x, y), colour = "grey75", size = 0.8) +
  geom_point(data = train_points, aes(x, y), size = 2.2, colour = navy) +
  geom_line(data = curves, aes(x, y), colour = copper, linewidth = 1.1) +
  facet_wrap(~ model) + coord_cartesian(ylim = c(-2, 2)) + labs(x = NULL, y = NULL)
```

The line **underfits**. Degree 3 follows the pattern. Degree 12 threads every training point and swings away from the new ones.

---

## The bias-variance trade-off

```{r}
#| fig-width: 8
#| fig-height: 3.5
errors |>
  tidyr::pivot_longer(-degree, names_to = "data", values_to = "error") |>
  ggplot(aes(x = degree, y = error, colour = data)) +
  geom_line(linewidth = 1.2) + geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12) + scale_y_log10() +
  scale_colour_manual(values = c(training = "grey55", new_data = copper),
                      labels = c(training = "Training points", new_data = "New points")) +
  labs(x = "Flexibility (degree)", y = "RMSE (log)", colour = NULL)
```

Training error keeps falling. Error on new data is lowest at degree `r errors$degree[which.min(errors$new_data)]`, then rises.

---

## Bias and variance

| Too simple | Too flexible |
|---|---|
| high **bias** | high **variance** |
| misses part of the real pattern | follows each sample's noise |
| wrong whatever the data | changes greatly between samples |

The only way to find the model in between is to measure performance on data it did not learn from.

---

## tidymodels designs a fair test

| Package | Job |
|---|---|
| rsample | split the data |
| recipes | prepare it |
| parsnip | specify models |
| workflows | combine recipe and model |
| tune | tune hyperparameters |
| yardstick | measure performance |

`library(tidymodels)`, then `tidymodels_prefer()` to settle name clashes.

---

## The workflow of a fair test

```{mermaid}
flowchart TB
  A[All data] --> B[Split]
  B --> C[Training data]
  B --> T[Test data, set aside]
  C --> D[Recipe] --> E[Fit and tune with<br>cross-validation] --> F[Final model]
  T --> G[Evaluate once]
  F --> G
```

All choices use the training data. The test data is used once, at the end.

---

## Use only what is known at prediction time

```r
dropout_data <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1) |> select(-semester),
            join_by(student_id)) |>
  mutate(considering_dropout = factor(considering_dropout,
                                      levels = c("Yes", "No"))) |>
  select(-student_id, -supervisor_id, -workshop, -workshop_sessions)
```

The first factor level is the event to predict, so `"Yes"` comes first.

The workshop came after semester 1: using it would be **data leakage**.

---

## An imbalanced outcome

```r
dropout_data |> count(considering_dropout) |> mutate(share = n / sum(n))
```

Only about `r round(100 * mean(dropout_data$considering_dropout == "Yes"))`% of students answer "Yes".

Rare outcomes are common in real prediction, from disease to fraud, and they fool anyone who judges by accuracy alone.

---

## Never judge a model on the data it learned from

```r
set.seed(2026)
dropout_split <- initial_split(dropout_data, prop = 0.75,
                               strata = considering_dropout)
dropout_train <- training(dropout_split)
dropout_test  <- testing(dropout_split)
```

`r nrow(dropout_train)` students to train, `r nrow(dropout_test)` locked away for the final test.

`strata` keeps the same share of "Yes" in both.

---

## A recipe prepares the data

```r
dropout_recipe <- recipe(considering_dropout ~ ., data = dropout_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

| Step | Does |
|---|---|
| `step_impute_median()` | fills missing numbers with the median |
| `step_dummy()` | turns categories into 0/1 columns |
| `step_normalize()` | turns numbers into z-scores |

---

## `prep()` and `bake()`

```r
dropout_recipe |>
  prep() |>
  bake(new_data = NULL) |>
  glimpse()
```

`prep()` estimates medians, means, and SDs from the training data; `bake()` applies them.

`gender` becomes `gender_Male`; `faculty` becomes four dummy columns.

---

## No peeking at the test data

The medians, means, and SDs come from the **training** data only.

They are applied unchanged to the test data.

Calculated from all the data, they would leak test information into the model. Recipes and workflows prevent this automatically.

---

## Model and workflow

```r
logistic_spec <- logistic_reg()

logistic_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(logistic_spec)

logistic_fit <- fit(logistic_wf, data = dropout_train)
```

parsnip specifies models the same way whatever package does the work.

A workflow keeps recipe and model together; `fit()` prepares and fits in one step.

---

## Predictions on the test students

```r
test_results <- augment(logistic_fit, new_data = dropout_test)
test_results |> accuracy(truth = considering_dropout, estimate = .pred_class)
```

`augment()` adds `.pred_class` and the probabilities `.pred_Yes`, `.pred_No`.

Accuracy: `r round(100 * acc)`%. It sounds good.

---

## Accuracy misleads for rare outcomes

| "Model" | Accuracy | At-risk students found |
|---|---|---|
| logistic regression | `r round(100 * acc)`% | some |
| always say "No" | `r round(100 * always_no)`% | none |

A "model" that ignores every predictor scores as well or better, and helps no one.

---

## ROC AUC

The probability that, of one student who considered dropping out and one who did not, the model ranks the first higher.

```r
test_results |> roc_auc(truth = considering_dropout, .pred_Yes)
```

| AUC | 0.5 | 1.0 | this model |
|---|---|---|---|
| meaning | coin toss | perfect | `r f2(auc_test)` |

Not fooled by imbalance. Chapter 12 goes further, including thresholds.

---

## Overfitting in real data: 1-nearest neighbour

```r
knn1_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(nearest_neighbor(neighbors = 1) |> set_mode("classification"))
```

| Data | AUC |
|---|---|
| training | `r f2(knn_train)` |
| test | `r f2(knn_test)` |

On training data each student's nearest neighbour is themselves. The model memorised; it did not learn.

---

## Comparing models without the test set

Models and settings need comparing during building.

- the training data would reward overfitting;
- the test set would be used up.

**Cross-validation** reuses the training data.

---

## Cross-validation

```{r}
#| fig-width: 7
#| fig-height: 3
expand.grid(fold = 1:5, round = 1:5) |>
  mutate(role = if_else(fold == round, "Evaluate", "Fit")) |>
  ggplot(aes(x = fold, y = factor(round, levels = 5:1), fill = role)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  scale_fill_manual(values = c(Evaluate = navy, Fit = "grey82")) +
  labs(x = "Fold", y = "Round", fill = NULL) + theme(panel.grid = element_blank())
```

Fit on four folds, evaluate on the fifth, rotate, average. Every student is evaluated once, by a model that never saw them.

---

## Cross-validation in tidymodels

```r
set.seed(2026)
dropout_folds <- vfold_cv(dropout_train, v = 10, strata = considering_dropout)

logistic_cv <- fit_resamples(logistic_wf, resamples = dropout_folds,
                             metrics = metric_set(roc_auc, accuracy))
collect_metrics(logistic_cv)
```

Cross-validated AUC ≈ `r f2(cv_auc$mean)` (standard error `r sprintf("%.3f", cv_auc$std_err)`): an honest estimate, without touching the test set.

---

## Parameters and hyperparameters

| | Learned from data? | Example |
|---|---|---|
| parameter | yes | regression coefficients |
| hyperparameter | no, chosen beforehand | k in k-NN, tree depth, polynomial degree |

**Tuning** tries several hyperparameter values and compares them with cross-validation.

---

## Tuning a decision tree's depth

```r
tree_spec <- decision_tree(tree_depth = tune(), min_n = 10,
                           cost_complexity = 0) |>
  set_mode("classification")

tree_tuning <- tune_grid(tree_wf, resamples = dropout_folds,
                         grid = tibble(tree_depth = 1:10),
                         metrics = metric_set(roc_auc))
```

A tree asks yes-or-no questions in a row. Its depth controls flexibility. `tune()` marks what to tune.

---

## Depth against AUC

```{r}
#| fig-width: 8
#| fig-height: 3.4
collect_metrics(tree_tuning) |>
  ggplot(aes(tree_depth, mean)) +
  geom_line(colour = navy, linewidth = 1.2) + geom_point(colour = navy, size = 2.5) +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Tree depth", y = "Cross-validated AUC")
```

Depth 1 is chance; AUC rises, then levels off and dips as deep trees overfit. `select_best()` picks depth `r best_depth$tree_depth`.

---

## The final test

```r
final_tree <- tree_wf |>
  finalize_workflow(best_depth) |>
  last_fit(dropout_split, metrics = metric_set(roc_auc, accuracy))

final_logistic <- last_fit(logistic_wf, dropout_split,
                           metrics = metric_set(roc_auc, accuracy))
```

`last_fit()` fits on the whole training set and evaluates once on the test set.

---

## The simple model wins

| Model | Test AUC |
|---|---|
| tuned decision tree | `r f2(auc_tree)` |
| logistic regression | `r f2(auc_lr)` |

Dropout risk rises steadily with stress and falls with support: a smooth model captures that; a tree cuts it into boxes.

Try a simple model first, and make a complex one earn its place.

---

## Predictions about people

- a student flagged by mistake may be treated differently;
- a student who is missed receives no help;
- a good overall AUC can hide unfairness to one group;
- the right use here is to offer support earlier, never to penalise.

Check the model separately for women and men, part-time and full-time students.

---

## In your field: business and finance

```r
credit_wf <- workflow() |>
  add_recipe(recipe(Status ~ ., data = training(credit_split)) |>
               step_impute_median(all_numeric_predictors()) |>
               step_impute_mode(all_nominal_predictors()) |>
               step_dummy(all_nominal_predictors())) |>
  add_model(logistic_reg())
last_fit(credit_wf, credit_split) |> collect_metrics()
```

Loan applications and whether they turned out well. Only the data changed.

The **Brier score** measures probability quality: 0 is perfect.

---

## Practical lab: the Chapter 11 playground

Work through the playground exercises in your browser, with hints and solutions.

The browser version uses base R for AUC and cross-validation; the download uses tidymodels.

---

## Practical exercises 1–3: data and models

1. Remove the questionnaire scores: how much does the AUC fall?
2. Split 80/20 with another seed: how much does the test AUC move?
3. Tune k-NN over 5, 11, 21, 41 neighbours and compare with logistic regression.

---

## Practical exercises 4–6: judgement

4. What does `step_zv()` do, and when is it useful?
5. Explain to an administrator why 85% accuracy may be useless.
6. With 50 training points, how does the new-data error curve change?

---

## Try this yourself

Think of a prediction your own field needs.

- what outcome, and when must the prediction be made?
- which variables would be known at that moment?
- how rare is the outcome, and which measure would you use?
- what would the cost of each kind of error be?

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| a near-perfect training score | overfitting; judge on new data |
| a model better than seems possible | data leakage from the future |
| high accuracy, no cases found | an imbalanced outcome |
| the event level is predicted backwards | the outcome factor's first level is not the event |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| test set used to pick the model | it is no longer a test; use cross-validation |
| preprocessing on all the data | test information leaked; use a recipe |
| unequal "Yes" shares in train and test | split not stratified |
| a complex model loses to a simple one | the pattern is smooth; that is fine |

---

## Completion checklist

- [ ] predictors are known at prediction time;
- [ ] the outcome's first level is the event;
- [ ] the split is stratified and the test set locked away;
- [ ] preparation lives in a recipe;
- [ ] models are compared by cross-validation;
- [ ] hyperparameters are tuned, not guessed;
- [ ] the test set is used once;
- [ ] AUC, not only accuracy, is reported.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| a good training fit predicts well | only new-data performance counts |
| more complex is better | flexibility adds variance |
| high accuracy means a good classifier | not for imbalanced outcomes |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| the test set can choose between models | choose by cross-validation; test once |
| a good predictor is a cause | prediction needs association only |

---

## The chapter in one sentence

> **A predictive model is only as good as its performance on data it has never seen, so design the test before you build the model.**

---

## Next: Chapter 12

The next chapter compares classifiers:

- decision trees and random forests;
- k-nearest neighbours and support vector machines;
- comparing models fairly;
- the confusion matrix and choosing a threshold;
- imbalanced outcomes and fairness.

---

## Questions

If a model flagged students at risk, who should see the flag?

What would you do for a student it missed?
