---
title: "Predictive Regression"
description: "Chapter 13: measuring prediction error, overfitting with many predictors, ridge, lasso, and elastic net, boosting with XGBoost, model comparison, and the final test."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidymodels)
tidymodels_prefer()

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 13)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x); f3 <- function(x) sprintf("%.3f", x)

year_one <- semesters |> filter(semester <= 2) |>
  pivot_wider(id_cols = student_id, names_from = semester,
              values_from = gpa:wellbeing, names_glue = "{.value}_s{semester}")
final_gpa <- semesters |> filter(semester == 4, !is.na(gpa)) |> select(student_id, final_gpa = gpa)
gpa_data <- students |>
  left_join(questionnaire, join_by(student_id)) |>
  left_join(year_one, join_by(student_id)) |>
  inner_join(final_gpa, join_by(student_id)) |>
  select(-student_id, -supervisor_id, -considering_dropout)
set.seed(2026)
gpa_split <- initial_split(gpa_data, prop = 0.75, strata = final_gpa)
gpa_train <- training(gpa_split); gpa_test <- testing(gpa_split)
set.seed(2026)
gpa_folds <- vfold_cv(gpa_train, v = 10, strata = final_gpa)
gpa_recipe <- recipe(final_gpa ~ ., data = gpa_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
n_pred <- ncol(bake(prep(gpa_recipe), new_data = NULL)) - 1
regression_metrics <- metric_set(rmse, mae, rsq)

set.seed(2026)
null_cv <- fit_resamples(workflow(gpa_recipe, null_model(mode = "regression")),
                         resamples = gpa_folds, metrics = metric_set(rmse, mae))
null_rmse <- collect_metrics(null_cv) |> filter(.metric == "rmse") |> pull(mean)
lm_wf <- workflow(gpa_recipe, linear_reg())
set.seed(2026)
lm_cv <- fit_resamples(lm_wf, resamples = gpa_folds, metrics = regression_metrics)
lm_cv_rmse <- collect_metrics(lm_cv) |> filter(.metric == "rmse") |> pull(mean)
set.seed(3)
small_train <- gpa_train |> slice_sample(n = 60)
lm_small <- fit(lm_wf, data = small_train)
small_train_rmse <- augment(lm_small, new_data = small_train) |> rmse(final_gpa, .pred) |> pull(.estimate)
small_test_rmse  <- augment(lm_small, new_data = gpa_test) |> rmse(final_gpa, .pred) |> pull(.estimate)
lasso_small <- workflow(gpa_recipe, linear_reg(penalty = 0.03, mixture = 1) |> set_engine("glmnet")) |>
  fit(data = small_train)
lasso_small_rmse <- augment(lasso_small, new_data = gpa_test) |> rmse(final_gpa, .pred) |> pull(.estimate)
coefficient_comparison <- tidy(lm_small) |> select(term, ordinary = estimate) |>
  inner_join(tidy(lasso_small) |> select(term, lasso = estimate), join_by(term)) |>
  filter(term != "(Intercept)")
n_nonzero_small <- sum(coefficient_comparison$lasso != 0)
coef_of <- function(model, term) f2(coefficient_comparison[[model]][coefficient_comparison$term == term])

glmnet_wf <- workflow(gpa_recipe, linear_reg(penalty = tune(), mixture = tune()) |> set_engine("glmnet"))
glmnet_grid <- expand_grid(penalty = 10^seq(-4, 0, length.out = 20), mixture = c(0, 0.5, 1))
set.seed(2026)
glmnet_tuning <- tune_grid(glmnet_wf, resamples = gpa_folds, grid = glmnet_grid, metrics = metric_set(rmse))
best_glmnet <- show_best(glmnet_tuning, metric = "rmse", n = 1)
best_penalty <- select_best(glmnet_tuning, metric = "rmse")
lasso_fit <- glmnet_wf |> finalize_workflow(best_penalty) |> fit(data = gpa_train)
lasso_coefs <- tidy(lasso_fit)
n_kept <- sum(lasso_coefs$estimate != 0 & lasso_coefs$term != "(Intercept)")
n_all  <- sum(lasso_coefs$term != "(Intercept)")

xgb_wf <- workflow(gpa_recipe, boost_tree(trees = 500, learn_rate = tune(), tree_depth = tune()) |>
                     set_engine("xgboost") |> set_mode("regression"))
set.seed(2026)
xgb_tuning <- tune_grid(xgb_wf, resamples = gpa_folds,
                        grid = expand_grid(learn_rate = c(0.01, 0.03, 0.1), tree_depth = c(1, 2, 4)),
                        metrics = metric_set(rmse))
best_xgb <- show_best(xgb_tuning, metric = "rmse", n = 1)
comparison <- bind_rows(
  collect_metrics(null_cv) |> mutate(model = "baseline (mean)"),
  collect_metrics(lm_cv) |> mutate(model = "linear regression"),
  show_best(glmnet_tuning |> filter_parameters(mixture == 0), metric = "rmse", n = 1) |> mutate(model = "ridge"),
  show_best(glmnet_tuning |> filter_parameters(mixture == 0.5), metric = "rmse", n = 1) |> mutate(model = "elastic net"),
  show_best(glmnet_tuning |> filter_parameters(mixture == 1), metric = "rmse", n = 1) |> mutate(model = "lasso"),
  show_best(xgb_tuning, metric = "rmse", n = 1) |> mutate(model = "XGBoost")) |>
  filter(.metric == "rmse") |> select(model, rmse = mean, std_err) |> arrange(rmse)

final_lasso <- glmnet_wf |> finalize_workflow(best_penalty) |> last_fit(gpa_split, metrics = regression_metrics)
final_m <- collect_metrics(final_lasso)
fm <- function(m) final_m$.estimate[final_m$.metric == m]
gpa_only_fit <- last_fit(workflow(recipe(final_gpa ~ gpa_s1 + gpa_s2, data = gpa_train), linear_reg()),
                         gpa_split, metrics = regression_metrics)
gpa_only_rmse <- collect_metrics(gpa_only_fit) |> filter(.metric == "rmse") |> pull(.estimate)

data(concrete, package = "modeldata")
set.seed(1)
concrete_split <- initial_split(concrete)
lm_concrete_fit <- last_fit(workflow(compressive_strength ~ ., linear_reg()), concrete_split)
xgb_concrete_fit <- last_fit(workflow(compressive_strength ~ .,
  boost_tree(trees = 500, learn_rate = 0.05, tree_depth = 4) |> set_engine("xgboost") |> set_mode("regression")),
  concrete_split)
concrete_rmse <- function(f) collect_metrics(f) |> filter(.metric == "rmse") |> pull(.estimate)
```

# Predictive Regression

Many predictors, few cases, and a number to predict

**Chapter 13**

Polla Fattah

---

## By the end of today you can

- measure numeric predictions with RMSE, MAE, and R², against a baseline;
- explain why ordinary regression overfits with many predictors;
- fit and tune ridge, lasso, and elastic net regression;
- read which predictors the lasso keeps, and what that does not mean;
- explain how boosting builds a model from small trees, and tune XGBoost;
- choose a model by cross-validation, and report the final test.

---

## Many predictors, few cases

A questionnaire, several rounds of measurement, and a background survey can supply 50 predictors for 60 people.

Ordinary regression estimates a coefficient for each, fits its own data almost perfectly, and predicts new cases badly.

Correlated predictors, like questionnaire items, make it worse.

---

## The question: final GPA from year one

The university's advisers meet students at the end of year one.

| May use | May not use |
|---|---|
| background | anything after year one |
| all 22 questionnaire items | the workshop's later effects |
| semester 1 and 2 records | |

The outcome, GPA in semester 4, is a number: a **regression** problem.

---

## One row per student, wide

```r
year_one <- semesters |>
  filter(semester <= 2) |>
  pivot_wider(id_cols = student_id, names_from = semester,
              values_from = gpa:wellbeing,
              names_glue = "{.value}_s{semester}")
```

`names_glue` builds names such as `gpa_s1` and `sleep_hours_s2`.

---

## Only students who stayed have a final GPA

```r
gpa_data <- students |>
  left_join(questionnaire, join_by(student_id)) |>
  left_join(year_one, join_by(student_id)) |>
  inner_join(final_gpa, join_by(student_id)) |>
  select(-student_id, -supervisor_id, -considering_dropout)
```

`inner_join()` keeps students with a final GPA: `r nrow(students) - nrow(gpa_data)` who left are excluded.

The model predicts only for students who stay, and must not be used to judge the leavers.

---

## Split, folds, and recipe

```r
gpa_split <- initial_split(gpa_data, prop = 0.75, strata = final_gpa)
gpa_folds <- vfold_cv(gpa_train, v = 10, strata = final_gpa)

gpa_recipe <- recipe(final_gpa ~ ., data = gpa_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

With a numeric outcome, `strata` samples within quartiles. After dummies: `r n_pred` predictors.

---

## Errors of a numeric prediction

| Actual | Predicted | Error |
|---|---|---|
| 3.2 | 3.0 | 0.2 |
| 2.8 | 2.9 | −0.1 |
| 3.6 | 3.3 | 0.3 |
| 3.0 | 3.1 | −0.1 |
| 2.5 | 2.9 | −0.4 |

Each **error** is actual minus predicted.

---

## Three summaries of the errors

| Measure | Definition | Here |
|---|---|---|
| MAE | average size of the errors | 0.22 |
| RMSE | square root of the average squared error | larger; punishes big misses |
| R² | squared correlation of predicted and actual | share captured, 0 to 1 |

```r
regression_metrics <- metric_set(rmse, mae, rsq)
```

MAE and RMSE are in GPA points, lower is better: "typically off by about 0.2".

---

## Always compare with a baseline

```r
null_wf <- workflow(gpa_recipe, null_model(mode = "regression"))
null_cv <- fit_resamples(null_wf, resamples = gpa_folds,
                         metrics = metric_set(rmse, mae))
```

Predict the training mean for everyone: RMSE `r f2(null_rmse)`.

Any useful model must do clearly better.

---

## Ordinary regression with every predictor

```r
lm_wf <- workflow(gpa_recipe, linear_reg())
lm_cv <- fit_resamples(lm_wf, resamples = gpa_folds,
                       metrics = regression_metrics)
```

Cross-validated RMSE `r f3(lm_cv_rmse)`: much better than the baseline.

With `r nrow(gpa_train)` students and `r n_pred` predictors, it works reasonably. Theses often have 60.

---

## The same model on 60 students

```r
set.seed(3)
small_train <- gpa_train |> slice_sample(n = 60)
lm_small <- fit(lm_wf, data = small_train)
```

| Evaluated on | RMSE |
|---|---|
| its own 60 students | `r f2(small_train_rmse)` |
| new students | `r f2(small_test_rmse)` |
| baseline (the mean) | `r f2(null_rmse)` |

Excellent on its own data, worse than predicting the average on new data.

---

## Multicollinearity

Six stress items, two semesters of GPA: many predictors are strongly correlated.

Regression struggles to divide credit between them.

Coefficients become large, unstable, even of opposite signs, cancelling out on training data but not on new data.

---

## Regularisation penalises large coefficients

$$
\text{sum of squared errors} + \lambda \times \text{size of the coefficients}
$$

| $\lambda$ | Effect |
|---|---|
| 0 | ordinary regression |
| larger | coefficients **shrunk** towards zero |

A little shrinkage costs almost no training fit and buys much stability. Predictors must be normalised.

---

## Ridge, lasso, and elastic net

| Method | Size measured as | Effect |
|---|---|---|
| ridge | sum of squared coefficients | shrinks all; keeps every predictor |
| lasso | sum of absolute coefficients | shrinks, and sets some to exactly zero |
| elastic net | a mix, `mixture` 0 to 1 | between the two |

In tidymodels: `linear_reg(penalty = , mixture = )` with the glmnet engine.

---

## A lasso on the same 60 students

```r
lasso_small <- workflow(gpa_recipe,
                        linear_reg(penalty = 0.03, mixture = 1) |>
                          set_engine("glmnet")) |>
  fit(data = small_train)
```

| Model on 60 students | RMSE on new students |
|---|---|
| ordinary regression | `r f2(small_test_rmse)` |
| lasso | `r f2(lasso_small_rmse)` |

The penalty kept the model from chasing noise.

---

## Shrinkage, coefficient by coefficient

```{r}
#| fig-width: 7
#| fig-height: 3.8
ggplot(coefficient_comparison, aes(x = ordinary, y = lasso)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_point(size = 2.5, alpha = 0.7, colour = navy) +
  labs(x = "Ordinary regression coefficient", y = "Lasso coefficient")
```

The lasso keeps `r n_nonzero_small` of `r nrow(coefficient_comparison)` predictors and shrinks even those.

---

## Coefficients that make no sense

| Predictor | Ordinary regression | Lasso |
|---|---|---|
| semester 1 GPA | `r coef_of("ordinary", "gpa_s1")` | `r coef_of("lasso", "gpa_s1")` |
| semester 2 GPA | `r coef_of("ordinary", "gpa_s2")` | `r coef_of("lasso", "gpa_s2")` |
| semester 2 caffeine | `r coef_of("ordinary", "caffeine_mg_s2")` | `r coef_of("lasso", "caffeine_mg_s2")` |

With 60 students, ordinary regression fits chance patterns. Shrunken coefficients cannot chase one sample's noise.

---

## Tuning penalty and mixture

```r
glmnet_wf <- workflow(gpa_recipe,
  linear_reg(penalty = tune(), mixture = tune()) |> set_engine("glmnet"))

glmnet_grid <- expand_grid(penalty = 10^seq(-4, 0, length.out = 20),
                           mixture = c(0, 0.5, 1))

glmnet_tuning <- tune_grid(glmnet_wf, resamples = gpa_folds,
                           grid = glmnet_grid, metrics = metric_set(rmse))
```

20 penalties on a log scale, for ridge, an even elastic net, and lasso.

---

## Too little, just right, too much

```{r}
#| fig-width: 8
#| fig-height: 3.6
collect_metrics(glmnet_tuning) |>
  mutate(mixture = factor(mixture, labels = c("ridge (0)", "elastic net (0.5)", "lasso (1)"))) |>
  ggplot(aes(penalty, mean, colour = mixture)) +
  geom_line(linewidth = 1.1) + geom_point(size = 1.5) +
  scale_x_log10() + scale_colour_manual(values = c("grey55", copper, navy)) +
  labs(x = "Penalty (log scale)", y = "Cross-validated RMSE", colour = NULL)
```

At the far right the lasso has zeroed every coefficient: back to the baseline. Best: mixture `r best_glmnet$mixture`, penalty `r signif(best_glmnet$penalty, 2)`.

---

## What the lasso keeps

```r
best_penalty <- select_best(glmnet_tuning, metric = "rmse")
lasso_fit <- glmnet_wf |> finalize_workflow(best_penalty) |> fit(data = gpa_train)
tidy(lasso_fit) |> filter(estimate != 0) |> arrange(desc(abs(estimate)))
```

Of `r n_all` predictors, the lasso keeps `r n_kept`. Year-one GPA dominates.

One SD higher semester 2 GPA: about `r f2(lasso_coefs$estimate[lasso_coefs$term == "gpa_s2"])` points higher final GPA.

---

## The lasso path

```{r}
#| fig-width: 8
#| fig-height: 3.6
lasso_path <- workflow(gpa_recipe, linear_reg(penalty = 0.01, mixture = 1) |> set_engine("glmnet")) |>
  fit(data = gpa_train) |> extract_fit_engine() |> tidy() |> filter(term != "(Intercept)")
ggplot(lasso_path, aes(x = lambda, y = estimate, group = term)) +
  geom_line(colour = "grey75") +
  geom_line(data = filter(lasso_path, term %in% c("gpa_s1", "gpa_s2")), aes(colour = term), linewidth = 1.2) +
  geom_vline(xintercept = best_penalty$penalty, linetype = "dashed", colour = soft) +
  scale_x_continuous(transform = scales::transform_compose("log10", "reverse")) +
  scale_colour_manual(values = c(gpa_s1 = copper, gpa_s2 = navy)) +
  labs(x = "Penalty (decreasing)", y = "Coefficient", colour = NULL)
```

As the penalty falls, predictors enter one by one. The two GPA measures enter first.

---

## Selected is not the same as causal

Sleep affects GPA (Chapter 8), yet the lasso drops it.

Its effect is already visible in year-one GPA, so it adds little to the **prediction**.

Among correlated predictors the lasso keeps one, and which one can change between samples. Predict with the lasso; explain with Chapters 8 and 10.

---

## Boosting

```{mermaid}
flowchart LR
    A["Start:<br>the average"] --> B["Fit a small tree<br>to the errors"] --> C["Add it,<br>scaled down"]
    C -->|repeat hundreds of times| B
```

Each tree learns what the model so far missed. The **learning rate** keeps each step small.

Many weak trees add up to a strong model.

---

## Boosting with one predictor

```{r}
#| fig-width: 9
#| fig-height: 3.4
boost_data <- gpa_train |> filter(!is.na(gpa_s2))
boost_grid <- tibble(gpa_s2 = seq(min(boost_data$gpa_s2), max(boost_data$gpa_s2), length.out = 200))
set.seed(2026)
boost_curves <- purrr::map(c(1, 10, 100), \(n) {
  fitted <- boost_tree(trees = n, tree_depth = 1, learn_rate = 0.3) |>
    set_engine("xgboost") |> set_mode("regression") |> fit(final_gpa ~ gpa_s2, data = boost_data)
  predict(fitted, boost_grid) |> bind_cols(boost_grid) |>
    mutate(trees = paste(n, if (n == 1) "tree" else "trees"))
}) |> purrr::list_rbind() |> mutate(trees = factor(trees, levels = c("1 tree", "10 trees", "100 trees")))
ggplot(boost_data, aes(gpa_s2, final_gpa)) +
  geom_point(colour = "grey75", size = 0.8) +
  geom_step(data = boost_curves, aes(y = .pred), colour = copper, linewidth = 1.2) +
  facet_wrap(~ trees) + labs(x = "Semester 2 GPA", y = "Final GPA")
```

One step, a staircase, then a curve that follows the data.

---

## Forests and boosting

| | Random forest | Boosting |
|---|---|---|
| trees grown | independently | in sequence |
| combined by | averaging | adding corrections |
| main settings | trees, `mtry` | trees, learning rate, depth |

**XGBoost** is a fast, popular implementation, among the most successful methods for tables of data.

---

## Tuning XGBoost

```r
xgb_wf <- workflow(gpa_recipe,
  boost_tree(trees = 500, learn_rate = tune(), tree_depth = tune()) |>
    set_engine("xgboost") |> set_mode("regression"))

tune_grid(xgb_wf, resamples = gpa_folds,
          grid = expand_grid(learn_rate = c(0.01, 0.03, 0.1),
                             tree_depth = c(1, 2, 4)))
```

Best: depth `r best_xgb$tree_depth`, RMSE `r f3(best_xgb$mean)`.

Depth-1 trees add separate effects with no interactions. Shallow winning means few interactions to find.

---

## Comparing the models

```{r}
#| results: asis
comparison |> mutate(rmse = f3(rmse), std_err = f3(std_err)) |>
  rename(`cross-validated RMSE` = rmse, `std. error` = std_err) |> knitr::kable()
```

Lasso and elastic net on top, XGBoost and ridge close behind, all better than ordinary regression. The differences are within a standard error; the lasso is chosen for using only `r n_kept` predictors.

---

## The final test

```r
final_lasso <- glmnet_wf |>
  finalize_workflow(best_penalty) |>
  last_fit(gpa_split, metrics = regression_metrics)
```

| MAE | RMSE | R² |
|---|---|---|
| `r f2(fm("mae"))` | `r f2(fm("rmse"))` | `r f2(fm("rsq"))` |

Off by `r f2(fm("mae"))` GPA points on average, capturing `r round(100 * fm("rsq"))`% of the variation.

---

## Predicted against actual

```{r}
#| fig-width: 5
#| fig-height: 4
collect_predictions(final_lasso) |>
  ggplot(aes(.pred, final_gpa)) +
  geom_abline(linetype = "dashed", colour = "grey55") +
  geom_point(alpha = 0.6, colour = navy) +
  coord_obs_pred() + labs(x = "Predicted final GPA", y = "Actual final GPA")
```

Predictions are less spread than reality: extreme students are pulled towards the average.

---

## How much does year one add to past grades?

```r
gpa_only_wf <- workflow(recipe(final_gpa ~ gpa_s1 + gpa_s2, data = gpa_train),
                        linear_reg())
```

| Model | Test RMSE |
|---|---|
| lasso with everything | `r f3(fm("rmse"))` |
| two year-one GPAs only | `r f3(gpa_only_rmse)` |

Past grades are by far the best predictor. Sleep, stress, and support matter (Chapter 8), but their influence is already in the grades.

---

## In your field: engineering

```r
lm_concrete  <- workflow(compressive_strength ~ ., linear_reg())
xgb_concrete <- workflow(compressive_strength ~ .,
  boost_tree(trees = 500, learn_rate = 0.05, tree_depth = 4) |>
    set_engine("xgboost") |> set_mode("regression"))
```

Concrete strength from its recipe and age: curves and interactions.

Boosting cuts RMSE from `r sprintf("%.1f", concrete_rmse(lm_concrete_fit))` to `r sprintf("%.1f", concrete_rmse(xgb_concrete_fit))` MPa. Flexible models win when the data has structure for them.

---

## Practical lab: the Chapter 13 playground

Work through the playground exercises in your browser, with hints and solutions.

The browser version builds ridge regression by formula; the download uses glmnet and XGBoost.

---

## Practical exercises 1–3: errors and shrinkage

1. MAE and RMSE by hand; change one prediction by 1.0 and compare.
2. Repeat the 60-student experiment with 150 students.
3. Ridge against lasso coefficients: how many are exactly zero?

---

## Practical exercises 4–6: design and communication

4. Scale scores instead of 22 items: does the lasso's RMSE change?
5. XGBoost with 100, 500, and 1000 trees at learning rate 0.03.
6. Explain an RMSE of 0.2 to academic advisers in two sentences.

---

## Try this yourself

Take a prediction problem from your own field.

- count your cases and your predictors;
- name a baseline prediction;
- decide whether a lasso, a ridge, or boosting suits the structure;
- plan how you would report the final test.

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| near-perfect training fit, poor new predictions | too many predictors for the cases |
| coefficients huge and of opposite signs | multicollinearity |
| regularisation changes nothing | predictors not normalised |
| the lasso predicts the mean for everyone | penalty too large |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| a different lasso selection each sample | correlated predictors; selection is unstable |
| the lasso "drops" a known cause | its effect is carried by another predictor |
| RMSE reported without a baseline | the number cannot be judged |
| extremes predicted too close to the mean | normal for imperfect models |

---

## Completion checklist

- [ ] predictors are known at prediction time;
- [ ] a baseline is reported;
- [ ] RMSE, MAE, and R² are reported;
- [ ] predictors are normalised before regularising;
- [ ] penalty and mixture are tuned by cross-validation;
- [ ] selected predictors are not described as causes;
- [ ] the test set is used once.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| more predictors always help | each is another chance to fit noise |
| the lasso's predictors are what matters | they help prediction, chosen partly by chance |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| a low RMSE means every prediction is accurate | extremes are pulled to the average |
| boosting always beats regression | only when curves and interactions exist |

---

## The chapter in one sentence

> **With many predictors, shrink the coefficients or build from small trees, compare against a baseline by cross-validation, and never read a selected predictor as a cause.**

---

## Next: Chapter 14

The next chapter revisits clustering:

- what counts as a group;
- Gaussian mixture models and soft membership;
- clusters as dense regions: DBSCAN;
- evaluating a clustering;
- reporting the choices behind the clusters.

---

## Questions

How many predictors, and how many cases, would your model have?

What is the simplest prediction your model must beat?
