---
title: "Advanced Clustering"
description: "Chapter 14: three definitions of a group, Gaussian mixture models with BIC and membership probabilities, DBSCAN and noise points, and evaluating a clustering."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(mclust)
library(dbscan)
library(cluster)

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"; green <- "#5b8c5a"; purple <- "#8a5a9c"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 13)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x)

q <- questionnaire; q$stress_4 <- 6 - q$stress_4
scores <- q |>
  mutate(stress = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
         support = rowMeans(pick(support_1:support_6), na.rm = TRUE),
         satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)) |>
  select(student_id, stress, support, satisfaction)
profiles <- semesters |>
  summarise(across(c(sleep_hours, study_hours, caffeine_mg, exercise_days), ~ mean(.x, na.rm = TRUE)),
            .by = student_id) |>
  left_join(scores, join_by(student_id)) |> na.omit()
profile_data <- scale(profiles |> select(-student_id))
set.seed(123)
kmeans_clusters <- kmeans(profile_data, centers = 4, nstart = 25)

set.seed(8)
shapes <- bind_rows(
  tibble(x = rnorm(100, 0, 0.4), y = rnorm(100, 0, 0.4), truth = "Round group"),
  tibble(x = rnorm(100, 3, 2), y = rnorm(100, 1.4, 0.2), truth = "Long group"),
  tibble(x = runif(15, -2, 7), y = runif(15, -2, 4), truth = "Scattered"))
xy <- shapes |> select(x, y)
shape_clusters <- bind_rows(
  shapes |> mutate(method = "k-means", cluster = factor(kmeans(xy, 2, nstart = 25)$cluster)),
  shapes |> mutate(method = "Mixture model", cluster = factor(Mclust(xy, G = 2, verbose = FALSE)$classification)),
  shapes |> mutate(method = "DBSCAN", cluster = factor(dbscan(xy, eps = 0.5, minPts = 5)$cluster))) |>
  mutate(method = factor(method, levels = c("k-means", "Mixture model", "DBSCAN")),
         cluster = forcats::fct_recode(cluster, noise = "0"))
agreement <- shape_clusters |> filter(truth != "Scattered") |>
  summarise(ari = adjustedRandIndex(truth, cluster), .by = method)
ari_of <- function(m) f2(agreement$ari[agreement$method == m])
db_noise_found <- shape_clusters |> filter(method == "DBSCAN", truth == "Scattered") |>
  summarise(n = sum(cluster == "noise")) |> pull(n)

geyser <- Mclust(faithful$waiting, verbose = FALSE)
gp <- geyser$parameters
geyser_z <- round(predict(geyser, newdata = c(50, 67, 85))$z, 2)

set.seed(123)
profile_gmm <- Mclust(profile_data, G = 1:8, verbose = FALSE)
bic_gap <- round(max(profile_gmm$BIC, na.rm = TRUE) - max(profile_gmm$BIC["4", ], na.rm = TRUE), 1)
profiles <- profiles |> mutate(gmm_cluster = profile_gmm$classification)
certainty <- apply(profile_gmm$z, 1, max)

profile_db <- dbscan(profile_data, eps = 2, minPts = 8)
n_noise <- sum(profile_db$cluster == 0); n_db_clusters <- max(profile_db$cluster)
noise_rows <- profiles |> filter(profile_db$cluster == 0)
n_extreme_work <- sum(noise_rows$study_hours > 55); n_extreme_rest <- sum(noise_rows$study_hours < 10)

profile_distances <- dist(profile_data)
sil_km  <- mean(silhouette(kmeans_clusters$cluster, profile_distances)[, "sil_width"])
sil_gmm <- mean(silhouette(profile_gmm$classification, profile_distances)[, "sil_width"])
ari_methods <- adjustedRandIndex(kmeans_clusters$cluster, profile_gmm$classification)
quake_db <- dbscan(quakes[, c("long", "lat")], eps = 1.5, minPts = 10)
```

# Advanced Clustering

Every method carries a definition of what a group is

**Chapter 14**

Polla Fattah

---

## By the end of today you can

- explain three definitions of a group;
- name the limits of k-means: hard assignments, round clusters, no outliers;
- fit a Gaussian mixture model, choose the number of clusters with BIC, and read membership probabilities;
- find dense clusters and noise points with DBSCAN, and choose its settings;
- evaluate a clustering with the silhouette, the adjusted Rand index, and interpretability.

---

## Three definitions of a group

| A group is | Method | Allows |
|---|---|---|
| cases close to a common centre | k-means | nothing else |
| a population with its own distribution | Gaussian mixture model | partial membership |
| a dense region separated by sparse ones | DBSCAN | cases in no group |

The definition decides what the method can find.

---

## Questions Chapter 9 left open

- how sure can we be which profile a student belongs to?
- is four really the right number?
- do some students fit no profile at all?

Same profile data: average habits across semesters, plus stress, support, satisfaction, scaled to z-scores.

---

## Three assumptions of k-means

| Assumption | Problem |
|---|---|
| every case in exactly one cluster, with certainty | a student halfway between is assigned as firmly as one at the centre |
| clusters round and similar in size | elongated or unequal groups are cut wrongly |
| every case in some cluster | no way to say "fits nowhere" |

Real groups of people rarely satisfy these.

---

## A simulation with known structure

```r
shapes <- bind_rows(
  tibble(x = rnorm(100, 0, 0.4), y = rnorm(100, 0, 0.4), truth = "Round group"),
  tibble(x = rnorm(100, 3, 2),   y = rnorm(100, 1.4, 0.2), truth = "Long group"),
  tibble(x = runif(15, -2, 7),   y = runif(15, -2, 4),     truth = "Scattered"))

kmeans(xy, 2, nstart = 25)
Mclust(xy, G = 2)
dbscan(xy, eps = 0.5, minPts = 5)
```

A round group, a long thin group, and 15 points that belong to neither.

---

## Three methods, three answers

```{r}
#| fig-width: 10
#| fig-height: 2.9
ggplot(shape_clusters, aes(x, y, colour = cluster)) +
  geom_point(size = 1.2) + facet_wrap(~ method) +
  scale_colour_manual(values = c("1" = navy, "2" = copper, "3" = green, "4" = purple, noise = "grey70")) +
  coord_equal() + labs(x = NULL, y = NULL, colour = NULL)
```

| k-means | mixture model | DBSCAN |
|---|---|---|
| ARI `r ari_of("k-means")` | ARI `r ari_of("Mixture model")` | ARI `r ari_of("DBSCAN")`, and `r db_noise_found` of 15 scattered points as noise |

---

## What each method saw

- **k-means** drew a straight boundary through the long group;
- the **mixture model** allowed the long group its elongated shape;
- **DBSCAN** joined the groups where they touch, one continuous dense region, but alone could say "no group".

No method is right in general. Choose the definition deliberately, and state it.

---

## Gaussian mixture models

The data is a mix of several groups, each with its own normal distribution.

The model estimates each group's **centre**, **spread**, and **size**.

---

## One variable: Old Faithful

```r
geyser <- Mclust(faithful$waiting)
summary(geyser, parameters = TRUE)
```

| | Short waits | Long waits |
|---|---|---|
| mixing probability | `r round(100 * gp$pro[1])`% | `r round(100 * gp$pro[2])`% |
| mean (minutes) | `r round(gp$mean[1])` | `r round(gp$mean[2])` |

Mclust tried one to nine groups and chose `r geyser$G`.

---

## Two normal curves over the data

```{r}
#| fig-width: 8
#| fig-height: 3.4
curves <- tibble(waiting = seq(40, 100, length.out = 300)) |>
  mutate(`Short waits` = gp$pro[1] * dnorm(waiting, gp$mean[1], sqrt(gp$variance$sigmasq)),
         `Long waits`  = gp$pro[2] * dnorm(waiting, gp$mean[2], sqrt(gp$variance$sigmasq))) |>
  tidyr::pivot_longer(-waiting, names_to = "group", values_to = "density")
ggplot(faithful, aes(x = waiting)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2, fill = "grey82", colour = "white") +
  geom_line(data = curves, aes(y = density, colour = group), linewidth = 1.3) +
  scale_colour_manual(values = c(`Short waits` = copper, `Long waits` = navy)) +
  labs(x = "Waiting time (minutes)", y = "Density", colour = NULL)
```

Waiting times between 272 eruptions: two humps, two groups.

---

## Membership probabilities

```r
predict(geyser, newdata = c(50, 67, 85))$z
```

| Wait | P(short) | P(long) |
|---|---|---|
| 50 minutes | `r geyser_z[1, 1]` | `r geyser_z[1, 2]` |
| 67 minutes | `r geyser_z[2, 1]` | `r geyser_z[2, 2]` |
| 85 minutes | `r geyser_z[3, 1]` | `r geyser_z[3, 2]` |

Not just which group, but how sure: **soft assignments**.

---

## Shapes and BIC

With several variables, each group has a centre and a **covariance matrix**: round or elongated, tilted any way.

mclust names each shape with three letters, such as `EEE` (all equal) or `VVV` (all variable).

The **Bayesian information criterion** rewards fit and penalises parameters. In mclust, **higher** BIC is better.

---

## Choosing the number of profiles

```r
set.seed(123)
profile_gmm <- Mclust(profile_data, G = 1:8)
summary(profile_gmm)
```

BIC chooses `r profile_gmm$G` clusters with the `r profile_gmm$modelName` structure.

The best four-cluster model is `r bic_gap` BIC points behind: above 6 is strong evidence, above 10 very strong.

---

## BIC by number of clusters

```{r}
#| fig-width: 9
#| fig-height: 3.8
bic <- as.data.frame(unclass(profile_gmm$BIC)); bic$G <- as.integer(rownames(bic))
bic |> tidyr::pivot_longer(-G, names_to = "model", values_to = "BIC") |>
  filter(!is.na(BIC)) |>
  ggplot(aes(G, BIC, group = model)) +
  geom_line(colour = "grey75") +
  geom_line(data = ~ filter(.x, model == profile_gmm$modelName), colour = copper, linewidth = 1.3) +
  scale_x_continuous(breaks = 1:8) + labs(x = "Number of clusters", y = "BIC (higher is better)")
```

Each grey line is one covariance structure; the chosen one is copper.

---

## Describing the three profiles

```{r}
#| results: asis
profiles |>
  summarise(students = n(), across(sleep_hours:satisfaction, ~ round(mean(.x), 1)), .by = gmm_cluster) |>
  arrange(gmm_cluster) |>
  rename(cluster = gmm_cluster, sleep = sleep_hours, study = study_hours,
         caffeine = caffeine_mg, exercise = exercise_days) |>
  knitr::kable()
```

Balanced, overloaded, and disengaged or isolated. Cluster numbers are arbitrary labels.

---

## Compared with k-means

```r
table(gmm = profile_gmm$classification, kmeans = kmeans_clusters$cluster)
```

- balanced and disengaged groups largely match;
- k-means's two "overloaded" clusters are joined into one.

With flexible shapes, the mixture model did not need to split the overloaded students.

---

## How certain is each assignment?

```r
head(round(profile_gmm$z, 2))
certainty <- apply(profile_gmm$z, 1, max)
sum(certainty < 0.8)
```

`apply(..., 1, max)` takes each row's largest probability.

`r sum(certainty < 0.8)` students have less than 80% probability for their most likely profile.

---

## Uncertain students lie where profiles meet

```{r}
#| fig-width: 8
#| fig-height: 3.9
pc <- prcomp(profile_data)$x[, 1:2]
tibble(PC1 = pc[, 1], PC2 = pc[, 2], cluster = factor(profile_gmm$classification),
       uncertainty = profile_gmm$uncertainty) |>
  ggplot(aes(PC1, PC2, colour = cluster, size = uncertainty)) +
  geom_point(alpha = 0.6) +
  scale_colour_manual(values = c(navy, copper, green)) + scale_size(range = c(0.6, 4)) +
  labs(colour = "Cluster", size = "Uncertainty")
```

Report the share who clearly fit, and treat the rest as mixtures.

---

## DBSCAN: clusters as dense regions

- a cluster is where cases are packed closely together;
- clusters are separated by sparser regions;
- cases in sparse regions are **noise**.

No number of clusters in advance, any shape, and "this case fits nowhere".

---

## Core points, border points, noise

| Setting or term | Meaning |
|---|---|
| `eps` | radius of each case's neighbourhood |
| `minPts` | cases a neighbourhood needs to be dense |
| core point | at least `minPts` cases within `eps` |
| border point | near a core point, but not itself core |
| noise | everything else |

Core points within `eps` of each other join the same cluster.

---

## A small example

```r
tiny <- tibble(
  x = c(1.0, 1.2, 1.1, 0.9, 1.3,   4.0, 4.2, 3.9, 4.1, 4.3,   2.5, 5.5),
  y = c(1.0, 1.1, 1.3, 1.2, 0.9,   3.0, 3.2, 3.1, 2.8, 3.0,   4.5, 0.5))
dbscan(tiny, eps = 0.5, minPts = 3)$cluster
#>  [1] 1 1 1 1 1 2 2 2 2 2 0 0
```

Two tight groups become clusters 1 and 2; the two isolated points are 0, noise.

k-means would have had to put them into a group.

---

## Choosing eps

```r
kNNdistplot(profile_data, k = 7)
abline(h = 2, lty = "dashed")
```

Each student's distance to their seventh-nearest neighbour, sorted.

The curve bends upward where isolated students begin: about 2. `minPts = 8`, roughly the number of variables plus one.

---

## One cloud, a few outsiders

```r
profile_db <- dbscan(profile_data, eps = 2, minPts = 8)
```

DBSCAN finds `r n_db_clusters` cluster and `r n_noise` noise points.

Not a failure: students form one continuous cloud. The profiles are regions of that cloud, not separate islands.

Smaller `eps` breaks the cloud into fragments and labels hundreds as noise.

---

## Students who fit no profile

| Kind | Pattern |
|---|---|
| extreme workers (`r n_extreme_work`) | about 60 study hours, 4 hours' sleep, over 700 mg caffeine, no exercise |
| extreme rest (`r n_extreme_rest`) | nearly 10 hours' sleep, very little study or caffeine, exercise almost daily |

DBSCAN finds students unusual in their **combination** of values, not one variable at a time.

Keep them, report them separately, and check the profiles without them.

---

## Evaluating a clustering

There are usually no true answers, so evaluation rests on several kinds of evidence:

| Evidence | Asks |
|---|---|
| internal measures | compact and well separated? |
| agreement between methods | do reasonable methods agree? |
| external validation | do clusters match known categories? |
| stability | do they survive small changes? |
| interpretability | do they make sense, and predict outcomes? |

---

## Internal: the silhouette

```r
profile_distances <- dist(profile_data)
mean(silhouette(kmeans_clusters$cluster, profile_distances)[, "sil_width"])
mean(silhouette(profile_gmm$classification, profile_distances)[, "sil_width"])
```

| k-means | mixture model |
|---|---|
| `r f2(sil_km)` | `r f2(sil_gmm)` |

Both low on a −1 to 1 scale: the profiles overlap, as DBSCAN suggested.

---

## The adjusted Rand index

```r
first  <- c(1, 1, 1, 2, 2, 2)
second <- c("B", "B", "B", "A", "A", "A")
third  <- c(1, 1, 2, 2, 3, 3)
adjustedRandIndex(first, second)   #> 1
adjustedRandIndex(first, third)    #> much lower
```

Counts pairs put together or apart by both clusterings, adjusted for chance. Labels do not matter.

k-means against the mixture model: ARI `r f2(ari_methods)`, substantial agreement for four clusters against three.

---

## Interpretability: do profiles predict outcomes?

```{r}
#| results: asis
profiles |>
  left_join(students |> select(student_id, considering_dropout), join_by(student_id)) |>
  summarise(students = n(), considering_dropout = paste0(round(100 * mean(considering_dropout == "Yes")), "%"),
            .by = gmm_cluster) |>
  arrange(gmm_cluster) |> rename(cluster = gmm_cluster) |> knitr::kable()
```

Dropout played no part in finding the profiles, yet they differ clearly on it.

---

## Report the choices, not just the clusters

- which variables, and how they were scaled;
- which method, and why that definition of a group;
- how many clusters, with the evidence (BIC, silhouette);
- how clearly cases fit, such as the share above 0.8 probability;
- how robust the profiles are to other reasonable choices.

---

## In your field: earth sciences

```{r}
#| fig-width: 5
#| fig-height: 3.9
quakes |> mutate(cluster = factor(quake_db$cluster)) |>
  ggplot(aes(long, lat, colour = cluster, shape = cluster == "0")) +
  geom_point(alpha = 0.7) +
  scale_colour_manual(values = c("0" = "grey65", "1" = navy, "2" = copper)) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4), guide = "none") +
  coord_quickmap() + labs(x = "Longitude", y = "Latitude", colour = "Cluster")
```

1,000 earthquakes near Fiji: DBSCAN follows the two curved trenches. Here dense regions really are separated by gaps.

---

## Practical lab: the Chapter 14 playground

Work through the playground exercises in your browser, with hints and solutions.

The browser version uses a hand-written DBSCAN and ARI; the download uses mclust and dbscan.

---

## Practical exercises 1–4: mixture models and DBSCAN

1. Four mixture profiles: which three-cluster profile splits?
2. How many students have a membership probability above 0.95?
3. DBSCAN with `eps` of 1.2, 1.6, and 2.4.
4. Refit the mixture model without the noise students.

---

## Practical exercises 5–7: evaluation

5. ARI between the mixture model and Ward's hierarchical clustering.
6. Compare the profiles on final GPA.
7. Separate the simulated groups with a gap: what changes for each method?

---

## Try this yourself

For grouping in your own field:

- which definition of a group fits your theory?
- would partial membership be meaningful?
- should some cases be allowed to fit no group?
- which outcome, not used to build the groups, could validate them?

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| an elongated group cut in two | k-means assumes round clusters |
| outliers forced into groups | k-means and mixtures place every case |
| BIC favours 3 and 4 almost equally | report both, with the evidence |
| lower BIC treated as better | in mclust, higher is better |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| DBSCAN finds one cluster | a continuous cloud with no gaps |
| hundreds of noise points | `eps` too small |
| cluster labels differ between runs | labels are arbitrary; compare with ARI |
| low silhouette for every method | groups overlap; no clustering will be crisp |

---

## Completion checklist

- [ ] the definition of a group is chosen and stated;
- [ ] the number of clusters is justified with BIC or silhouette;
- [ ] membership certainty is reported;
- [ ] students who fit no profile are identified and kept;
- [ ] at least two methods are compared with the ARI;
- [ ] stability is checked;
- [ ] profiles are validated on an outcome not used to build them.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| a sophisticated method finds the true groups | each finds groups matching its definition |
| every case belongs to some group | probabilities and noise describe reality better |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| the best BIC proves the number of groups | nearby models may be almost as good |
| DBSCAN failed with one cluster | one continuous cloud is a finding |

---

## The chapter in one sentence

> **Choose a definition of a group that fits your data, report how sure each assignment is and who fits nowhere, and judge clusters by agreement, stability, and meaning.**

---

## Next: Chapter 15

The next chapter introduces neural networks:

- a single neuron;
- from neurons to networks;
- a neural network for the dropout question;
- when neural networks are worth using;
- deep learning.

---

## Questions

In your field, can a case belong partly to two groups?

What would a case that fits no group tell you?
