---
title: "Neural Networks"
description: "Chapter 15: a single neuron as logistic regression, layers and activations, gradient descent by hand, weight decay, a tuned network for the dropout question, and when deep learning is worth it."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(tidymodels)
tidymodels_prefer()

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 14)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x); f3 <- function(x) sprintf("%.3f", x)

sigmoid <- function(z) 1 / (1 + exp(-z))
neuron <- function(stress, support) sigmoid(-4 + 1.2 * stress - 0.8 * support)
six <- data.frame(stress = c(2.0, 2.5, 3.0, 3.5, 4.0, 4.5) - 3.25, dropout = c(0, 0, 1, 0, 1, 1))
log_loss <- function(b0, b1) { p <- sigmoid(b0 + b1 * six$stress)
  -mean(six$dropout * log(p) + (1 - six$dropout) * log(1 - p)) }
p0 <- sigmoid(0 + 0 * six$stress)
g0 <- mean(p0 - six$dropout); g1 <- mean((p0 - six$dropout) * six$stress)
b0_1 <- -g0; b1_1 <- -g1
b0 <- 0; b1 <- 0; loss_history <- numeric(2000)
for (step in 1:2000) {
  p <- sigmoid(b0 + b1 * six$stress)
  b0 <- b0 - mean(p - six$dropout); b1 <- b1 - mean((p - six$dropout) * six$stress)
  loss_history[step] <- log_loss(b0, b1)
}
glm_coef <- coef(glm(dropout ~ stress, data = six, family = binomial))

scores <- questionnaire |>
  mutate(stress_4 = 6 - stress_4,
         stress = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
         burnout = rowMeans(pick(burnout_1:burnout_6), na.rm = TRUE),
         support = rowMeans(pick(support_1:support_6), na.rm = TRUE),
         satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)) |>
  select(student_id, stress, burnout, support, satisfaction)
dropout_data <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1) |> select(-semester), join_by(student_id)) |>
  mutate(considering_dropout = factor(considering_dropout, levels = c("Yes", "No"))) |>
  select(-student_id, -supervisor_id, -workshop, -workshop_sessions)
set.seed(2026)
dropout_split <- initial_split(dropout_data, prop = 0.75, strata = considering_dropout)
dropout_train <- training(dropout_split); dropout_test <- testing(dropout_split)
dropout_recipe <- recipe(considering_dropout ~ ., data = dropout_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
set.seed(2026)
dropout_folds <- vfold_cv(dropout_train, v = 10, strata = considering_dropout)

big_net <- mlp(hidden_units = 20, penalty = 0, epochs = 1000) |>
  set_engine("nnet", MaxNWts = 5000) |> set_mode("classification")
set.seed(2026)
big_fit <- fit(workflow(dropout_recipe, big_net), data = dropout_train)
n_weights <- length(extract_fit_engine(big_fit)$wts)
big_train_auc <- augment(big_fit, new_data = dropout_train) |> roc_auc(considering_dropout, .pred_Yes) |> pull(.estimate)
big_test_auc <- augment(big_fit, new_data = dropout_test) |> roc_auc(considering_dropout, .pred_Yes) |> pull(.estimate)

net_wf <- workflow(dropout_recipe, mlp(hidden_units = tune(), penalty = tune(), epochs = 500) |>
                     set_engine("nnet", MaxNWts = 5000) |> set_mode("classification"))
net_grid <- tidyr::expand_grid(hidden_units = c(1, 3, 5, 10, 20), penalty = 10^seq(-2, 1.5, by = 0.5))
set.seed(2026)
net_tuning <- tune_grid(net_wf, resamples = dropout_folds, grid = net_grid, metrics = metric_set(roc_auc))
best_net <- select_best(net_tuning, metric = "roc_auc")
best_net_auc <- show_best(net_tuning, metric = "roc_auc", n = 1)$mean
set.seed(2026)
logistic_cv <- fit_resamples(workflow(dropout_recipe, logistic_reg()), resamples = dropout_folds,
                             metrics = metric_set(roc_auc))
logistic_auc <- collect_metrics(logistic_cv)$mean
set.seed(2026)
final_net <- net_wf |> finalize_workflow(best_net) |> last_fit(dropout_split, metrics = metric_set(roc_auc, accuracy))
final_net_auc <- collect_metrics(final_net) |> filter(.metric == "roc_auc") |> pull(.estimate)
n_best_weights <- length(extract_fit_engine(extract_workflow(final_net))$wts)
final_logistic_auc <- last_fit(workflow(dropout_recipe, logistic_reg()), dropout_split) |>
  collect_metrics() |> filter(.metric == "roc_auc") |> pull(.estimate)

data(concrete, package = "modeldata")
set.seed(1)
concrete_split <- initial_split(concrete)
concrete_recipe <- recipe(compressive_strength ~ ., data = training(concrete_split)) |>
  step_log(age) |> step_normalize(all_numeric_predictors())
set.seed(1)
concrete_lm_fit <- last_fit(workflow(concrete_recipe, linear_reg()), concrete_split)
set.seed(1)
concrete_net_fit <- last_fit(workflow(concrete_recipe,
  mlp(hidden_units = 10, penalty = 0.1, epochs = 1000) |> set_engine("nnet", MaxNWts = 5000) |>
    set_mode("regression")), concrete_split)
concrete_rmse <- function(f) collect_metrics(f) |> filter(.metric == "rmse") |> pull(.estimate)
```

# Neural Networks

Less mysterious than the reputation suggests

**Chapter 15**

Polla Fattah

---

## By the end of today you can

- explain how a neuron combines its inputs;
- explain why a single sigmoid neuron is a logistic regression;
- describe layers, weights, activation functions, and training;
- fit and tune a network with `mlp()`, using weight decay;
- compare a network fairly with simpler models;
- say when deep learning is, and is not, the right tool.

---

## A claim worth testing

"A neural network would surely predict better than logistic regression."

Neural networks power image recognition, speech, translation, and large language models.

On a table of research data, the claim should be tested, not accepted.

---

## A single neuron

| Step | Does |
|---|---|
| weight | multiplies each input |
| bias | a constant added to the total |
| activation function | turns the total into the output |

Two inputs, stress and support, with weights 1.2 and −0.8, and a bias of −4.

---

## Working one neuron by hand

For a student with stress 4 and support 2:

$$
-4 + 1.2 \times 4 - 0.8 \times 2 = -0.8
$$

The **sigmoid** squeezes any number into 0 to 1:

$$
\text{sigmoid}(z) = \frac{1}{1 + e^{-z}}
$$

---

## The neuron in R

```r
sigmoid <- function(z) 1 / (1 + exp(-z))
neuron  <- function(stress, support) {
  sigmoid(-4 + 1.2 * stress - 0.8 * support)
}
neuron(stress = 4, support = 2)   #> 0.31
neuron(stress = 2, support = 4)   #> 0.008
```

---

## A single neuron is a logistic regression

| Neuron | Logistic regression (Chapter 8) |
|---|---|
| bias | intercept |
| weights | coefficients |
| sigmoid | turns log odds into a probability |

Everything a single neuron can do, Chapter 8 has already done.

---

## From neurons to networks

```{mermaid}
flowchart LR
  I1((Stress)) --> H1((H1))
  I1 --> H2((H2))
  I1 --> H3((H3))
  I2((Support)) --> H1
  I2 --> H2
  I2 --> H3
  I3((Sleep)) --> H1
  I3 --> H2
  I3 --> H3
  H1 --> O((P of<br>dropout))
  H2 --> O
  H3 --> O
```

A **multilayer perceptron**: inputs, a **hidden layer**, an output. Every arrow is a weight.

---

## Why layers add power

- each hidden neuron learns its own combination of inputs;
- the output neuron combines those;
- curved activations let the network represent curves and interactions.

With enough hidden neurons, a network can approximate almost any relationship. That flexibility is its strength, and its danger.

---

## Two activation functions

```{r}
#| fig-width: 8
#| fig-height: 3.2
activations <- data.frame(z = rep(seq(-5, 5, length.out = 200), 2),
                          name = rep(c("Sigmoid", "ReLU"), each = 200))
activations$output <- ifelse(activations$name == "Sigmoid", 1 / (1 + exp(-activations$z)), pmax(0, activations$z))
activations$name <- factor(activations$name, levels = c("Sigmoid", "ReLU"))
ggplot(activations, aes(z, output)) +
  geom_line(colour = navy, linewidth = 1.3) +
  facet_wrap(~ name, scales = "free_y") + labs(x = "Total input", y = "Output")
```

The sigmoid squeezes into 0 to 1. The **ReLU** zeroes negatives and is standard in deep networks.

---

## How a network learns

- start with small random weights;
- for every weight, find whether a small change would reduce the error: the **gradient**;
- move all weights a little in the helpful direction: **gradient descent**;
- one pass through the data is an **epoch**; training runs for hundreds.

---

## Six students, one neuron

```r
six <- data.frame(stress  = c(2.0, 2.5, 3.0, 3.5, 4.0, 4.5) - 3.25,
                  dropout = c(0,   0,   1,   0,   1,   1))
```

Stress is measured from the group's average.

The error to reduce is the **log loss**: small when high probabilities go to students who did consider dropping out.

---

## Starting from zero

```r
log_loss <- function(b0, b1) {
  p <- sigmoid(b0 + b1 * six$stress)
  -mean(six$dropout * log(p) + (1 - six$dropout) * log(1 - p))
}
log_loss(0, 0)   #> 0.693
```

With both weights at zero, every student gets a probability of 0.5.

---

## The gradient

```r
p <- sigmoid(b0 + b1 * six$stress)
gradient_b0 <- mean(p - six$dropout)
gradient_b1 <- mean((p - six$dropout) * six$stress)
```

The average prediction error, multiplied by the weight's input (1 for the bias).

The stress gradient is `r f3(g1)`: negative, so increasing the weight would reduce the error.

---

## One step of gradient descent

```r
learning_rate <- 1
b0 <- b0 - learning_rate * gradient_b0
b1 <- b1 - learning_rate * gradient_b1
```

| | Before | After one step |
|---|---|---|
| stress weight | 0 | `r f3(b1_1)` |
| log loss | `r f3(log_loss(0, 0))` | `r f3(log_loss(b0_1, b1_1))` |

Training is nothing more than this step, repeated.

---

## 2,000 steps

```{r}
#| fig-width: 8
#| fig-height: 2.5
ggplot(data.frame(step = 1:2000, loss = loss_history), aes(step, loss)) +
  geom_line(colour = navy, linewidth = 1.2) + scale_x_log10() +
  labs(x = "Training step (log scale)", y = "Log loss")
```

| | Bias | Stress weight |
|---|---|---|
| gradient descent | `r f3(b0)` | `r f3(b1)` |
| `glm()` | `r f3(abs(glm_coef[1]))` | `r f3(glm_coef[2])` |

---

## Two consequences

- training starts from random weights, so two runs differ: `set.seed()` matters;
- many weights can keep reducing training error by memorising noise.

**Weight decay** penalises large weights, exactly like the ridge penalty of Chapter 13. In tidymodels it is `penalty`.

Backpropagation calculates gradients efficiently for thousands of weights.

---

## A network for the dropout question

```r
mlp(hidden_units = , penalty = , epochs = ) |>
  set_engine("nnet", MaxNWts = 5000) |>
  set_mode("classification")
```

Same data, split, recipe, and folds as Chapters 11 and 12. Networks need normalised inputs; the recipe provides them.

The nnet engine, which comes with R, fits one hidden layer.

---

## A network that memorises

```r
big_net <- mlp(hidden_units = 20, penalty = 0, epochs = 1000) |>
  set_engine("nnet", MaxNWts = 5000) |> set_mode("classification")
big_fit <- fit(workflow(dropout_recipe, big_net), data = dropout_train)
```

| Data | AUC |
|---|---|
| training | `r f2(big_train_auc)` |
| test | `r f2(big_test_auc)` |

`r n_weights` weights for `r nrow(dropout_train)` students: more than enough freedom to memorise every one.

---

## Tuning size and weight decay

```r
net_spec <- mlp(hidden_units = tune(), penalty = tune(), epochs = 500) |>
  set_engine("nnet", MaxNWts = 5000) |> set_mode("classification")

net_grid <- expand_grid(hidden_units = c(1, 3, 5, 10, 20),
                        penalty = 10^seq(-2, 1.5, by = 0.5))
```

Weight decay on a log scale from 0.01 to about 30.

---

## The penalty matters more than the size

```{r}
#| fig-width: 9
#| fig-height: 3.6
collect_metrics(net_tuning) |>
  mutate(hidden_units = factor(hidden_units)) |>
  ggplot(aes(penalty, mean, colour = hidden_units)) +
  geom_line(linewidth = 1) + geom_point(size = 1.6) + scale_x_log10() +
  scale_colour_manual(values = c("#b9c7d6", "#7d97b3", navy, "#d9a66b", copper)) +
  labs(x = "Weight decay (log scale)", y = "Cross-validated AUC", colour = "Hidden units")
```

Little decay: networks overfit, the big ones most. Enough decay: every size does about equally well.

---

## A fair comparison

| Model | Cross-validated AUC | Test AUC |
|---|---|---|
| tuned network (`r best_net$hidden_units` hidden units) | `r f3(best_net_auc)` | `r f2(final_net_auc)` |
| logistic regression | `r f3(logistic_auc)` | `r f2(final_logistic_auc)` |

With `r nrow(dropout_test)` test students, `r sum(dropout_test$considering_dropout == "Yes")` at risk, this difference is well within chance.

A neural network does not predict dropout better.

---

## The costs of the network

- `r n_best_weights` weights that cannot be interpreted: no odds ratios;
- results depend on random starting weights without strong decay;
- careful tuning needed to avoid overfitting.

Logistic regression is kept. Reporting that a network was tried and did not help is itself a finding.

---

## When neural networks are worth using

| Situation | Why networks help |
|---|---|
| unstructured data: images, sound, text | features must be learned from raw inputs |
| very large datasets | thousands of weights need many cases |
| complex non-linear relationships | flexibility finds structure |

A table of a few hundred cases and a few dozen variables, the typical thesis, rarely qualifies.

---

## Deep learning

- many hidden layers, millions or billions of weights;
- each layer builds on the last: edges, then shapes, then objects;
- **convolutional networks** for images, **transformers** for text;
- large language models (Chapter 18) are transformers with billions of weights.

---

## Deep learning in R

| Tool | Notes |
|---|---|
| keras3 | interface to Keras; needs Python alongside R |
| torch | the PyTorch engine, without Python |
| brulee | tidymodels deep networks through torch, with `mlp()` |

For most researchers, the practical route is a **pretrained model**, as Chapter 18 does with a language model.

---

## In your field: engineering

```r
concrete_recipe <- recipe(compressive_strength ~ ., data = training(concrete_split)) |>
  step_log(age) |>
  step_normalize(all_numeric_predictors())
concrete_net <- mlp(hidden_units = 10, penalty = 0.1, epochs = 1000) |>
  set_engine("nnet", MaxNWts = 5000) |> set_mode("regression")
```

| Model | Test RMSE (MPa) |
|---|---|
| linear regression | `r sprintf("%.1f", concrete_rmse(concrete_lm_fit))` |
| neural network | `r sprintf("%.1f", concrete_rmse(concrete_net_fit))` |

Curves and interactions give the flexibility something to capture.

---

## Practical lab: the Chapter 15 playground

Work through the playground exercises in your browser, with hints and solutions.

nnet runs in the browser; the download uses tidymodels.

---

## Practical exercises 1–3: neurons and networks

1. Double the importance of support in the neuron: what does a negative weight mean?
2. One hidden neuron with decay 1 against logistic regression.
3. Three seeds with decay 0.01 and with the best decay: when does the seed matter?

---

## Practical exercises 4–6: training

4. The 20-neuron network for 10, 100, and 1000 epochs.
5. Explain to the research group why the thesis keeps logistic regression.
6. Gradient descent with learning rates 0.1 and 10: what goes wrong?

---

## Try this yourself

Someone suggests "just use AI" for your prediction problem.

- how many cases and variables do you have?
- is the data a table, or images, sound, or text?
- what simple model would you compare against?
- what would you lose in interpretability?

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| perfect training AUC | memorisation; add weight decay |
| different results on each run | random starting weights; set a seed, add decay |
| "too many weights" error | raise `MaxNWts` |
| a network no better than regression | a smooth pattern; that is a finding |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| loss explodes during training | learning rate too large |
| loss barely moves | learning rate too small, or too few epochs |
| inputs on very different scales | missing `step_normalize()` |
| no one can explain the model | many uninterpretable weights |

---

## Completion checklist

- [ ] a simple model is fitted first as the benchmark;
- [ ] inputs are normalised;
- [ ] hidden units and weight decay are tuned by cross-validation;
- [ ] the seed is set;
- [ ] training and test performance are both reported;
- [ ] the comparison with the benchmark is fair;
- [ ] the loss in interpretability is weighed.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| a neural network is always more accurate | on typical tables, often not |
| networks learn mysteriously | gradient descent: small steps that reduce error |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| a perfect training fit means learning | usually memorisation |
| neural networks and AI are the same | networks are one family; LLMs are one kind |

---

## The chapter in one sentence

> **A neural network is logistic regression units stacked in layers and trained by small steps; it earns its complexity on large, unstructured, or strongly non-linear data, not on every table.**

---

## Next: Chapter 16

The next chapter turns to data ordered in time:

- time series data and its components;
- the counselling service's weekly visits;
- decomposition into trend and seasonality;
- forecasting, and judging forecasts;
- what forecasts assume about the future.

---

## Questions

What would convince you that a neural network is worth its cost?

What would you report if it were not?
