---
title: "Time Series Forecasting"
description: "Chapter 16: dependence over time, trend, seasonality and noise, tsibbles, autocorrelation, STL decomposition, benchmarks, ETS and Fourier ARIMA, test periods, and prediction intervals."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(tsibble)
library(fable)
library(feasts)

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft),
        strip.text = element_text(colour = navy, face = "bold", size = 13)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f1 <- function(x) sprintf("%.1f", x); f2 <- function(x) sprintf("%.2f", x)

clinic <- tsibble(day = 1:14, visits = c(12, 10, 9, 11, 7, 3, 2, 14, 11, 10, 12, 8, 4, 2), index = day)
clinic_fc <- clinic |>
  model(mean = MEAN(visits), naive = NAIVE(visits), snaive = SNAIVE(visits ~ lag(7))) |>
  forecast(h = 7)

visits <- counselling_visits |> mutate(week = yearweek(week_start)) |> as_tsibble(index = week)
visits_acf <- visits |> ACF(visits, lag_max = 60)
visits_stl <- visits |> model(STL(visits ~ season(period = 52), robust = TRUE)) |> components()
odd_week <- visits_stl |> as_tibble() |> slice_max(abs(remainder), n = 1)

visits_train <- visits |> filter(week_start < as.Date("2024-09-01"))
visits_test  <- visits |> filter(week_start >= as.Date("2024-09-01"))
visits_fit <- visits_train |>
  model(mean = MEAN(visits), naive = NAIVE(visits), snaive = SNAIVE(visits ~ lag(52)),
        stl_ets = decomposition_model(STL(log(visits) ~ season(period = 52), robust = TRUE),
                                      ETS(season_adjust ~ season("N"))),
        fourier = ARIMA(log(visits) ~ fourier(period = 52, K = 6) + PDQ(0, 0, 0)))
visits_fc <- visits_fit |> forecast(h = 52)
acc <- visits_fc |> accuracy(visits) |> arrange(RMSE)

final_fit <- visits |>
  model(stl_ets = decomposition_model(STL(log(visits) ~ season(period = 52), robust = TRUE),
                                      ETS(season_adjust ~ season("N"))))
next_year <- final_fit |> forecast(h = 52)
first_hilo <- hilo(next_year$visits[1], 95)
set.seed(2026)
futures <- final_fit |> generate(h = 52, times = 1000)
yearly_totals <- futures |> as_tibble() |> summarise(total = sum(.sim), .by = .rep)
q <- round(quantile(yearly_totals$total, c(0.025, 0.5, 0.975)))
last_year_total <- sum(visits_test$visits)

co2_ts <- as_tsibble(co2)
co2_accuracy <- co2_ts |> filter(index <= yearmonth("1990 Dec")) |>
  model(snaive = SNAIVE(value), ets = ETS(value), arima = ARIMA(value)) |>
  forecast(h = "7 years") |> accuracy(co2_ts) |> select(.model, RMSE, MAE)
fmt_n <- function(x) format(x, big.mark = ",")
```

# Time Series Forecasting

When the order of observations carries information

**Chapter 16**

Polla Fattah

---

## By the end of today you can

- explain dependence over time, and why forecasts are ranges;
- name the parts of a series: trend, seasonality, noise;
- store a series as a tsibble, and plot it;
- read time plots, seasonal plots, and autocorrelation;
- decompose a series with STL;
- forecast with benchmarks, ETS, and ARIMA in fable;
- evaluate forecasts on a test period, and report prediction intervals.

---

## A different way of thinking

| Survey data | Time series |
|---|---|
| observations independent | this week relates to last week |
| order does not matter | order carries information |
| one value per case | trend, season, and noise together |

A forecast of the future is honestly a range, never a single number.

---

## The counselling service's question

Five years of weekly visit counts: overwhelmed before exams, quiet in summer, staffing planned by guesswork.

How many visits should the service expect next year, week by week?

An **internal analysis**, done for the service rather than for publication.

---

## A tsibble knows which column is time

```r
clinic <- tsibble(
  day    = 1:14,
  visits = c(12, 10, 9, 11, 7, 3, 2,   14, 11, 10, 12, 8, 4, 2),
  index  = day
)
```

Two weeks of daily visits to a small clinic, from a Monday.

The **index** keeps observations in order and each time unique. Busy early in the week, quiet at the weekend.

---

## Three parts of a pattern

```r
parts <- tibble(
  week     = 1:156,
  trend    = 30 + 0.05 * week,
  seasonal = 12 * sin(2 * pi * week / 52),
  noise    = rnorm(156, sd = 3)
) |>
  mutate(series = trend + seasonal + noise)
```

An artificial weekly series for three years, built from its parts.

---

## Only the sum is observed

```{r}
#| fig-width: 9
#| fig-height: 4.3
set.seed(4)
parts <- tibble(week = 1:156, trend = 30 + 0.05 * (1:156),
                seasonal = 12 * sin(2 * pi * (1:156) / 52), noise = rnorm(156, sd = 3)) |>
  mutate(series = trend + seasonal + noise)
parts |> tidyr::pivot_longer(-week, names_to = "part", values_to = "value") |>
  mutate(part = factor(part, levels = c("trend", "seasonal", "noise", "series"))) |>
  ggplot(aes(week, value)) + geom_line(colour = navy) +
  facet_wrap(~ part, ncol = 2, scales = "free_y") + labs(x = "Week", y = NULL)
```

Forecasting extends trend and season. Noise cannot be forecast: the main reason every forecast has a range.

---

## The counselling data as a tsibble

```r
visits <- counselling_visits |>
  mutate(week = yearweek(week_start)) |>
  as_tsibble(index = week)
```

Five academic years, September 2020 to August 2025: `r nrow(counselling_visits)` weeks.

`yearweek()` turns each Monday's date into a week; the header reads `[1W]`.

---

## Plot the series first

```{r}
#| fig-width: 9
#| fig-height: 3.4
autoplot(visits, visits, colour = navy) + labs(x = NULL, y = "Visits per week")
```

A seasonal pattern each academic year, a gentle upward trend, noise, and one unusual week in autumn 2023.

---

## A seasonal plot

```{r}
#| fig-width: 9
#| fig-height: 3.5
visits |>
  mutate(academic_year = paste0(20 + (row_number() - 1) %/% 52, "/", 21 + (row_number() - 1) %/% 52),
         week_of_year = (row_number() - 1) %% 52 + 1) |>
  ggplot(aes(week_of_year, visits, colour = academic_year)) +
  geom_line(linewidth = 0.8) + scale_colour_viridis_d(end = 0.9) +
  labs(x = "Week of the academic year", y = "Visits", colour = "Year")
```

Every year has the same shape; later years lie slightly higher. The period is 52 weeks.

---

## Busy weeks follow busy weeks

```{r}
#| fig-width: 5.5
#| fig-height: 3.9
visits |> mutate(last_week = lag(visits)) |>
  ggplot(aes(last_week, visits)) + geom_point(alpha = 0.6, colour = navy) +
  labs(x = "Visits in the previous week", y = "Visits this week")
```

Knowing last week's visits says a good deal about this week's.

---

## Autocorrelation

The correlation between the series and itself $k$ steps earlier.

```r
visits_acf <- visits |> ACF(visits, lag_max = 60)
```

| Lag | 1 | 26 | 52 |
|---|---|---|---|
| autocorrelation | `r f2(visits_acf$acf[1])` | `r f2(visits_acf$acf[26])` | `r f2(visits_acf$acf[52])` |

A week resembles the same week a year earlier. Half a year apart, a busy term week often meets a quiet summer week.

---

## The autocorrelation function

```{r}
#| fig-width: 9
#| fig-height: 3.4
autoplot(visits_acf) + labs(x = "Lag (weeks)", y = "Autocorrelation")
```

A peak at the seasonal lag is the signature of seasonality. No autocorrelation would mean nothing to forecast beyond the average.

---

## Decomposition with STL

```r
visits_stl <- visits |>
  model(STL(visits ~ season(period = 52), robust = TRUE)) |>
  components()
```

| Component | Is |
|---|---|
| trend | the slow movement |
| seasonal | the repeating academic-year pattern |
| remainder | what is left: the noise |

`robust = TRUE` stops unusual weeks from distorting trend and season.

---

## The recovered parts

```{r}
#| fig-width: 9
#| fig-height: 4.3
visits_stl |> as_tibble() |>
  select(week, visits, trend, season_52, remainder) |>
  tidyr::pivot_longer(-week, names_to = "part", values_to = "value") |>
  mutate(part = factor(part, levels = c("visits", "trend", "season_52", "remainder"))) |>
  ggplot(aes(as.Date(week), value)) + geom_line(colour = navy) +
  facet_wrap(~ part, ncol = 2, scales = "free_y") + labs(x = NULL, y = NULL)
```

The trend rises from about `r round(min(visits_stl$trend))` to `r round(max(visits_stl$trend))` visits a week, levelling off in the last year.

---

## The unusual week

```r
visits_stl |> as_tibble() |>
  slice_max(abs(remainder), n = 3) |>
  select(week, visits, trend, season_52, remainder)
```

The week of `r format(as.Date(odd_week$week), "%d %B %Y")`: `r odd_week$visits` visits, about `r round(odd_week$remainder)` more than trend and season predict.

It was the mental health awareness week. A real week, not an error: kept, and the robust fit stops it distorting forecasts.

---

## Growing seasonal swings

The peaks get higher as the trend rises, while the summer lows stay low.

When the season grows with the level, model the **logarithm** of the series.

In fable, write `log(visits)` in the model; forecasts are transformed back automatically.

---

## Benchmarks first

| Method | Forecasts |
|---|---|
| mean | the average of all past observations |
| naive | the last observed value |
| seasonal naive | the value from the same season last cycle |

Every forecasting method must beat these, as every model in Chapters 11–13 had to beat a baseline.

---

## Benchmarks on the clinic

```r
clinic |>
  model(mean   = MEAN(visits),
        naive  = NAIVE(visits),
        snaive = SNAIVE(visits ~ lag(7))) |>
  forecast(h = 7)
```

| Method | Forecast for next week |
|---|---|
| mean | `r f1(mean(clinic$visits))` every day |
| naive | 2 every day: a quiet Sunday repeated |
| seasonal naive | last week's pattern repeated |

---

## Two families of models

| Family | Forecasts with |
|---|---|
| ETS, exponential smoothing | weighted averages, recent weeks counting most; tracks level, trend, season |
| ARIMA | how each value depends on earlier values and shocks |

`ETS()` and `ARIMA()` choose their details automatically.

Both work best with short seasons: 4 quarters, 7 days, 12 months. A 52-week season is long.

---

## Two strategies for weekly data

```{mermaid}
flowchart LR
    A["Weekly series"] --> B["STL: remove<br>the season"] --> C["ETS on the<br>adjusted series"] --> D["Add the<br>season back"]
    A --> E["ARIMA with<br>Fourier waves"]
```

`decomposition_model()` does the first in one step. `fourier(period = 52, K = 6)` uses six pairs of smooth waves.

---

## Training and test periods

```r
visits_train <- visits |> filter(week_start <  as.Date("2024-09-01"))
visits_test  <- visits |> filter(week_start >= as.Date("2024-09-01"))
```

The test period must come **after** the training period: a forecast can only use the past.

Train on four academic years, test on the fifth.

---

## Five models at once

```r
visits_fit <- visits_train |>
  model(
    mean    = MEAN(visits),
    naive   = NAIVE(visits),
    snaive  = SNAIVE(visits ~ lag(52)),
    stl_ets = decomposition_model(
                STL(log(visits) ~ season(period = 52), robust = TRUE),
                ETS(season_adjust ~ season("N"))),
    fourier = ARIMA(log(visits) ~ fourier(period = 52, K = 6) + PDQ(0, 0, 0)))
```

`season("N")`: no season inside ETS, it comes from STL. `PDQ(0, 0, 0)`: the Fourier terms handle the season.

---

## Accuracy on the test year

```{r}
#| results: asis
acc |> select(.model, RMSE, MAE) |>
  mutate(across(c(RMSE, MAE), f1)) |> rename(model = .model) |> knitr::kable()
```

STL with ETS is best. The Fourier model struggles with sharp steps, such as the summer drop from one week to the next.

---

## The two best, against reality

```{r}
#| fig-width: 9
#| fig-height: 3.5
visits_fc |> filter(.model %in% c("stl_ets", "snaive")) |>
  autoplot(visits_test, level = NULL) +
  scale_colour_manual(values = c(snaive = "grey55", stl_ets = copper)) +
  labs(x = NULL, y = "Visits per week", colour = "Model")
```

The seasonal naive copies last year's noise, even the awareness-week spike. STL with ETS is smoother and closer.

---

## Forecasting next year

```r
final_fit <- visits |>
  model(stl_ets = decomposition_model(
    STL(log(visits) ~ season(period = 52), robust = TRUE),
    ETS(season_adjust ~ season("N"))))
next_year <- final_fit |> forecast(h = 52)
```

Refitted on all five years, so the forecasts use the most recent information.

---

## Prediction intervals

```{r}
#| fig-width: 9
#| fig-height: 3.4
next_year |> autoplot(visits |> filter(week_start >= as.Date("2023-09-01")), colour = copper) +
  labs(x = NULL, y = "Visits per week")
```

80% and 95% ranges. First week: about `r round(next_year$.mean[1])` visits, anywhere from `r round(first_hilo$lower)` to `r round(first_hilo$upper)` unsurprising.

---

## A total for the year needs simulation

```r
set.seed(2026)
futures <- final_fit |> generate(h = 52, times = 1000)
yearly_totals <- futures |> as_tibble() |>
  summarise(total = sum(.sim), .by = .rep)
quantile(yearly_totals$total, c(0.025, 0.5, 0.975))
```

Weeks do not vary independently, so weekly intervals cannot simply be added.

`generate()` produces 1,000 possible futures; their totals give the range.

---

## The report to the service

- about **`r fmt_n(q[2])`** visits expected in 2025/26 (95% interval `r fmt_n(q[1])` to `r fmt_n(q[3])`), against `r fmt_n(last_year_total)` in 2024/25;
- the busiest weeks will again be just before and during the exam periods;
- an awareness campaign can raise demand sharply for a week: plan extra capacity.

---

## Forecasts assume the future resembles the past

A new booking system, a changed exam calendar, or a crisis like COVID-19 can make any forecast wrong.

- report forecasts with their intervals;
- say what they assume;
- update them as new data arrives.

Prophet, once popular, is no longer actively developed; fable is well supported.

---

## In your field: environmental science

```r
co2_ts |>
  filter(index <= yearmonth("1990 Dec")) |>
  model(snaive = SNAIVE(value), ets = ETS(value), arima = ARIMA(value)) |>
  forecast(h = "7 years") |> accuracy(co2_ts)
```

| Model | MAE (ppm) |
|---|---|
| seasonal naive | `r f1(co2_accuracy$MAE[co2_accuracy$.model == "snaive"])` |
| ETS | `r f1(co2_accuracy$MAE[co2_accuracy$.model == "ets"])` |
| ARIMA | `r f1(co2_accuracy$MAE[co2_accuracy$.model == "arima"])` |

Mauna Loa CO₂: a strong trend and a 12-month cycle that ETS and ARIMA model directly.

---

## Practical lab: the Chapter 16 playground

Work through the playground exercises in your browser, with hints and solutions.

The browser version builds Fourier terms with `lm()`; the download uses fable.

---

## Practical exercises 1–4: parts and models

1. Which benchmark suits a trend without seasonality? Test it on the clinic.
2. Plot the seasonally adjusted series: what does it reveal?
3. Fourier terms with `K = 2` and `K = 12`.
4. Train on three years, test on the fourth.

---

## Practical exercises 5–7: forecasts and communication

5. Expected visits in the four weeks before the first exams, with an interval.
6. Explain a 95% prediction interval to the service.
7. Double the noise in the artificial series: what happens to forecast intervals?

---

## Try this yourself

Find a series in your own field.

- plot it, and name its trend, season, and noise;
- choose the seasonal period;
- pick a benchmark it must beat;
- decide what event could make a forecast wrong.

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| "duplicated index" error | the same time appears twice |
| ETS or ARIMA struggles with weekly data | the 52-week season is long; decompose first |
| forecasts copy last year's spikes | seasonal naive repeats the noise |
| one week dominates the trend | use `robust = TRUE` |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| test error suspiciously low | test period not after training period |
| seasonal swings grow with the level | model `log()` of the series |
| a yearly total without an interval | weekly intervals cannot be added; simulate |
| forecasts wrong after a change | the past no longer resembles the future |

---

## Completion checklist

- [ ] the series is a tsibble with the right index;
- [ ] time, seasonal, and ACF plots were inspected;
- [ ] the series was decomposed, and unusual weeks explained;
- [ ] benchmarks are reported;
- [ ] models are tested on a later period;
- [ ] forecasts include prediction intervals;
- [ ] totals come from simulation;
- [ ] assumptions are stated.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| a forecast predicts the exact value | a forecast is a range |
| more history always helps | only if old patterns still hold |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| any random subset can be the test set | the test period comes after training |
| an unusual week should be removed | keep it; use robust methods |

---

## The chapter in one sentence

> **Separate trend, season, and noise, beat the simple benchmarks on a later test period, and always report a forecast as a range with its assumptions.**

---

## Next: Chapter 17

The next chapter makes research reproducible:

- reproducibility and credibility;
- documents that contain their own analysis;
- a results chapter in Quarto;
- fixed package versions with renv;
- interactive dashboards with Shiny, and open science.

---

## Questions

What regular measurement in your field has a seasonal pattern?

What event could break its forecast?
