---
title: "Reproducible Research & Quarto"
subtitle: "Literate Programming, Parameterized Reporting, renv, Shiny Dashboards, and Open Science"
author: "Dr. Polla Fattah"
institute: "From Data to Thesis: Research Data Analysis with R"
format:
  revealjs:
    theme: [default, ../../styles/theme.scss]
    slide-number: true
    preview-links: auto
    chalkboard: true
    transition: slide
    highlight-style: github
    code-line-numbers: false
    width: 1200
    height: 750
execute:
  freeze: auto
---

# Reproducible Research

### Science, Integrity, and Computational Transparency

**Chapter 17 · From Data to Thesis**

Dr. Polla Fattah

---

## The Supervisor's Midnight Nightmare

Imagine this classic graduate scenario:

> **Three weeks before thesis submission:**  
> Your supervisor discovers that 12 survey respondents were recorded twice due to a network glitch.  
> You fix the script to remove them.  
> **Now what?**

- Every summary mean in Chapter 4 is off by 0.04.
- Every standard error, $t$-statistic, and $p$-value in Chapter 5 has shifted.
- Figures were exported as PNGs and pasted into Word six months ago.
- Dozens of tables were manually typed cell-by-cell.

**Result**: Weeks of panic, sleepless nights, and inevitable copy-paste errors.

---

## The Core Thesis Principle

::: {.callout-important}
### Scientific Findings Must Be Auditable
A research finding is credible **only if it can be checked**. If another scholar takes your raw data and scripts, they must arrive at the exact same numbers.
:::

- Science is built on **verifiability**, not trust.
- If an analysis cannot be recomputed, there is no foundation to evaluate whether it replicates in new populations.
- **Reproducibility is a matter of research integrity**, not merely technical neatness.

---

## Reproducibility vs Replicability

The National Academies and scientific literature draw a clean distinction:

| Concept | Data Used | Code / Methods | Question Answered |
|---|---|---|---|
| **Reproducibility** | **Same** raw data | **Same** code / pipeline | Can independent researchers recompute the identical numbers? |
| **Replicability** | **New** sample / data | **Identical** study protocol | Does the observed effect hold up in an independent investigation? |

> Reproducibility is the **bare minimum threshold** for scientific validity.

---

## The Psychology Replication Shock (2015)

In 2015, the Open Science Collaboration replicated 100 benchmark empirical psychology studies published in top-tier journals:

```
Original Studies:     [■■■■■■■■■■■■■■■■■■■■■■■■■■■■■■] 97% Significant (p < .05)
Replication Studies:  [■■■■■■■■■■                    ] 36% Significant
Effect Sizes:         Averaged ~50% of the original claimed magnitudes!
```

### Why did this happen?
1. Small, underpowered samples.
2. The **file-drawer effect** (journals rejecting null results).
3. Post-hoc decisions (**p-hacking** / researcher degrees of freedom).
4. Irreproducible data preparation pipelines.

---

## Where Irreproducibility Comes From

True scientific fraud is rare. Most failures stem from everyday friction:

- **Point-and-click software**: Clicking buttons in SPSS or Excel without an audit trail.
- **Manual transcriptions**: Copying numbers from console output into Word tables.
- **Lost versions**: `analysis_v1.R`, `final_model_v2.R`, `final_really_final.R`.
- **Silent dependency shifts**: A package updates and changes a default algorithm parameter.
- **Hard-coded local paths**: `read.csv("C:/Users/John/Desktop/data.csv")`.

**The Solution**: An automated, code-driven literate programming pipeline.

---

# Literate Programming with Quarto

### Unifying Code, Data, Text, and Graphics

---

## What is Quarto?

**Quarto** is an open-source scientific and technical publishing system developed by Posit (makers of RStudio).

- Written in plain-text markdown files (`.qmd`).
- Combines explanatory text, mathematical derivations, executable code chunks, and live output.
- Language-agnostic: Works natively with **R**, **Python**, **Julia**, and **Observable JS**.
- Successor to **R Markdown** (`.Rmd`), offering unified syntax and multi-language support.

> Every number, table, and graph in your thesis document is computed on the fly directly from raw data!

---

## The Quarto Render Engine Architecture

How does a `.qmd` file become a polished PDF, Word doc, or HTML slide deck?

```
+---------------+      knitr       +---------------+      pandoc      +-------------------+
|  report.qmd   | --------------> |   report.md   | -------------->  | PDF / Word / HTML |
| (Code + Text) | (Executes R &   | (Text + Live  | (Compiles into   | Presentations /   |
|               |  catches plots) |  Results)     |  final document) | Entire Theses     |
+---------------+                 +---------------+                  +-------------------+
```

1. **knitr** parses the file, isolates R code chunks, executes them sequentially, and converts output into standard Markdown.
2. **Pandoc** converts the resulting Markdown into LaTeX/PDF, Microsoft Word (`.docx`), HTML, or RevealJS slides.

---

## The Three Components of a `.qmd` File

Every Quarto document consists of three well-defined layers:

```markdown
---
title: "First Semester Student Wellbeing"
author: "Elaf PhD"
format: html
---

# 1. YAML Frontmatter (Configuration & Metadata)
Controls styling, outputs, bibliographies, and document structure.

# 2. Markdown Prose (Narrative & Explanations)
Uses lightweight syntax: **bold**, *italics*, # Headers, bullet points.

# 3. Executable Code Chunks
R code enclosed in ```{r} ... ``` blocks that execute during render.
```

---

## Inline Code: Never Type a Number by Hand

Consider this sentence in your thesis draft:

```markdown
The final sample included `r knitr::inline_expr("nrow(students)")` graduate students. 
Their average baseline wellbeing was `r knitr::inline_expr("round(mean(semesters$wellbeing, na.rm = TRUE), 1)")`.
```

When rendered by Quarto, it evaluates dynamically into:

> *"The final sample included 800 graduate students. Their average baseline wellbeing was 64.2."*

If 12 students are excluded or survey data updates:
- **You re-render the document (`quarto render`).**
- **Every inline number throughout 200 pages updates instantly!**

---

## Anatomy of a Code Chunk

```markdown
```{{r}}
#| label: tbl-faculty-summary
#| tbl-cap: "Descriptive wellbeing statistics across academic faculties."
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(knitr)

semesters |>
  filter(semester == 1) |>
  left_join(students, join_by(student_id)) |>
  summarise(
    N    = n(),
    Mean = round(mean(wellbeing, na.rm = TRUE), 2),
    SD   = round(sd(wellbeing, na.rm = TRUE), 2),
    .by  = faculty
  ) |>
  kable(col.names = c("Faculty", "Sample (N)", "Mean Score", "Std Dev"))
```
```

---

## Key Chunk Execution Options (`#|`)

| Option | Values | Purpose |
|---|---|---|
| `#| echo:` | `true` / `false` | Should the R code itself be printed in the output document? (Typically `false` for thesis chapters). |
| `#| eval:` | `true` / `false` | Should the code block be executed? |
| `#| message:` | `true` / `false` | Suppresses package startup and tidyverse messages. |
| `#| warning:` | `true` / `false` | Suppresses statistical warning messages (e.g. ties in rank tests). |
| `#| tbl-cap:` | `"String"` | Generates an official numbered table caption. |
| `#| fig-cap:` | `"String"` | Generates an official numbered figure caption. |
| `#| fig-width:` | `6.5` | Sets plot width in inches. |

---

## Cross-Referencing: Automatic Numbering

Quarto provides automatic, dynamic cross-referencing:

```markdown
In our survey, faculty wellbeing varied significantly (@tbl-faculty-summary), 
mirroring trends visualized in @fig-wellbeing-trajectory.
```

- When the chunk label begins with `tbl-`, reference it as `@tbl-...`.
- When the chunk label begins with `fig-`, reference it as `@fig-...`.
- When headings have IDs (`## Methods {#sec-methods}`), reference as `@sec-methods`.

### Benefit
If you insert a new table before Table 2, Pandoc automatically renumbers **every subsequent table and reference** across your thesis.

---

## Bibliographies and Citation Management

Academic theses require rigorous citation tracking. Quarto integrates with BibTeX:

In `thesis.qmd` YAML header:
```yaml
---
title: "Doctoral Thesis"
format: pdf
bibliography: references.bib
csl: apa-7th-edition.csl
---
```

In your BibTeX file (`references.bib`):
```bibtex
@book{kuhn2022,
  author    = {Kuhn, Max and Silge, Julia},
  title     = {Tidy Modeling with R},
  publisher = {O'Reilly Media},
  year      = {2022}
}
```

In your text:
```markdown
Machine learning workflows were executed using tidymodels [@kuhn2022].
```

---

## Mathematical Typesetting with LaTeX

Quarto parses standard LaTeX mathematical expressions across all output formats (HTML, Word, PDF):

### Inline Math:
```markdown
The sample mean is defined as $\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i$.
```
The sample mean is defined as $\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i$.

### Display Math:
```markdown
$$
\text{logit}(p_i) = \beta_0 + \beta_1 \cdot \text{stress}_i + \beta_2 \cdot \text{support}_i + u_{j[i]}
$$
```

$$
\text{logit}(p_i) = \beta_0 + \beta_1 \cdot \text{stress}_i + \beta_2 \cdot \text{support}_i + u_{j[i]}
$$

---

## Multi-Format Rendering

You write **one single `.qmd` file**, but render to whatever medium is required:

```yaml
---
title: "Wellbeing Analysis"
format:
  html:
    toc: true
    theme: cosmo
  docx:
    reference-doc: university-template.docx
  pdf:
    documentclass: scrreprt
    keep-tex: false
---
```

- **HTML**: Interactive exploration, supplementary websites, online reports.
- **Word (`docx`)**: Feedback loops with supervisors and co-authors using tracked changes.
- **PDF**: Camera-ready doctoral thesis submission (compiled via TinyTeX).

---

## Parameterized Reporting

What if the Dean requests separate customized reports for all 5 faculties?

**Do not create 5 separate `.qmd` files!** Use document parameters:

```yaml
---
title: "Faculty Wellbeing Evaluation"
format: html
params:
  target_faculty: "Education"
---
```

In your code chunk:
```r
faculty_data <- semesters |>
  left_join(students, join_by(student_id)) |>
  filter(faculty == params$target_faculty)
```

In PowerShell or Bash:
```bash
quarto render report.qmd -P target_faculty:"Engineering" -o "Engineering-Report.html"
quarto render report.qmd -P target_faculty:"Science" -o "Science-Report.html"
```

---

# Package Management with `renv`

### Freezing Computational Environments Across Time and Space

---

## The Dependency Rot Dilemma

> *"It worked on my laptop two years ago, but now the script crashes with an error!"*

Package ecosystems evolve constantly:
- Functions get deprecated or renamed.
- Default arguments shift (e.g. `stringsAsFactors = FALSE`).
- Underlying C/C++ dependencies get updated.

If you publish a paper today, will another researcher's R installation run your code in 2030?

**Solution**: Isolated project environments with **`renv`**.

---

## The `renv` Workflow

`renv` gives your project its own private package library and records the exact version of every package in a `renv.lock` file.

```
       1. Initialize                2. Develop & Lock                 3. Replicate
       
   +--------------------+         +--------------------+         +--------------------+
   |   renv::init()     | ------> |  renv::snapshot()  | ------> |   renv::restore()  |
   | (Creates private   |         | (Records versions  |         | (Installs exact    |
   |  project library)  |         |  to renv.lock)     |         |  lockfile versions)|
   +--------------------+         +--------------------+         +--------------------+
```

- **`renv::init()`**: Disconnects project from the global user library; builds isolated library.
- **`renv::snapshot()`**: Scans project scripts and pins exact package versions to `renv.lock`.
- **`renv::restore()`**: Reinstalls the exact frozen package ecosystem on any new machine.

---

## Inspecting the `renv.lock` File

The lockfile is a simple, human-readable JSON manifest:

```json
{
  "R": {
    "Version": "4.4.1",
    "Repositories": [
      { "Name": "CRAN", "URL": "https://cloud.r-project.org" }
    ]
  },
  "Packages": {
    "lme4": {
      "Package": "lme4",
      "Version": "1.1-35.5",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "c20df5d68d1f2b604e304b"
    }
  }
}
```

Committing `renv.lock` to your repository ensures anyone can rebuild your exact analysis environment!

---

## Documenting Software in Your Methodology

Even without `renv`, rigorous thesis reporting requires logging your runtime environment:

```{r}
#| echo: true
R.version.string
packageVersion("lme4")
packageVersion("ggplot2")
```

To output the full software manifest for thesis appendices:
```r
sessionInfo()
```

---

# Interactive Dashboards with Shiny

### Giving Thesis Stakeholders Exploratory Control

---

## From Static Reports to Dynamic Exploration

Static reports answer only the specific questions the author formulated.

What if a committee member asks:
- *"How does wellbeing vary if we isolate only full-time students?"*
- *"Can we zoom into Semester 3 and inspect sleep duration instead?"*

**Shiny** turns R analyses into responsive web applications **without requiring HTML, CSS, or JavaScript knowledge**.

---

## The Architecture of a Shiny App

Every Shiny application consists of two communicating halves:

```
               USER INTERFACE (UI)                    SERVER FUNCTION
          +---------------------------+          +-------------------------+
          |  Inputs:                  |          |                         |
          |  - Select Faculty Dropdown| -------> |  input$faculty          |
          |  - Metric Radio Buttons   |          |           |             |
          |                           |          |           v             |
          |  Outputs:                 |          |  selected <- reactive() |
          |  - plotOutput("trend")    | <------- |           |             |
          |  - tableOutput("summary") |          |  output$trend <-        |
          +---------------------------+          |    renderPlot({...})    |
                                                 +-------------------------+
```

---

## Reactivity: The Heart of Shiny

Shiny is built on a **directed acyclic reactivity graph**:

```
[ Input: input$faculty ] ---> [ reactive(selected) ] ---> [ output$trend (renderPlot) ]
                                                     ---> [ output$summary (renderTable) ]
[ Input: input$measure ] --------------------------------> [ output$trend (renderPlot) ]
```

### The Reactive Rule
- Code inside `reactive({...})` or `renderPlot({...})` automatically monitors its inputs.
- When `input$faculty` changes, Shiny invalidates downstream calculations and updates only the necessary screen widgets.

---

## Complete Shiny Dashboard: `app.R`

```r
library(shiny)
library(bslib)
library(dplyr)
library(ggplot2)
library(data2thesis)

wellbeing_data <- semesters |>
  left_join(students |> select(student_id, faculty, study_mode), join_by(student_id))

measures <- c("Wellbeing" = "wellbeing", "Sleep Hours" = "sleep_hours", "Study Hours" = "study_hours")

ui <- page_sidebar(
  title = "Doctoral Student Wellbeing Portal",
  sidebar = sidebar(
    selectInput("faculty", "Select Faculty:", choices = sort(unique(wellbeing_data$faculty))),
    radioButtons("measure", "Outcome Metric:", choices = measures)
  ),
  card(plotOutput("trend")),
  card(tableOutput("summary"))
)
```

---

## Shiny Server Logic & Dynamic Column Selection

```r
server <- function(input, output, session) {
  
  # Reactive subset: recomputed ONLY when input$faculty changes
  selected <- reactive({
    wellbeing_data |> filter(faculty == input$faculty)
  })

  # Dynamic plot: re-rendered when selected() or input$measure changes
  output$trend <- renderPlot({
    selected() |>
      summarise(mean = mean(.data[[input$measure]], na.rm = TRUE),
                .by = c(semester, study_mode)) |>
      ggplot(aes(x = semester, y = mean, colour = study_mode)) +
      geom_line(linewidth = 1.2) + geom_point(size = 3) +
      labs(y = names(measures)[measures == input$measure], x = "Semester") +
      theme_minimal(base_size = 14)
  })

  output$summary <- renderTable({
    selected() |>
      summarise(Students = n_distinct(student_id),
                Mean = mean(.data[[input$measure]], na.rm = TRUE),
                .by = semester)
  })
}

shinyApp(ui, server)
```

---

## Dynamic Column Selection: `.data[[...]]`

Notice the syntax inside `summarise()`:

```r
mean(.data[[input$measure]], na.rm = TRUE)
```

### Why not `mean(input$measure)`?
- `input$measure` evaluates to a character string: `"wellbeing"`.
- Calling `mean("wellbeing")` in R produces an error (`NA`).
- The tidy evaluation pronoun `.data[[input$measure]]` tells dplyr: *"Look inside the current dataframe and extract the column whose name matches the string in `input$measure`."*

---

## Dashboard Output: Faculty of Education

```{r}
#| label: fig-dashboard-demo
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
#| fig-align: center
library(dplyr)
library(ggplot2)
library(data2thesis)

semesters |>
  left_join(students |> select(student_id, faculty, study_mode), join_by(student_id)) |>
  filter(faculty == "Education") |>
  summarise(mean = mean(wellbeing, na.rm = TRUE), .by = c(semester, study_mode)) |>
  ggplot(aes(x = semester, y = mean, colour = study_mode)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = c("Full-time" = "#2b5c8f", "Part-time" = "#d95f02")) +
  labs(title = "Education Faculty: Wellbeing Across Semesters",
       x = "Semester", y = "Average Wellbeing (0-100)", colour = "Study Mode") +
  theme_minimal(base_size = 14)
```

---

## Where Can You Host a Shiny Dashboard?

How do thesis examiners or policymakers view your dashboard?

1. **shinyapps.io**:
   - Free cloud hosting for small applications provided by Posit.
   - Deploy directly with a single click from RStudio (`rsconnect::deployApp()`).
2. **Posit Connect / University Server**:
   - Institutional hosting behind firewalls.
3. **Shinylive (Serverless WebR)**:
   - Uses WebAssembly (Wasm) to compile R entirely inside the browser!
   - Requires zero backend servers; can be hosted free on GitHub Pages!

---

# Open Science & Research Integrity

### Transparency, Ethics, and the Garden of Forking Paths

---

## Open Science Practices

**Open Science** extends reproducibility from your personal laptop to the global research community:

```
  Open Data            Open Code            Open Protocol           Open Access
      |                    |                      |                      |
Anonymized CSVs/RData    Clean scripts       Preregistered design   Preprints & public
with full codebooks     with renv.lock       & analysis plan        DOIs (Zenodo/OSF)
```

### Permanent Identification: The DOI
- Never link to personal Google Drive or Dropbox links in your thesis (link rot).
- Deposit materials in **Zenodo** or the **Open Science Framework (OSF)**.
- Receive a permanent, citable **Digital Object Identifier (DOI)**.

---

## Open Licensing

Attaching an explicit license clarifies reuse rights for future researchers:

| Asset | Recommended License | Meaning |
|---|---|---|
| **Data & Prose** | **Creative Commons CC-BY 4.0** | Anyone may share and adapt, provided they give credit/citation. |
| **Public Domain**| **CC0 (Public Domain)** | Complete dedication to public domain with zero restrictions. |
| **R Code / Scripts**| **MIT License** | Highly permissive: anyone can run, modify, and distribute code. |
| **Academic Software**| **GPL-3.0** | Copyleft: derivative works must also remain open source. |

---

## Protecting Participants: The Small Cell Trap

::: {.callout-caution}
### Removing Names Does NOT Guarantee Anonymity!
Quasi-identifiers (age, gender, faculty, marital status, children) can easily isolate unique individuals.
:::

```{r}
#| echo: true
students |>
  count(faculty, study_mode, gender, has_children) |>
  arrange(n) |>
  head(4)
```

A cell with $N=1$ reveals the exact identity of a colleague!  
**Remedies**: Category collapsing, cell suppression ($N < 5$), or synthetic data.

---

## The Garden of Forking Paths

In any empirical study, researchers face hundreds of defensible choices:
- Should we drop outliers ($Z > 2.0$ or $Z > 3.0$)?
- Should we log-transform skewed response distributions?
- Should we run an independent $t$-test or a Mann-Whitney rank test?
- Should we control for age or leave it out?

> **Gelman & Loken (2013)**:  
> When these choices are made **after** viewing the data, researchers unconsciously steer toward whichever path yields $p < .05$.

Let us demonstrate this via an R simulation.

---

## Simulating the False-Positive Explosion

```{r}
#| echo: true
forking_paths <- function() {
  group     <- rep(c("A", "B"), each = 30)
  score     <- rexp(60, rate = 1 / 10)       # Null hypothesis true: No group difference!
  covariate <- rnorm(60)
  z <- abs(as.numeric(scale(score)))
  
  p <- c(
    planned_t_test = t.test(score ~ group)$p.value,
    no_outliers    = t.test(score[z < 2] ~ group[z < 2])$p.value,
    log_scale      = t.test(log(score) ~ group)$p.value,
    rank_test      = wilcox.test(score ~ group)$p.value,
    with_covariate = summary(lm(score ~ group + covariate))$coefficients[2, 4]
  )
  c(planned = p[["planned_t_test"]] < 0.05, any_post_hoc = any(p < 0.05))
}

set.seed(17)
sim_results <- rowMeans(replicate(2000, forking_paths()))
```

---

## Simulation Findings: Inflation from 5% to 23%!

```{r}
#| echo: true
sim_results
```

```
Planned Analysis: [■                   ]  5.3% False Positive Rate (Expected alpha)
Any of 5 Choices: [■■■■■■■■■           ] 23.4% False Positive Rate!
```

- Each individual analysis was standard, defensible, and legitimate.
- But testing 5 options post-hoc inflated the error rate to **nearly 1 in 4**!
- **Preregistration** eliminates this vulnerability.

---

## Preregistration & Registered Reports

How do we eliminate the Garden of Forking Paths?

### 1. Standard Preregistration
- Freeze your hypotheses, sample size, exclusion rules, and planned statistical tests on **OSF** or **AsPredicted.org** *before* viewing the data.
- Draw a transparent line between:
  - **Confirmatory Analyses**: Tested against the preregistered protocol.
  - **Exploratory Analyses**: Serendipitous findings discovered during data exploration.

### 2. Registered Reports
- The study protocol and design are peer-reviewed and **provisionally accepted for publication before data collection begins**!
- Publication does not depend on whether $p < .05$.

---

## Software Citation & Professional Attribution

Scientific software is built by researchers whose careers depend on citation:

```{r}
#| echo: true
citation("lme4")
```

---

# Version Control with Git (Optional)

### Eliminating `thesis_v2_final_final.docx`

---

## Why Version Control for Theses?

```
Graduate Student Folder:
├── thesis_draft.qmd
├── thesis_draft_supervisor_comments.qmd
├── thesis_draft_supervisor_comments_v2.qmd
├── thesis_draft_final.qmd
└── thesis_draft_final_REALLY_FINAL_SUBMIT.qmd
```

### Git replaces this chaos:
- Keeps a single, canonical filename (`thesis.qmd`).
- Records a complete, immutable time-capsule history of every single edit.
- Lets you inspect historical diffs or revert to any prior state safely.

---

## The Git Mental Model

```
   WORKING DIRECTORY                 STAGING AREA                LOCAL REPOSITORY
+-----------------------+        +-------------------+        +--------------------+
|  Edited thesis.qmd    | -----> |   git add .       | -----> | git commit -m "..."|
|  Added new chart      |        | (Marked for save) |        | (Snapshot saved)   |
+-----------------------+        +-------------------+        +--------------------+
                                                                        |
                                                                        | git push
                                                                        v
                                                              +--------------------+
                                                              |  REMOTE (GitHub)   |
                                                              | (Offsite backup &  |
                                                              |  collaboration)    |
                                                              +--------------------+
```

---

## RStudio Git Pane & `.gitignore`

RStudio provides a visual interface for Git:

1. Turn any project into a Git repository:
   ```r
   usethis::use_git()
   usethis::use_github()
   ```
2. Tick modified files in the **Git Pane**, write a descriptive commit message (*"Standardize APA table styling"*), and click **Commit**.

::: {.callout-warning}
### Data Privacy and `.gitignore`
Never push private participant data to public repositories!  
Add data files to your project's `.gitignore`:
```
data/raw_surveys.csv
data/*.xlsx
.Renviron
```
:::

---

## The Thesis Reproducibility Checklist

Before submitting your doctoral thesis:

- [ ] **Single Command Render**: The complete thesis compiles from raw data via `quarto render`.
- [ ] **Zero Hard-Coded Numbers**: All descriptive statistics and test values in text use inline R code.
- [ ] **Frozen Dependencies**: `renv.lock` accurately captures all package versions.
- [ ] **Cross-References Verified**: All tables (`@tbl-`) and figures (`@fig-`) resolve cleanly.
- [ ] **Participant Privacy Audited**: Quasi-identifiers checked; small cells ($N < 5$) protected.
- [ ] **Public Archive Ready**: Anonymized data, cleaning scripts, and codebook assigned a permanent DOI.

---

## Chapter Summary

- **Reproducibility is Integrity**: An empirical result that cannot be recomputed cannot be trusted.
- **Quarto Literate Programming**: Seamlessly blends YAML metadata, Markdown narrative, executable R chunks, and inline computation into HTML, Word, and PDF.
- **Deterministic Numbering**: Dynamic cross-referencing and automatic BibTeX citations guarantee alignment across draft revisions.
- **Environment Pinning**: `renv` isolates project libraries and guarantees your analysis runs in 2030 exactly as it did today.
- **Interactive Dashboards**: Shiny brings your statistical models to life for examiners and policymakers through reactive data streams.
- **Open Science & Ethics**: Preregistration protects against the garden of forking paths; participant privacy is preserved through small-cell suppression.

---

## Recommended Exercises & Next Steps

1. **Inline Computation Practice**: Take your Chapter 6 descriptive summary and replace every hard-coded mean with ` `r knitr::inline_expr("...")` `.
2. **Multi-Format Compilation**: Render your results chapter simultaneously to HTML and Microsoft Word with an APA reference template.
3. **Build a Shiny Filter**: Extend the wellbeing dashboard with a `sliderInput()` that filters data by academic semester range.
4. **Environment Freezing**: Initialize `renv` in your research project and create your first `renv.lock` snapshot.
5. **Simulate Forking Paths**: Modify `forking_paths()` with a 6th analysis (e.g. median test) and measure the false-positive shift.
