---
title: "Using AI with R"
description: "Chapter 18: the reliability of coding and Cohen's kappa, how language models work, AI as a coding assistant, coding open-ended answers with a language model through ellmer, and using AI responsibly."
format: deck-html
execute:
  echo: false
  freeze: auto
---

```{r}
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(stringr)
library(yardstick)

navy <- "#1d588c"; copper <- "#b4762e"; soft <- "#46586d"
theme_set(theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "transparent", colour = NA),
        axis.title = element_text(colour = soft)))
knitr::opts_chunk$set(dev = "svg", dev.args = list(bg = "transparent"),
                      message = FALSE, warning = FALSE)
f2 <- function(x) sprintf("%.2f", x)

coder_a <- c(rep("Workload", 16), rep("Family", 4))
coder_b <- c(rep("Workload", 14), rep("Family", 2), rep("Workload", 2), rep("Family", 2))
observed <- mean(coder_a == coder_b)
chance <- sum(prop.table(table(coder_a)) * prop.table(table(coder_b)))
kappa_ex <- (observed - chance) / (1 - chance)

wrong <- students |> summarise(percent = 100 * sum(considering_dropout == "Yes") / nrow(students), .by = faculty)
right <- students |> summarise(percent = 100 * mean(considering_dropout == "Yes"), .by = faculty)

keywords <- c(
  Supervision = "\\b(supervis|feedback|guidance|meeting)",
  Workload    = "\\b(time|deadline|workload|too much|assignment|reading|busy)",
  Finances    = "\\b(money|fee|scholarship|rent|afford|salary|income|funding|stipend|loan)",
  Family      = "\\b(famil|child|kid|son\\b|daughter|baby|parent|mother|father|husband|wife)",
  Health      = "\\b(sleep|tired|stress|anxi|health|ill\\b|burnout|headache|coffee|exhaust)",
  Isolation   = "\\b(lonel|alone|friends|outsider|miss (home|my)|far from|know no one)",
  Other       = "\\b(ethic|participant|statistic|library|procedure|english|power cut|laptop|internet|software)")
code_by_keywords <- function(answer) {
  hits <- sapply(keywords, \(pattern) str_count(str_to_lower(answer), pattern))
  if (all(hits == 0)) "Other" else names(keywords)[which.max(hits)]
}
themes <- names(keywords)
keyword_check <- open_responses_coded |>
  mutate(keyword_theme = sapply(biggest_challenge, code_by_keywords),
         theme = factor(theme, levels = themes), keyword_theme = factor(keyword_theme, levels = themes))
kw_acc <- accuracy(keyword_check, theme, keyword_theme)$.estimate
kw_kap <- kap(keyword_check, theme, keyword_theme)$.estimate

ai <- read.csv("../../data-raw/ai_coding_results.csv")
ai_meta <- read.csv("../../data-raw/ai_coding_meta.csv")
ai_check <- open_responses_coded |>
  left_join(ai, join_by(student_id)) |>
  mutate(across(c(theme, ai_theme, ai_theme_rerun), \(x) factor(x, levels = themes)))
ai_acc <- accuracy(ai_check, theme, ai_theme)$.estimate
ai_kap <- kap(ai_check, theme, ai_theme)$.estimate
rerun_agree <- mean(ai_check$ai_theme == ai_check$ai_theme_rerun, na.rm = TRUE)
kappa_label <- function(k) as.character(cut(k, c(-Inf, 0.2, 0.4, 0.6, 0.8, Inf),
  labels = c("slight", "fair", "moderate", "substantial", "almost perfect")))
q_support <- questionnaire |>
  mutate(support = rowMeans(pick(support_1:support_6), na.rm = TRUE)) |> select(student_id, support)
```

# Using AI with R

Draft with AI, check everything, and stay responsible

**Chapter 18**

Polla Fattah

---

## By the end of today you can

- explain how the reliability of coding is measured, and why chance matters;
- explain how large language models work, and why they make confident mistakes;
- use an AI assistant to write, explain, and debug R code, and check its answers;
- call a language model from R with ellmer, with structured answers;
- validate AI coding against human coding with accuracy and Cohen's kappa;
- use AI responsibly: protect data, keep work reproducible, and disclose it.

---

## Text as research data

"What has been your biggest challenge during your studies?"

`r nrow(open_responses)` students answered in their own words (RQ11).

`r nrow(open_responses_coded)` answers were coded by hand into themes, which took most of two weeks. Could an AI tool code the rest reliably?

---

## Qualitative coding

Reading each text and assigning it to a theme defined in a **codebook**.

Coding is judgement, and judgements differ.

Whether the coder is a person, a keyword list, or a program, the question is the same: does it code reliably?

---

## Raw agreement can mislead

```r
coder_a <- c(rep("Workload", 16), rep("Family", 4))
coder_b <- c(rep("Workload", 14), rep("Family", 2),   # the first 16
             rep("Workload", 2),  rep("Family", 2))   # the last 4
```

| Observed agreement | Chance agreement | Cohen's kappa |
|---|---|---|
| `r round(100 * observed)`% | `r round(100 * chance)`% | `r f2(kappa_ex)` |

Each coder puts 80% under Workload, so random coders would agree 68% of the time.

---

## Cohen's kappa

$$
\kappa = \frac{\text{observed} - \text{chance}}{1 - \text{chance}}
$$

| Kappa | Agreement |
|---|---|
| 0 | no better than chance |
| 0.41 to 0.60 | moderate |
| 0.61 to 0.80 | substantial |
| above 0.80 | almost perfect |

`kap()` from yardstick. The standard measure of **inter-rater reliability**.

---

## How large language models work

- a neural network (Chapter 15) with billions of weights;
- trained on enormous amounts of text: books, websites, code;
- learns to predict the next **token**, a word or part of one;
- answers by generating one likely token at a time;
- further training turns it into an assistant that follows instructions.

---

## Why they make confident mistakes

| Strength or weakness | Because |
|---|---|
| fluent in R | much R was in the training text |
| **hallucination**: plausible inventions | it produces what is plausible, not verified |
| out of date | knowledge stops at the training cut-off |
| knows nothing about your data | unless you tell it |

> **Use AI to draft, and check everything it produces.**

---

## Asking good questions

| Include | Example |
|---|---|
| the goal | the percentage per faculty who considered dropping out |
| the data | column names and types, e.g. from `glimpse()`; never identifiable data |
| the tools | dplyr with the native pipe |
| for errors | the exact message and the code |

---

## A plausible answer that is wrong

```r
students |>
  summarise(percent = 100 * sum(considering_dropout == "Yes") / nrow(students),
            .by = faculty)
```

It runs, and the numbers look reasonable.

But they add up to `r round(sum(wrong$percent))`: `nrow(students)` counts every student, not those in each faculty.

---

## The corrected code

```r
students |>
  summarise(percent = 100 * mean(considering_dropout == "Yes"),
            .by = faculty)
```

The mean of TRUE/FALSE is the proportion TRUE, within each faculty (Chapter 6).

Not a hallucination but a plausible misunderstanding: the most dangerous kind, because nothing looks wrong.

---

## Four checks

- run the code on a **tiny example** with a known answer;
- check a number **by another route**: `count(faculty, considering_dropout)`;
- read the **help page**: does the function exist and do that?
- ask the assistant to **explain** each line.

Assistants are excellent teachers: "explain this line by line", "why does this give NA?".

---

## AI inside the editor

- **GitHub Copilot** in RStudio suggests code as you type, even from a comment;
- **Positron Assistant** and similar chats can see open files and, with permission, your data's structure.

Suggestions arrive fast and look authoritative. Read each one before accepting it.

---

## The answers to be coded

```r
open_responses |> slice(c(5, 9, 15, 55)) |> pull(biggest_challenge)
```

They vary in length and style, often mention more than one challenge, and contain spelling mistakes.

Seven themes. The hand-coded sample is the benchmark, like a test set (Chapter 11).

---

## The hand-coded themes

```{r}
#| results: asis
open_responses_coded |> count(theme, sort = TRUE) |> knitr::kable()
```

---

## Coding by keywords

```r
keywords <- c(
  Supervision = "\\b(supervis|feedback|guidance|meeting)",
  Finances    = "\\b(money|fee|scholarship|rent|afford|salary|income|funding)",
  Family      = "\\b(famil|child|kid|son\\b|daughter|baby|parent|husband|wife)"
)  # plus Workload, Health, Isolation, and Other in the full list
code_by_keywords <- function(answer) {
  hits <- sapply(keywords, \(pattern) str_count(str_to_lower(answer), pattern))
  if (all(hits == 0)) "Other" else names(keywords)[which.max(hits)]
}
```

A **regular expression**: `|` means "or", `\\b` marks the start of a word.

---

## Words are not meaning

```r
code_by_keywords("I miss my family and I cannot afford the fees.")
#> [1] "Finances"
```

One Family word, one Isolation phrase, two Finances words: Finances wins, where a reader might disagree.

| Keywords against hand coding | Accuracy | Kappa |
|---|---|---|
| | `r round(100 * kw_acc)`% | `r f2(kw_kap)` |

"My supervisor" in an answer about letting down one's family is about family.

---

## Calling a language model from R

| Kind | Examples | Needs |
|---|---|---|
| online | `chat_anthropic()`, `chat_openai()`, `chat_google_gemini()` | an API key; small cost per request |
| local | `chat_ollama()` through Ollama | a download; no key, no cost, text stays on your computer |

Store an API key in `.Renviron` (`usethis::edit_r_environ()`), never in a shared script.

---

## A chat from R

```r
library(ellmer)
chat <- chat_ollama(model = "gemma4:e4b")            # local
# chat <- chat_anthropic(model = "claude-sonnet-5") # online, needs a key
chat$chat("In one sentence, what does Cohen's kappa measure?")
```

Always set the `model` explicitly and record it: models are updated and retired.

The study used a local model, so answers never left the computer and anyone can repeat it for free.

---

## The codebook as a system prompt

```r
codebook <- "
You are helping a researcher code open-ended survey answers ...
Assign each answer to exactly ONE theme: the main challenge.
Themes:
- Supervision: the supervisor; feedback, guidance, meetings.
- Workload: too much work or too little time; deadlines, reading.
- Finances: money, fees, scholarships, living costs.
- Family: children, partners, parents, caring duties.
- Health: physical or mental health, sleep, stress, burnout.
- Isolation: loneliness, far from home, no colleagues.
- Other: anything else, such as ethics approval or statistics.
"
```

The same instructions for every answer.

---

## Structured answers for many texts

```r
theme_type <- type_object(
  theme = type_enum(themes, "The single main theme of the answer."))

chat <- chat_anthropic(system_prompt = codebook, model = "claude-sonnet-5",
                       params = params(temperature = 0))

ai_codes <- parallel_chat_structured(
  chat, prompts = as.list(open_responses$biggest_challenge),
  type = theme_type)
```

`type_enum()` restricts replies to the themes. `temperature = 0` makes answers as repeatable as possible.

---

## Checking a local model's replies in code

```r
chat <- chat_ollama(system_prompt = codebook, model = "gemma4:e4b",
                    params = params(temperature = 0))
replies  <- parallel_chat_text(chat, as.list(open_responses$biggest_challenge))
ai_codes <- themes[match(tolower(gsub("[^A-Za-z]", "", replies)),
                         tolower(themes))]
```

Smaller models may ignore the structured format. Ask for the theme name only, then let R check it.

A reply that is not a theme becomes `NA`: counted, not silently accepted.

---

## Run once, save, analyse the file

The script records the provider, model, date, software versions, and time taken, and codes the hand-coded sample twice.

| | Value |
|---|---|
| model | `r ai_meta$model` |
| requests | `r ai_meta$requests` |
| time | about `r round(ai_meta$minutes)` minutes |
| replies outside the themes | `r ai_meta$failed` |

---

## The model against the hand coding

```r
ai_check |> accuracy(truth = theme, estimate = ai_theme)
ai_check |> kap(truth = theme, estimate = ai_theme)
```

| Coder | Accuracy | Kappa |
|---|---|---|
| keywords | `r round(100 * kw_acc)`% | `r f2(kw_kap)` |
| language model | `r round(100 * ai_acc)`% | `r f2(ai_kap)` |

The model reaches `r kappa_label(ai_kap)` agreement with the hand coding.

---

## Read the disagreements

```{r}
#| results: asis
ai_check |> filter(theme != ai_theme) |>
  mutate(biggest_challenge = str_trunc(biggest_challenge, 70)) |>
  select(Answer = biggest_challenge, `Hand coding` = theme, Model = ai_theme) |>
  head(4) |> knitr::kable()
```

Is the model wrong, the codebook unclear, or the answer ambiguous? A clearer rule helps human and machine alike.

---

## Is the model repeatable?

Asked a second time, the model gave the same theme for `r round(100 * rerun_agree)`% of the answers.

That is not guaranteed for every model or setting; online models can change between runs.

So the outputs are saved and analysed from the file.

---

## What the students said

```{r}
#| fig-width: 8
#| fig-height: 3.6
ai |> count(ai_theme) |> mutate(share = n / sum(n)) |>
  ggplot(aes(share, reorder(ai_theme, share))) +
  geom_col(fill = navy) + scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Share of answers", y = NULL)
```

All `r nrow(ai)` answers, coded by the model once its agreement was checked.

---

## Text and numbers tell the same story

```{r}
#| results: asis
ai |> left_join(q_support, join_by(student_id)) |>
  summarise(students = n(), support = round(mean(support, na.rm = TRUE), 2), .by = ai_theme) |>
  arrange(support) |> rename(theme = ai_theme, `mean support` = support) |> knitr::kable()
```

Students whose main challenge is supervision report the least supervisor support: a check of **validity**.

---

## Check, and stay responsible

- run code on known cases;
- check facts and references against sources: assistants invent plausible references;
- check coding against human coding.

Whatever appears in your thesis is your responsibility, whoever or whatever drafted it.

---

## Protect your data

- text sent online leaves your computer, and may be stored;
- check the **ethics approval and consent forms** first;
- check the provider's terms: free accounts often allow training on your data;
- remove names, places, and identifying details;
- for sensitive data, use a **local model**, and validate it too.

---

## Keep it reproducible

| Record | Why |
|---|---|
| provider, exact model, date | models change and retire |
| settings such as temperature | they change answers |
| system prompt and codebook | the instructions are part of the method |
| the model's outputs | analyse the saved file, not a new call |

Code a sample twice to check consistency.

---

## Disclose it

An AI tool cannot be an author: it cannot take responsibility (COPE).

> Answers were coded into seven themes by a large language model (Gemma 4, e4b version, run locally with Ollama and the ellmer R package, temperature 0) using the codebook in Appendix X. Agreement with the author's hand coding of a random sample of `r nrow(open_responses_coded)` answers was assessed with Cohen's kappa.

---

## Be aware of bias

Models learn from human text and can reproduce its biases:

- in how they describe groups of people;
- in which answers they find typical;
- in how well they understand non-standard English.

Check whether accuracy differs between groups, as for any model about people.

---

## In your field: medicine and public health

```r
medication_type <- type_object(
  drug          = type_string("Name of the medication"),
  dose_mg       = type_number("Dose in milligrams"),
  times_per_day = type_integer("How many times a day it is taken"))

chat$chat_structured(
  "Patient to continue metformin 500 mg twice daily with meals.",
  type = medication_type)
```

Free text into fixed fields. Validate against clinician-checked records, and never send patient data to public services.

---

## Practical lab: the Chapter 18 playground

Work through the playground exercises in your browser, with hints and solutions.

The saved model coding is included, so no API key is needed.

---

## Practical exercises 1–3: assistants and keywords

1. Ask an assistant for average sleep by faculty and semester, and check it two ways.
2. Add three keywords: does kappa improve, or are you overfitting?
3. Read ten keyword disagreements: do you agree with the hand coding?

---

## Practical exercises 4–6: models and reliability

4. Code 20 answers with a language model; change one theme definition and recode.
5. Write the methods paragraph for AI coding in your own study.
6. A coder who always says Workload: raw agreement and kappa?

---

## Try this yourself

Plan AI use in your own research.

- what would you ask it to do, and what would you check by hand?
- does your consent form allow sending data to a company?
- which local model could you use instead?
- what would you record, and how would you disclose it?

---

## Troubleshooting guide (Part 1)

| Symptom | Likely cause |
|---|---|
| high agreement, low kappa | one theme dominates; chance inflates agreement |
| a function that does not exist | hallucination; check the help page |
| code runs but answers the wrong question | a plausible misunderstanding; test it |
| an outdated way of doing something | the training cut-off |

---

## Troubleshooting guide (Part 2)

| Symptom | Likely cause |
|---|---|
| replies in the wrong format | a small model ignored structure; check in code |
| different answers on another day | model or settings changed; save outputs |
| an API key in a shared script | store it in `.Renviron` |
| the model disagrees on two-challenge answers | the codebook's rule is unclear |

---

## Completion checklist

- [ ] AI-drafted code is tested on known cases;
- [ ] AI coding is validated against human coding with kappa;
- [ ] disagreements are read, and the codebook improved;
- [ ] data sent to a model is allowed by consent and anonymised;
- [ ] provider, model, date, settings, and prompts are recorded;
- [ ] outputs are saved and analysed from the file;
- [ ] AI use is disclosed.

---

## Misconceptions to leave behind (Part 1)

| Misconception | Better mental model |
|---|---|
| high agreement means reliable coding | kappa shows agreement beyond chance |
| the AI knows the answer | it produces plausible text, to be validated |

---

## Misconceptions to leave behind (Part 2)

| Misconception | Better mental model |
|---|---|
| a model's answers never change | versions, settings, and dates change them |
| AI use must be hidden | most allow it with disclosure |

---

## The chapter in one sentence

> **Treat AI output as a draft: validate it against human judgement, protect participants, record what you did, and say so.**

---

## Next: Chapter 19

The final chapter puts everything together:

- the shape of a research project;
- from raw export to clean data, and describing the sample;
- answering each research question;
- the chain of reasoning and the results chapter;
- choosing a method, and where to go next.

---

## Questions

Which part of your own analysis would you trust an AI to draft?

How would you check it?
