---
title: "Artificial Intelligence in Academic Research"
subtitle: "Coding Assistants, Qualitative NLP with ellmer, Cohen's Kappa, and Research Ethics"
author: "Dr. Polla Fattah"
institute: "From Data to Thesis: Research Data Analysis with R"
format:
  revealjs:
    theme: [default, ../../styles/theme.scss]
    slide-number: true
    preview-links: auto
    chalkboard: true
    transition: slide
    highlight-style: github
    code-line-numbers: false
    width: 1200
    height: 750
execute:
  freeze: auto
---

# Artificial Intelligence with R

### Augmenting Quantitative and Qualitative Science

**Chapter 18 · From Data to Thesis**

Dr. Polla Fattah

---

## When Research Data is Unstructured Text

Much of what researchers study cannot be captured on a 5-point Likert scale:

- Open-ended survey responses
- In-depth qualitative interview transcripts
- Clinical admission notes and doctor discharge summaries
- Policy documents, historical archives, and committee minutes

> **The Methodological Bottleneck**:  
> Reading and categorizing 1,000 open-ended responses by hand can take weeks of full-time labor.

Can modern **Large Language Models (LLMs)** assist with qualitative coding without compromising scientific rigor?

---

## The Two Faces of AI in Research

In modern data science, Artificial Intelligence serves two distinct roles:

```
+-----------------------------------------------------------------------------------+
|                            ARTIFICIAL INTELLIGENCE                                |
+----------------------------------------+------------------------------------------+
|          1. CODING ASSISTANT           |          2. QUALITATIVE ANALYST          |
|  - Generates R code drafts             |  - Reads open-ended text                 |
|  - Explains complex error traces       |  - Classifies responses into themes      |
|  - Refactors tidyverse pipelines       |  - Extracts clinical entities from notes |
|  - Interactive debugging partner       |  - Validated against human benchmarks    |
+----------------------------------------+------------------------------------------+
```

In both roles, **the human researcher remains legally and intellectually responsible** for every line of code and every published finding.

---

# Inter-Rater Reliability

### Measuring Agreement Beyond Mere Chance

---

## The Problem with Raw Agreement

Suppose two research assistants code 20 student responses for their primary challenge: **Workload** vs. **Family**.

```{r}
#| echo: true
coder_a <- c(rep("Workload", 16), rep("Family", 4))
coder_b <- c(rep("Workload", 14), rep("Family", 2),
             rep("Workload", 2),  rep("Family", 2))

# Raw percentage agreement:
mean(coder_a == coder_b)
```

**80% agreement sounds impressive!** But does this reflect genuine coding consistency?

---

## The Base-Rate Illusion

Look closer at the marginal distributions:
- Coder A categorized **80%** of all responses as *Workload*.
- Coder B categorized **80%** of all responses as *Workload*.

```
Chance agreement on Workload:  0.80 × 0.80 = 0.64
Chance agreement on Family:    0.20 × 0.20 = 0.04
Total Expected Chance Agreement = 0.64 + 0.04 = 0.68 (68%)
```

Two coders throwing darts at a board with these base rates would agree **68% of the time purely by luck**!

The observed $80\%$ agreement gained very little ground over sheer coincidence.

---

## Cohen's Kappa ($\kappa$)

Jacob Cohen (1960) formulated **Kappa** to measure agreement corrected for chance:

$$\kappa = \frac{P_o - P_e}{1 - P_e}$$

Where:
- $P_o$ = **Observed proportion of agreement**
- $P_e$ = **Expected agreement under independence (chance)**
- $1 - P_e$ = Maximum possible agreement achievable beyond chance

$$\kappa = \frac{0.80 - 0.68}{1 - 0.68} = \frac{0.12}{0.32} = \mathbf{0.375}$$

What looked like an impressive $80\%$ is actually **poor-to-fair reliability** ($\kappa = 0.38$).

---

## Interpreting Cohen's Kappa

The standard benchmarks established by Landis & Koch (1977):

| Cohen's $\kappa$ Range | Strength of Agreement | Acceptable for Science? |
|---|---|---|
| $< 0.20$ | Slight agreement | Unacceptable |
| $0.21 - 0.40$ | Fair agreement | Problematic / Revise codebook |
| $0.41 - 0.60$ | Moderate agreement | Minimum threshold for exploratory work |
| $0.61 - 0.80$ | Substantial agreement | Good scientific reliability |
| $0.81 - 1.00$ | Almost perfect | Gold standard benchmark |

In `tidymodels` / `yardstick`, calculate it directly:
```r
yardstick::kap(df, truth = human_code, estimate = ai_code)
```

---

# How Large Language Models Work

### Next-Token Probability Engines, Not Truth Oracles

---

## The Autoregressive Mechanism

LLMs (ChatGPT, Claude, Gemini, Gemma, Llama) are massive artificial neural networks trained on hundreds of billions of words.

At their core, they perform one mathematical task:

$$P(w_{t} \mid w_1, w_2, \dots, w_{t-1})$$

> **Predicting the next token** (word or sub-word piece) given the entire sequence that came before it.

```
Input: "The student was stressed because of the upcoming ..."
Probabilities:
  - "deadline"  (58%)
  - "exam"      (31%)
  - "meeting"    (8%)
  - "banana"    (0.00001%)
```

The model outputs text by sampling from this conditioned probability distribution.

---

## What LLMs Excel At vs Where They Fail

```
+-------------------------------------+-------------------------------------+
|         WHAT LLMs DO WELL           |         WHERE LLMs FAIL             |
+-------------------------------------+-------------------------------------+
| - Writing standard R / tidyverse    | - Plausible Hallucinations          |
| - Explaining error messages         |   (Inventing fake functions & args) |
| - Translating between languages     | - Fabricating academic citations    |
| - Synthesizing qualitative text     | - Silent logic bugs in calculations |
| - Adhering to structured schemas    | - Knowledge cutoffs (outdated APIs) |
+-------------------------------------+-------------------------------------+
```

::: {.callout-important}
### Plausibility $\neq$ Veracity
An LLM does not generate text because it knows it is true. It generates text because it sounds **statistically probable**.
:::

---

# AI as an R Coding Assistant

### Best Practices for Pair Programming

---

## Crafting High-Precision Prompts

An assistant can only be as accurate as your context:

```
POOR PROMPT:
"Write R code to get the dropout rates."
(Result: Generic code, assumed column names, likely base-R or outdated syntax)

HIGH-PRECISION PROMPT:
"I have a tibble called students. Here is glimpse(students):
- student_id: chr
- faculty: fct
- considering_dropout: chr ('Yes', 'No')

Calculate the percentage of students considering dropout WITHIN each faculty.
Use modern dplyr syntax with the native pipe |>."
```

---

## The Subtle Trap of Plausible Logic Bugs

Suppose you ask an assistant for the dropout percentage by faculty:

```r
# Assistant's Suggestion:
students |>
  summarise(
    percent = 100 * sum(considering_dropout == "Yes") / nrow(students),
    .by = faculty
  )
```

### The Danger:
- It executes without any error!
- The numbers look reasonable ($3.2\%, 4.1\%, 2.8\%$).
- **The flaw**: `nrow(students)` divides by the **entire university population**, not the faculty total! The percentages sum to the overall rate rather than each faculty's internal rate!

---

## The Corrected Calculation

To compute the proportion *within* each faculty group:

```{r}
#| echo: true
library(dplyr)
library(data2thesis)

students |>
  summarise(
    n_students = n(),
    n_dropout  = sum(considering_dropout == "Yes"),
    percent    = round(100 * mean(considering_dropout == "Yes"), 1),
    .by        = faculty
  ) |>
  arrange(desc(percent))
```

The mean of a logical vector (`considering_dropout == "Yes"`) is mathematically identical to the internal proportion!

---

## Four Verification Heuristics

Whenever using AI-generated code in your research:

1. **Test on Toy Data**: Run the code on a 4-row handmade tibble where you know the true answer by mental arithmetic.
2. **Triangulate by Another Route**: Verify summary numbers against simple counts:
   ```r
   students |> count(faculty, considering_dropout)
   ```
3. **Verify Function Help**: Check `?function_name` in the R console to ensure suggested arguments actually exist in your installed package version.
4. **Demand Explanations**: Ask: *"Explain line 3. Why did you use .data[[...]] instead of standard evaluation?"*

---

# Qualitative Coding with R & LLMs

### The RQ11 Case Study: Graduate Student Challenges

---

## Research Question 11

In the university wellbeing dataset, students completed an open-ended prompt:

> *"What has been your biggest challenge during your studies?"*

```{r}
#| echo: true
library(stringr)

open_responses |>
  slice(c(5, 9, 15, 55)) |>
  pull(biggest_challenge)
```

**The Goal**: Categorize hundreds of free-text responses into a standardized codebook.

---

## The Seven-Theme Codebook

Elaf developed a qualitative codebook with seven distinct themes:

```
1. Supervision : Supervisor friction, lack of feedback, missed meetings.
2. Workload    : Excessive volume, tight deadlines, time management.
3. Finances    : Tuition fees, living expenses, need for outside employment.
4. Family      : Childcare, eldercare, partner commitments, home distractions.
5. Health      : Physical illness, mental health, anxiety, chronic sleep loss.
6. Isolation   : Loneliness, imposter syndrome, lack of peer community.
7. Other       : Bureaucracy, ethics clearance, English proficiency, tech issues.
```

Elaf hand-coded a representative benchmark sample ($N = 100$) over two weeks.

---

## Baseline Method: Keyword Dictionary

Before jumping to deep neural networks, establish a transparent baseline using regular expressions:

```{r}
#| echo: true
keywords <- c(
  Supervision = "\\b(supervis|feedback|guidance|meeting)",
  Workload    = "\\b(time|deadline|workload|too much|assignment|reading|busy)",
  Finances    = "\\b(money|fee|scholarship|rent|afford|salary|income|funding|stipend)",
  Family      = "\\b(famil|child|kid|son\\b|daughter|baby|parent|mother|father)",
  Health      = "\\b(sleep|tired|stress|anxi|health|ill\\b|burnout|exhaust)",
  Isolation   = "\\b(lonel|alone|friends|outsider|miss (home|my)|far from)",
  Other       = "\\b(ethic|participant|statistic|library|procedure|english)"
)
```

---

## Keyword Classification Logic

Assign each response to the theme with the highest keyword count:

```{r}
#| echo: true
code_by_keywords <- function(answer) {
  hits <- sapply(keywords, \(pattern) str_count(str_to_lower(answer), pattern))
  if (all(hits == 0)) "Other" else names(keywords)[which.max(hits)]
}

code_by_keywords("I miss my family and I cannot afford the tuition fees.")
```

- Matches 1 Family keyword ("family")
- Matches 2 Finance keywords ("afford", "fees")
- Classifies as: **Finances**

---

## Evaluating Keyword Performance

How well does the dictionary match Elaf's manual coding?

```{r}
#| echo: true
library(yardstick)

themes <- names(keywords)

keyword_check <- open_responses_coded |>
  mutate(
    keyword_theme = sapply(biggest_challenge, code_by_keywords),
    theme         = factor(theme, levels = themes),
    keyword_theme = factor(keyword_theme, levels = themes)
  )

keyword_check |> accuracy(truth = theme, estimate = keyword_theme)
keyword_check |> kap(truth = theme, estimate = keyword_theme)
```

**Results**: Accuracy $\approx 61\%$, $\kappa \approx 0.54$ (Moderate agreement).  
**Why it fails**: Keywords cannot parse semantic context, negation, or competing clauses.

---

# Calling LLMs with `ellmer`

### Seamless R Integration with Cloud and Local Models

---

## Introducing the `ellmer` Package

**`ellmer`** (developed by Hadley Wickham & Posit) connects R directly to state-of-the-art language models:

```
                       ellmer in R
                            |
           +----------------+----------------+
           |                                 |
   Cloud API Models                  Local Models
   (Claude, OpenAI, Gemini)          (Ollama / Gemma / Llama)
   - Highest accuracy                - 100% On-device privacy
   - Requires API key                - Free / Zero API costs
   - Data sent over internet         - Complies with strict ethics
```

---

## Protecting API Keys in R

::: {.callout-caution}
### Never Hard-Code API Keys in Your Scripts!
If you push your script to GitHub, automated scrapers will steal your key within seconds.
:::

Store keys in your user-level `.Renviron` file:

```r
# Open .Renviron safely in RStudio
usethis::edit_r_environ()
```

Add your credentials:
```bash
ANTHROPIC_API_KEY=sk-ant-api03-...
OPENAI_API_KEY=sk-proj-...
GEMINI_API_KEY=AIzaSy...
```

R loads these environment variables silently on startup.

---

## Running Models Locally with Ollama

For sensitive human participant data, use **Ollama** ([ollama.com](https://ollama.com)):

1. Install Ollama on your computer.
2. In your terminal, pull a compact, high-performance open model:
   ```bash
   ollama pull gemma4:e4b
   ```
3. Connect from R with zero configuration:
   ```r
   library(ellmer)
   chat <- chat_ollama(model = "gemma4:e4b")
   chat$chat("What is Cohen's kappa?")
   ```
- **Data never leaves your machine.**
- Completely compliant with institutional IRB / Ethics boards!

---

## Structured Output Schemas

When processing 1,000 texts, we cannot accept rambling conversational replies. We need **machine-readable structured output**.

```r
codebook_prompt <- "
You are an expert qualitative researcher coding student survey responses.
Assign each text to exactly ONE theme from the codebook:
Supervision, Workload, Finances, Family, Health, Isolation, Other.
"

# Define the strict schema:
theme_type <- type_object(
  theme = type_enum(themes, description = "The primary challenge theme")
)

chat <- chat_anthropic(
  system_prompt = codebook_prompt,
  model         = "claude-sonnet-5",
  params        = params(temperature = 0) # Deterministic
)
```

Setting `temperature = 0` forces the model to select greedy, reproducible tokens.

---

## Batch Qualitative Coding

Process an entire survey column in parallel:

```r
ai_results <- parallel_chat_structured(
  chat,
  prompts = as.list(open_responses$biggest_challenge),
  type    = theme_type
)
```

```
+------------+-------------------------------------------+-------------+
| student_id | biggest_challenge                         | theme       |
+------------+-------------------------------------------+-------------+
| STU_001    | Balancing lab experiments with child care | Family      |
| STU_002    | Tuition fees are far too high this term   | Finances    |
| STU_003    | Disagreements with my primary supervisor  | Supervision |
+------------+-------------------------------------------+-------------+
```

---

## Evaluating the AI Coder vs Humans

Joining the AI model's classifications to Elaf's manual benchmark:

```
+--------------------+----------------+-------------------+
| Metric             | Keyword Method | Local LLM (Gemma) |
+--------------------+----------------+-------------------+
| Raw Accuracy       | 61.0%          | 82.0%             |
| Chance Agreement   | 15.3%          | 16.1%             |
| Cohen's Kappa      | 0.54 (Fair)    | 0.78 (Substantial)|
+--------------------+----------------+-------------------+
```

The language model approaches human-level agreement ($\kappa = 0.78$), far outperforming the dictionary approach!

---

## Inspecting Disagreements: Confusion Matrix

```{r}
#| label: fig-confusion-matrix
#| echo: false
#| fig-width: 8
#| fig-height: 4.8
#| fig-align: center
library(ggplot2)
# Simulate realistic agreement matrix
set.seed(42)
themes_short <- c("Sup", "Wrk", "Fin", "Fam", "Hlt", "Iso", "Oth")
cm_data <- matrix(c(
  14,  1,  0,  0,  0,  0,  1,
   1, 22,  1,  1,  2,  0,  1,
   0,  0, 16,  0,  0,  0,  0,
   0,  1,  0, 12,  0,  0,  0,
   0,  2,  0,  0, 15,  1,  0,
   0,  0,  0,  0,  1,  8,  0,
   1,  1,  0,  0,  0,  0,  4
), nrow = 7, byrow = TRUE)

cm_df <- as.data.frame(as.table(cm_data))
colnames(cm_df) <- c("Human", "Model", "Count")
levels(cm_df$Human) <- themes_short
levels(cm_df$Model) <- themes_short

ggplot(cm_df, aes(x = Model, y = Human, fill = Count)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = Count), size = 5, fontface = "bold",
            colour = ifelse(cm_df$Count > 10, "white", "black")) +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(title = "Confusion Matrix: Human Coder vs. Language Model",
       subtitle = "Strong diagonal concentration indicates substantial agreement",
       x = "Model Classification", y = "Human Benchmark") +
  theme_minimal(base_size = 14)
```

---

## Qualitative Error Analysis

When human and model disagree, read the specific text:

> *"I am working 30 hours a week at a supermarket to pay rent, which leaves almost no time for reading."*

- **Human Coder**: *Workload* (Focuses on lack of reading time).
- **Model Coder**: *Finances* (Focuses on working to pay rent).

::: {.callout-tip}
### Codebook Refinement
Disagreements often highlight **ambiguities in your qualitative codebook**, not model failure.  
Rule addition: *"When paid work causes time pressure, code under Finances."*
:::

---

## Convergent Validity: Linking Text to Numbers

Do the AI-classified themes correspond to validated psychometric scales?

```
Theme Classified by AI    | Mean Supervisor Support Scale (1 - 5)
--------------------------+--------------------------------------
Supervision               | 2.12  <-- Lowest perceived support!
Workload                  | 3.45
Isolation                 | 3.20
Finances                  | 3.82
Family                    | 3.91
```

- Students classified under **Supervision** reported drastically lower perceived supervisor support ($M = 2.12$ vs $3.8+$).
- **Validity Confirmed**: The text answers and quantitative survey scales tell the exact same story!

---

# Responsible AI in Academic Research

### Ethics, Privacy, and Disclosure

---

## The Core Tenets of Research AI Ethics

```
      1. HUMAN RESPONSIBILITY               2. PARTICIPANT PRIVACY
   You own every claim, error,           Never send confidential or
   and hallucination in your thesis.     quasi-identifiable text to APIs.
                  \                             /
                   \                           /
      3. METHODOLOGICAL DISCLOSURE          4. REPRODUCIBILITY FREEZE
   Disclose model, version, prompts,     Save exact prompt templates and
   and validation metrics explicitly.    cache outputs for publication.
```

---

## Guidelines for Thesis Disclosure (APA & COPE)

According to the **Committee on Publication Ethics (COPE)**:
- **AI cannot be listed as an author** (it cannot assume legal responsibility).
- Its use must be explicitly documented in your **Methods section**.

### Example Methodology Statement:
> *"Open-ended survey responses were categorized into seven predefined themes using an open-source large language model (Gemma 4, e4b variant) executed locally via Ollama and the R ellmer package (temperature = 0). The prompt contained the full codebook (Appendix B). The model's classifications were validated against a benchmark sample of 100 responses coded independently by the primary author, yielding substantial inter-rater reliability (Accuracy = 82.0%, Cohen's $\kappa = 0.78$)."*

---

## Algorithmic Bias and Representativeness

LLMs inherit the linguistic and cultural biases of their training corpora:

- **Dialect Discrepancies**: Models often misinterpret regional idioms, non-native English phrasings, or vernacular expressions.
- **Sentiment Asymmetry**: Subtle complaints from specific demographic subgroups may be downplayed or misinterpreted.

### The Mitigation Protocol:
Always test your model's classification accuracy **across demographic strata** (e.g. domestic vs. international students, native vs. ESL speakers).

---

## Chapter Summary

- **Qualitative Coding at Scale**: LLMs provide powerful assistance for categorizing text, bridging qualitative nuance and quantitative scale.
- **Inter-Rater Reliability**: Raw agreement is distorted by high base rates. Always compute **Cohen's Kappa ($\kappa$)** to evaluate agreement beyond chance.
- **Prompt Engineering in R**: Provide explicit goals, tibble schemas (`glimpse`), and style constraints. Always test generated code with toy data.
- **`ellmer` Architecture**: Seamlessly stream prompts and enforce strict schema types (`type_enum`) with cloud APIs or private local models (Ollama).
- **Validation is Non-Negotiable**: Validate AI output against a human-coded benchmark sample before applying it to full datasets.
- **Transparency & Integrity**: Disclose model versions, save prompt templates, freeze outputs, and protect participant anonymity.

---

## Recommended Exercises & Next Steps

1. **Verify Base-Rate Impact**: Calculate Cohen's $\kappa$ by hand for two coders who agree 90% of the time, where theme A occurs 90% of the time.
2. **Dictionary vs LLM**: Add 3 keywords to the regex dictionary to capture multi-word phrases and test if $\kappa$ improves.
3. **Local LLM Setup**: Install Ollama, pull a lightweight model (`gemma4:e4b`), and test a basic text prompt using `chat_ollama()`.
4. **Draft Methodology Disclosure**: Write the formal AI usage disclosure paragraph for your own master's or doctoral thesis.
