---
title: "The Complete Thesis Project"
subtitle: "End-to-End Pipeline, Project Architecture, Scientific Chain of Reasoning, and Defense Synthesis"
author: "Dr. Polla Fattah"
institute: "From Data to Thesis: Research Data Analysis with R"
format:
  revealjs:
    theme: [default, ../../styles/theme.scss]
    slide-number: true
    preview-links: auto
    chalkboard: true
    transition: slide
    highlight-style: github
    code-line-numbers: false
    width: 1200
    height: 750
execute:
  freeze: auto
---

# Putting It All Together

### From Messy Raw Survey to the Final Dissertation Defense

**Chapter 19 · From Data to Thesis**

Dr. Polla Fattah

---

## The Supervisor's Final Advice

In the final semester of graduate study:

> *"Before you type a single paragraph of your results chapter, rebuild your entire analysis—from the raw Excel export to the final regression table—as a single, self-contained pipeline that executes seamlessly from start to finish.*  
> 
> *Then you will know that every number in your thesis is true, and you can answer any examiner's challenge by pointing directly to the code."*

---

## A Thesis Is an Argument; Data Is the Evidence

A doctoral or master's thesis convinces examiners when every claim rests on an unbroken chain of reasoning:

```
+-----------------------------------------------------------------------------------+
|                            THE SCIENTIFIC CHAIN                                   |
+-------------+-------------+-------------+-------------+-------------+-------------+
|  Research   | Experimental| Operational | Statistical | Empirical   | Theoretical |
|  Question   |   Design    |  Variables  |    Model    |   Result    | Conclusion  |
+-------------+-------------+-------------+-------------+-------------+-------------+
```

If one link is broken:
- A table copied by hand before outlier cleaning.
- A causal claim made on purely observational data.
- An examiner unable to reproduce the findings.

The argument collapses. This capstone chapter locks every link together.

---

## The Research Project Lifecycle

Every empirical quantitative investigation follows this journey:

```
    Plan & Frame                  Collect & Import                 Clean & Tidy
  [ Ch 5: RQ / Design ] -------> [ Ch 1-2: R / Import ] -------> [ Ch 3: Tidyverse ]
                                                                        |
                                                                        v
   Communicate & Defend           Model & Test                   Explore & Plot
 [ Ch 17-19: Quarto / APA ] <--- [ Ch 7-16: Inference / ML ] <-- [ Ch 4, 6: EDA / Viz ]
```

- In reality, research is an iterative loop: exploratory plotting reveals dirty data, and model diagnostics trigger data transformations.
- **Writing every step in clean R scripts makes iteration painless.**

---

# Professional Project Architecture

### Structuring a Self-Contained Graduate Codebase

---

## The Directory Layout

A reproducible research repository must be immediately navigable:

```
wellbeing-thesis/
├── wellbeing-thesis.Rproj      # RStudio Project Root
├── README.md                   # Purpose, execution instructions, software requirements
├── data-raw/
│   └── wellbeing_raw.xlsx      # Read-only raw export (NEVER touch or edit)
├── data/
│   ├── clean_questionnaire.rds # Generated by 01-clean-data.R
│   └── clean_semesters.rds     # Clean longitudinal panel
├── R/
│   ├── 01-clean-data.R         # Raw survey -> Tidy analytical dataframes
│   ├── 02-descriptive-stats.R  # Table 1 generation & participant demographics
│   └── 03-models.R             # Inferential tests, mixed models, ML
├── output/
│   ├── figure-1-thesis.png     # High-res publication graphics (300 DPI)
│   └── table-regression.docx   # Rendered Word/LaTeX tables
├── results.qmd                 # Complete Quarto results chapter
├── references.bib              # BibTeX citation database
└── renv.lock                   # Frozen package dependencies
```

---

## Four Inviolable Principles of Project Design

::: {.callout-note appearance="simple"}
### 1. Raw Data is Read-Only
Never edit raw data files in Excel. Any correction must be documented as a reproducible line of R code.
:::

::: {.callout-note appearance="simple"}
### 2. Derived Files are Disposable
Everything in `data/` and `output/` can be deleted at any moment. Running the scripts regenerates them identically.
:::

::: {.callout-note appearance="simple"}
### 3. Chronological Script Numbering
Scripts are named with sequential prefixes (`01-...`, `02-...`) indicating the order of execution.
:::

::: {.callout-note appearance="simple"}
### 4. Zero Hard-Coded Machine Paths
Never write `C:/Users/Polla/...`. Use `here::here("data", "file.csv")` so the project runs on any computer.
:::

---

# The Cleaning & Reshaping Pipeline

### Transforming Dirty Raw Surveys into Tidy Analytical Tables

---

## End-to-End Data Cleaning in One Script

```{r}
#| echo: true
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(readxl)
library(data2thesis)

raw <- read_excel(data2thesis_example("wellbeing_raw.xlsx"))

item_names <- c(paste0("stress_", 1:6), paste0("burnout_", 1:6),
                paste0("support_", 1:6), paste0("satisfaction_", 1:4))

responses <- raw |>
  filter(!str_detect(str_to_upper(`Q1_Student ID`), "^TEST")) |>
  select(-`Response ID`) |>
  distinct() |>
  rename(student_id = `Q1_Student ID`, workshop = `Workshop group`,
         considering_dropout = `Y1_Considered leaving?`) |>
  rename_with(~ item_names, .cols = Q11_1:Q11_22) |>
  mutate(across(all_of(item_names), ~ as.integer(na_if(.x, "99"))))
```

---

## Longitudinal Panel Pivoting & Boundary Clamping

```{r}
#| echo: true
semesters_clean <- responses |>
  select(student_id, matches("_S[1-4]$")) |>
  pivot_longer(-student_id, names_to = c(".value", "semester"), names_sep = "_S") |>
  rename(gpa = GPA, sleep_hours = Sleep, study_hours = Study,
         exercise_days = Exercise, caffeine_mg = Caffeine,
         supervisor_meetings = Meetings, wellbeing = Wellbeing) |>
  mutate(
    semester    = as.integer(semester),
    sleep_hours = parse_number(str_replace(sleep_hours, ",", ".")),
    across(c(gpa, study_hours, exercise_days, caffeine_mg, supervisor_meetings, wellbeing),
           parse_number),
    # Biological / Logical Feasibility Boundary Clamping:
    sleep_hours = if_else(sleep_hours > 24, NA, sleep_hours),
    study_hours = if_else(study_hours > 168, NA, study_hours),
    gpa         = if_else(gpa > 4, NA, gpa)
  ) |>
  filter(!if_all(gpa:wellbeing, is.na))
```

---

## Psychometric Scale Scoring

Constructing psychometric composites with reverse-coded item inversion:

```{r}
#| echo: true
scores <- questionnaire |>
  mutate(
    # Invert negatively worded item: 1->5, 2->4, 3->3, 4->2, 5->1
    stress_4 = 6 - stress_4,
    stress   = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
    burnout  = rowMeans(pick(burnout_1:burnout_6), na.rm = TRUE),
    support  = rowMeans(pick(support_1:support_6), na.rm = TRUE)
  ) |>
  select(student_id, stress, burnout, support)

study <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1), join_by(student_id))
```

`study` now contains 800 graduate students with baseline demographic, survey, and academic variables.

---

# Describing the Sample: "Table 1"

### The Indispensable Demographic Baseline

---

## Modular Formatting Helper Functions

Avoid repetitive copy-pasting by writing small, dedicated string formatters:

```{r}
#| echo: true
mean_sd <- function(x) sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
percent <- function(x) sprintf("%.0f%%", 100 * mean(x, na.rm = TRUE))

describe <- function(d) {
  tibble(
    Characteristic = c("Sample Size (N)", "Age, mean (SD)", "Female (%)", "Part-time (%)",
                       "Perceived Stress (1-5)", "Supervisor Support (1-5)",
                       "Baseline Wellbeing (0-100)", "Considering Dropout (%)"),
    Value = c(as.character(nrow(d)), mean_sd(d$age), percent(d$gender == "Female"),
              percent(d$study_mode == "Part-time"), mean_sd(d$stress),
              mean_sd(d$support), mean_sd(d$wellbeing),
              percent(d$considering_dropout == "Yes"))
  )
}
```

---

## Generating Table 1: Master's vs PhD

```{r}
#| echo: true
describe(filter(study, programme == "Master's")) |>
  rename(`Master's` = Value) |>
  left_join(describe(filter(study, programme == "PhD")) |> rename(PhD = Value),
            join_by(Characteristic)) |>
  left_join(describe(study) |> rename(Total = Value), join_by(Characteristic)) |>
  knitr::kable(align = "lrrr")
```

Every number is computed directly from data; zero manual cell typing in Word!

---

# Answering the Research Questions

### Synthesizing the Core Methodological Arsenal

---

## RQ3: Does the Wellbeing Workshop Have a Causal Effect?

```{r}
#| echo: true
library(lme4)

panel <- semesters |>
  left_join(students, join_by(student_id)) |>
  mutate(time = semester - 1)

workshop_model <- lmer(wellbeing ~ factor(semester) * workshop + (time | student_id),
                       data = panel)

gaps <- expand.grid(semester = 1:4, workshop = c("Invited", "Not invited")) |>
  mutate(time = semester - 1)
gaps$wellbeing <- predict(workshop_model, newdata = gaps, re.form = NA)
gaps <- gaps |>
  pivot_wider(id_cols = semester, names_from = workshop, values_from = wellbeing) |>
  mutate(gap = round(Invited - `Not invited`, 1))
gaps
```

---

## Reporting RQ3: Academic Thesis Style

```
                               MODEL RESULTS SUMMARY
+--------------------+---------------------------------------------------------------+
| Semester 1 (Base)  | Difference = 0.4 points (Baseline equivalence confirmed)     |
| Semester 2 (Post)  | Difference = +6.8 points (Statistically significant spike)    |
| Semester 3 (Follow)| Difference = +3.2 points (Intervention effect fading)         |
| Semester 4 (Final) | Difference = +1.1 points (Convergence back to control)        |
+--------------------+---------------------------------------------------------------+
```

### Thesis Reporting Sentence:
> *"A linear mixed-effects model revealed that wellbeing was equivalent across experimental groups prior to intervention ($\Delta = 0.4$ points). Following the workshop, invited students exhibited a significant $6.8$-point advantage in Semester 2 (95% CI [$5.1$, $8.5$]). However, this effect attenuated over time to $3.2$ points in Semester 3 and $1.1$ points in Semester 4, indicating that booster sessions are required for sustained psychological impact."*

---

## RQ5: Explaining Academic Performance (GPA)

Multiple linear regression using `broom::tidy()`:

```{r}
#| echo: true
library(broom)

gpa_model <- lm(gpa ~ sleep_hours + study_hours + stress + support, data = study)

tidy(gpa_model, conf.int = TRUE) |>
  select(term, estimate, std.error, p.value, conf.low, conf.high) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

Total variance explained: $R^2 = 0.174$ ($F(4, 786) = 41.3$, $p < .001$).

---

## Reporting RQ5: Multiple Regression Synthesis

```
Predictor           | Estimate (B) | 95% CI          | Interpretation
--------------------+--------------+-----------------+---------------------------------
Sleep Hours         | +0.076       | [+0.05, +0.10]  | +1 hr sleep = +0.08 GPA boost
Study Hours         | +0.012       | [+0.01, +0.02]  | Modest positive contribution
Stress (1-5)        | -0.114       | [-0.16, -0.07]  | Chronic stress suppresses GPA
Supervisor Support  | +0.021       | [-0.02, +0.06]  | Not statistically significant
```

### Thesis Reporting Sentence:
> *"Multiple regression analysis indicated that sleep duration was significantly and positively associated with first-semester GPA ($B = 0.08, 95\%\text{ CI } [0.05, 0.10], p < .001$), holding study hours, stress, and supervisor support constant. The four predictors together accounted for $17.4\%$ of the variance in student GPA ($R^2 = .174, p < .001$)."*

---

## RQ9: Identifying Students at Risk of Dropout

Binary logistic regression with exponentiated coefficients (Odds Ratios):

```{r}
#| echo: true
dropout_model <- glm(
  I(considering_dropout == "Yes") ~ stress + support + financial_worry + employment + study_mode,
  data = study, family = binomial
)

tidy(dropout_model, conf.int = TRUE, exponentiate = TRUE) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

---

## Reporting RQ9: Odds Ratios & Confidence Intervals

```
Variable             | Odds Ratio (OR) | 95% CI         | Risk Direction
---------------------+-----------------+----------------+-----------------------------
Stress               | 2.45            | [1.98, 3.06]   | +145% odds per stress point
Supervisor Support   | 0.58            | [0.47, 0.71]   | -42% odds (Protective buffer)
Financial Worry      | 1.82            | [1.44, 2.31]   | Significant hazard
Employment (Working) | 1.48            | [1.02, 2.15]   | Competing time constraint
```

### Thesis Reporting Sentence:
> *"Binary logistic regression demonstrated that each one-point increase in perceived stress more than doubled the odds of considering dropout ($\text{OR} = 2.45, 95\%\text{ CI } [1.98, 3.06], p < .001$). Conversely, perceived supervisor support acted as a strong protective factor, reducing dropout odds by $42\%$ per unit increase ($\text{OR} = 0.58, 95\%\text{ CI } [0.47, 0.71], p < .001$)."*

---

# Multi-Panel Publication Visualizations

### Constructing Composite Figures with `patchwork`

---

## Assembling Figure 1: Intervention & Academic Trajectories

```{r}
#| label: fig-thesis-combined
#| echo: false
#| fig-width: 10
#| fig-height: 4.8
#| fig-align: center
library(ggplot2)
library(patchwork)

panel_a <- gaps |>
  pivot_longer(c(Invited, `Not invited`), names_to = "workshop", values_to = "wellbeing") |>
  ggplot(aes(x = semester, y = wellbeing, colour = workshop)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("Invited" = "#1b7837", "Not invited" = "#762a83")) +
  labs(title = "Intervention Trajectory", x = "Semester", y = "Wellbeing (0-100)", colour = "Group") +
  theme_minimal(base_size = 13)

panel_b <- ggplot(study, aes(x = sleep_hours, y = gpa)) +
  geom_point(alpha = 0.35, colour = "#2b5c8f") +
  geom_smooth(method = "lm", formula = y ~ x, colour = "#d95f02", fill = "#fee090", linewidth = 1.1) +
  labs(title = "Sleep & Academic GPA", x = "Sleep Hours / Night", y = "Semester 1 GPA") +
  theme_minimal(base_size = 13)

(panel_a + panel_b) +
  plot_annotation(tag_levels = "A")
```

---

## Saving Publication-Grade Figures

Export directly to disk in exact physical dimensions for journal print standards:

```r
library(ggplot2)

# Vector format (lossless scaling for PDF submission):
ggsave(here::here("output", "figure-1.pdf"),
       plot = combined_plot, width = 18, height = 9, units = "cm")

# High-resolution raster (300 DPI for Word submission):
ggsave(here::here("output", "figure-1.png"),
       plot = combined_plot, width = 18, height = 9, units = "cm", dpi = 300)
```

---

# The Scientific Chain of Reasoning

### Linking Data to Valid Intellectual Claims

---

## Comparing the Three Research Claims

| Step | Workshop Intervention (RQ3) | Explaining GPA (RQ5) | Considering Dropout (RQ9) |
|---|---|---|---|
| **Question** | Does the workshop improve wellbeing? | What factors predict GPA? | Who considers leaving university? |
| **Hypothesis** | Higher wellbeing in invited group | Sleep positively correlates with GPA | Stress increases dropout odds |
| **Design** | **Randomized Field Experiment** | **Cross-Sectional Observational** | **Longitudinal Cohort Survey** |
| **Model** | Linear Mixed-Effects (`lmer`) | Multiple OLS Regression (`lm`) | Logistic Regression (`glm`) |
| **Result** | $+6.8$ pts in S2; fades by S4 | $+0.08$ GPA points / hr of sleep | $\text{OR} = 2.45$ per point of stress |
| **Conclusion** | **Causal Effect**: Workshop works | **Association**: Sleep relates to GPA | **Risk Factor**: Stress elevates risk |
| **Limitation** | Invitation assigned, not attendance | Self-reported sleep; confounders | Intent to leave $\neq$ actual leaving |

---

## The Cardinal Rule of Scientific Claims

::: {.callout-important}
### Significant $p$-values Do NOT Establish Causality!
Statistical significance ($p < .05$) merely confirms that an observed pattern is unlikely under the null hypothesis of pure chance.
:::

- **Only Randomized Designs justify causal verbs** (*"caused"*, *"increased"*, *"led to"*).
- **Observational Designs require cautious associative verbs** (*"associated with"*, *"co-occurred"*, *"predicted"*).
- Every thesis discussion must transparently articulate its **methodological boundary conditions and limitations**.

---

# Choosing the Right Statistical Method

### A Comprehensive Decision Architecture

---

## Method Selection Decision Tree

```
                               WHAT IS YOUR RESEARCH GOAL?
                                            |
         +-------------------+--------------+--------------+-------------------+
         |                   |                             |                   |
      DESCRIBE            EXPLAIN / COMPARE             PREDICT             STRUCTURE / CLUSTER
         |                   |                             |                   |
     Histograms       What is your outcome?           Supervised ML         Unsupervised Learning
     Boxplots                |                        (Ch 11-15)            (Ch 9, 14)
     Medians / SD     +------+------+                 - Elastic Net         - PCA
     (Ch 4, 6)        |             |                 - Random Forest       - K-Means
                  CONTINUOUS      BINARY / CAT        - XGBoost             - GMM / DBSCAN
                      |             |
                 Are measures    Logistic Regression
                   repeated?     Odds Ratios (Ch 8)
                      |
           +----------+----------+
           |                     |
          NO                    YES
     OLS Regression       Mixed-Effects Models
     ANOVA (Ch 7-8)       Random Slopes (Ch 10)
```

---

## Complete Textbook Research Question Map

| RQ | Core Question | Primary Method | Textbook Chapter |
|---|---|---|---|
| **RQ1** | What does graduate student life look like? | Exploratory Visualizations & Descriptives | Ch 4, 6 |
| **RQ2** | Do graduate students sleep less than 7 hours? | One-Sample $t$-test & Confidence Intervals | Ch 7 |
| **RQ3** | Does the workshop improve wellbeing? | Two-Sample $t$-test & Mixed-Effects Model | Ch 7, 10 |
| **RQ4** | Do faculties and study modes differ? | Factorial Two-Way ANOVA & Post-Hocs | Ch 7, 8 |
| **RQ5** | What factors explain academic GPA? | Multiple OLS Linear Regression | Ch 8, 13 |
| **RQ6** | Does the survey measure what it claims? | Exploratory Factor Analysis & Cronbach's $\alpha$ | Ch 9 |
| **RQ7** | Are there distinct student burnout profiles? | $k$-Means Clustering & Mixture Modeling | Ch 9, 14 |
| **RQ8** | How do wellbeing and GPA evolve over time? | Growth-Curve Linear Mixed Models | Ch 10 |
| **RQ9** | Who is at risk of dropping out? | Logistic Regression, Decision Trees, ML | Ch 8, 11, 12 |
| **RQ10** | Can final GPA be accurately predicted? | Ridge, Lasso, and Gradient Boosted Trees | Ch 13 |
| **RQ11** | What challenges do students describe? | Qualitative NLP with LLMs (`ellmer`) | Ch 18 |
| **RQ12** | How many counseling visits will occur? | Time Series Forecasting (ARIMA / ETS) | Ch 16 |

---

# Universal Research Challenges

### Pitfalls Every Doctoral Researcher Encounter

---

## Six Perils of Real-World Research

```
1. DIRTY & MISSING DATA           2. SMALL SAMPLE UNDERPOWERING
   Cleaning takes 80% of time.       Wide CIs and high false-negative rates.
   Never drop NAs blindly.           Conduct prospective power analysis (Ch 5).
          \                                 /
           \                               /
3. MULTIPLE COMPARISON INFLATION   4. REGRESSION DIAGNOSTIC DRIFT
   Running 20 tests = 64% error.     Check heteroskedasticity and normality
   Preregister confirmatory tests.   with visual residual plots (Ch 8).
          /                                 \
         /                                   \
5. CONFOUNDING & SELECTION BIAS    6. OVERFITTING IN PREDICTION
   Correlation is not causation.     Models that memorize train noise fail
   Search for lurking third factors. on holdout test sets. Use cross-validation.
```

---

## Where to Journey Next

Completing this book places you in the top tier of data-fluent researchers.

Specialized methodologies to explore next:

| Advanced Discipline | What It Accomplishes | Key R Packages |
|---|---|---|
| **Structural Equation Modeling (SEM)** | Latent variable measurement and simultaneous path analysis | `lavaan` |
| **Bayesian Inference** | Incorporates prior knowledge; outputs full posterior distributions | `brms`, `rstanarm` |
| **Survival Analysis** | Models time-to-event outcomes (e.g. time until degree completion) | `survival`, `survminer` |
| **Automated Publication Tables** | Generates production-ready APA / medical tables | `gtsummary`, `modelsummary` |
| **Advanced Text Mining** | Sentiment tracking, LDA topic modeling, and tokenization | `tidytext`, `quanteda` |

---

## The Graduate Defense Checklist

Before walking into your thesis defense room:

- [ ] **100% Code-Rebuilt Results**: The entire thesis compiles from raw data in a single run.
- [ ] **Table 1 Verified**: Demographics match official institutional registration rolls.
- [ ] **No Naked $p$-Values**: Every reported test includes an effect size ($R^2$, Cohen's $d$, OR) and 95% CI.
- [ ] **Appropriate Language**: Associative verbs used for observational data; causal verbs reserved for randomized trials.
- [ ] **Package Archive Locked**: `renv.lock` safely frozen and committed to version control.
- [ ] **Limitations Acknowledged**: Boundary conditions clearly articulated in Chapter 5 Discussion.

---

## Capstone Summary

1. **Research is an Unbroken Pipeline**: Raw data $\to$ cleaning scripts $\to$ analysis $\to$ Quarto reporting $\to$ thesis defense.
2. **Reproducibility is Scientific Rigor**: By eliminating manual transcription and using inline R code, your thesis becomes immune to last-minute data adjustments.
3. **Intellectual Honesty Convinces**: Examiners trust researchers who clearly state the limitations of their observational data and defend the validity of their causal inferences.
4. **Mastery of the Toolset**: From basic `dplyr` wrangling to high-dimensional machine learning and deep language models, R empowers you to answer questions that matter.

**Congratulations on completing the journey from Data to Thesis!**
