# Introduction to Machine Learning in R {#sec-ch10}

```{r}
#| label: setup
#| include: false
library(data2thesis)
```

In Chapter 7, Elaf used logistic regression to *explain* who considers dropping out: stress, low support, and money worries all raise the odds. Her supervisor now asks a different question. "If the university knew, at the end of a student's first semester, who is at risk, it could offer help before it is too late. Could you build a model that picks out those students?"

That is a **prediction** question, and it changes what counts as a good model. For explanation, what matters is whether each coefficient is estimated well and makes sense. For prediction, what matters is how accurately the model works on *new* students it has never seen. **Machine learning** is the set of methods and practices for building and fairly testing predictive models. This chapter introduces its core ideas, using the **tidymodels** framework, and builds Elaf's first predictive model.

::: {.callout-tip appearance="simple"}
## By the end of this chapter you will be able to

- Explain supervised and unsupervised learning, and prediction versus explanation.
- Split data into training and test sets, and explain why a model must be tested on data it has not seen.
- Prepare data for modelling with a recipe, and combine a recipe and a model in a workflow.
- Evaluate a classification model, and explain why accuracy alone can mislead.
- Recognise overfitting, and use cross-validation to estimate performance honestly.
- Tune a model's hyperparameters and evaluate the final model on the test set.
:::

## What is machine learning?

**Machine learning** covers methods that learn patterns from data in order to make predictions or find structure. The two main kinds are:

- **Supervised learning**: the data includes the outcome you want to predict, and the model learns the relationship between the predictors and that outcome. If the outcome is a category, such as "considering dropout: yes or no", it is a **classification** problem; if it is a number, such as next semester's GPA, it is a **regression** problem. Chapters 10 to 12 are about supervised learning.
- **Unsupervised learning**: there is no outcome; the model looks for structure in the data itself. The clustering of Chapter 8 is unsupervised, and Chapter 13 returns to it.

Many machine learning methods are familiar statistics. Logistic regression is both a statistical model and one of the most useful classifiers. What changes is the goal, and with it the way models are judged:

| | Explanation (Chapters 6 to 9) | Prediction (Chapters 10 to 15) |
|---|---|---|
| Question | *Why* does something happen? | *What* will happen for a new case? |
| Judged by | Sensible, well-estimated effects | Accuracy on new data |
| Typical models | Simple, interpretable | Anything that predicts well |
| Main danger | Confounding | Overfitting |

: Explanation and prediction {#tbl-explain-predict}

## The tidymodels framework

**tidymodels** is a collection of packages for machine learning, designed like the tidyverse. Each package does one job: **rsample** splits data, **recipes** prepares it, **parsnip** specifies models, **workflows** combines the pieces, **tune** tunes models, and **yardstick** measures performance. Loading `tidymodels` loads them all:

```{r}
library(tidymodels)
tidymodels_prefer()
```

`tidymodels_prefer()` settles a few name clashes between packages in favour of tidymodels. @fig-pipeline shows how the pieces fit together; the rest of the chapter works through each step.

```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "The machine learning workflow in tidymodels."
flowchart LR
  A[All data] --> B[Split:<br/>training and test]
  B --> C[Training data]
  B --> T[Test data,<br/>set aside]
  C --> D[Recipe:<br/>prepare the data]
  D --> E[Model:<br/>fit and tune with<br/>cross-validation]
  E --> F[Final model]
  T --> G[Evaluate once<br/>on the test data]
  F --> G
```

## Data for prediction

A predictive model can only use information that would be available when the prediction is made. The university wants to act at the end of the first semester, so Elaf's model may use the students' background, their questionnaire scores from the start of the year, and their first-semester records, but nothing from later. The questionnaire scores are calculated as before:

```{r}
scores <- questionnaire |>
  mutate(
    stress_4     = 6 - stress_4,
    stress       = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
    burnout      = rowMeans(pick(burnout_1:burnout_6), na.rm = TRUE),
    support      = rowMeans(pick(support_1:support_6), na.rm = TRUE),
    satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)
  ) |>
  select(student_id, stress, burnout, support, satisfaction)

dropout_data <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1) |> select(-semester), join_by(student_id)) |>
  mutate(considering_dropout = factor(considering_dropout, levels = c("Yes", "No"))) |>
  select(-student_id, -supervisor_id, -workshop, -workshop_sessions)

dim(dropout_data)
```

Two details matter. The outcome must be a factor, and its first level is treated as the event to predict, so the levels are set to put `"Yes"` first. And the workshop variables are removed: the workshop took place after the first semester, so it would not be known when the prediction is made.

```{r}
dropout_data |> count(considering_dropout) |> mutate(share = n / sum(n))
```

Only about `r round(100 * mean(dropout_data$considering_dropout == "Yes"))`% of students answer "Yes". Such an **imbalanced** outcome is common in real prediction problems, from rare diseases to fraud, and, as you will see, it catches out anyone who judges a model by accuracy alone.

## Splitting the data

The most important rule of machine learning: **never judge a model on the data it learned from.** A model can learn the peculiarities of its data, including pure noise, and then look far better than it will be on new cases. The only honest test is data the model has never seen.

So the first step is to split the data. The **training set** is used to build and tune the model. The **test set** is locked away and used only once, at the very end, to estimate how the final model will perform on new students. `initial_split()` does the split; `strata` makes sure that both sets have the same share of "Yes" answers:

```{r}
set.seed(2026)
dropout_split <- initial_split(dropout_data, prop = 0.75, strata = considering_dropout)
dropout_train <- training(dropout_split)
dropout_test  <- testing(dropout_split)

nrow(dropout_train)
nrow(dropout_test)
```

Three quarters of the students are used for training and a quarter are set aside for the final test.

## Preparing the data with a recipe

Most models need the data prepared first: missing values filled in, categories turned into numbers, and numeric predictors put on a common scale. A **recipe** lists these steps:

```{r}
dropout_recipe <- recipe(considering_dropout ~ ., data = dropout_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

Reading it line by line:

- `recipe(considering_dropout ~ ., ...)` says that `considering_dropout` is the outcome and every other column (`.`) is a predictor.
- `step_impute_median()` replaces each missing numeric value with the median of that variable.
- `step_dummy()` turns each categorical predictor into **dummy variables**: 0/1 columns, one per category except a reference category, as `lm()` did automatically in Chapter 7.
- `step_normalize()` turns every numeric predictor into z-scores, so they are on the same scale. Some models, such as k-nearest neighbours below, depend on distances and need this.

The recipe only describes the steps. To see what it produces, `prep()` estimates what it needs from the training data (the medians, the means and standard deviations), and `bake()` applies it:

```{r}
dropout_recipe |>
  prep() |>
  bake(new_data = NULL) |>
  glimpse()
```

`new_data = NULL` means "the training data the recipe was prepared on". Notice that `gender` has become `gender_Male`, `faculty` has become four dummy columns, and all the numbers are now z-scores.

::: {.callout-important}
## No peeking at the test data

The medians, means, and standard deviations are estimated from the **training** data only, and then applied unchanged to the test data. If they were calculated from all the data, information from the test set would leak into the model, and the test would no longer be a fair one. Recipes and workflows handle this automatically, which is one of the best reasons to use them.
:::

## Specifying and fitting a model

**parsnip** specifies models in one consistent way, whatever package does the actual work. Here is logistic regression:

```{r}
logistic_spec <- logistic_reg()
logistic_spec
```

A **workflow** combines the recipe and the model, so they always travel together:

```{r}
logistic_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(logistic_spec)

logistic_fit <- fit(logistic_wf, data = dropout_train)
```

`fit()` prepares the recipe on the training data and fits the model to the prepared data, in one step.

## Evaluating a classifier

How well does the model do on the test students? `augment()` adds the model's predictions to the test data: the predicted class (`.pred_class`) and the predicted probability of each class (`.pred_Yes`, `.pred_No`):

```{r}
test_results <- augment(logistic_fit, new_data = dropout_test)
test_results |>
  select(considering_dropout, .pred_class, .pred_Yes) |>
  head()
```

The simplest measure is **accuracy**: the share of students classified correctly. **yardstick** calculates it:

```{r}
test_results |> accuracy(truth = considering_dropout, estimate = .pred_class)
```

```{r}
#| include: false
acc <- test_results |> accuracy(considering_dropout, .pred_class) |> pull(.estimate)
always_no <- mean(dropout_test$considering_dropout == "No")
```

`r round(100 * acc)`% sounds good. But consider a "model" that ignores every predictor and says "No" for every student. Because most students answer "No", it is right `r round(100 * always_no)`% of the time: *more* often than Elaf's model, while finding not one of the students at risk. For an imbalanced outcome, accuracy is a misleading measure.

A better measure is the **ROC AUC** (the area under the ROC curve), which asks: if you pick one student who considered dropping out and one who did not, how often does the model give the first a higher predicted probability than the second? An AUC of 0.5 is no better than a coin toss, and 1.0 is perfect. Unlike accuracy, it is not fooled by an imbalanced outcome. It uses the predicted probabilities:

```{r}
test_results |> roc_auc(truth = considering_dropout, .pred_Yes)
```

```{r}
#| include: false
auc_test <- test_results |> roc_auc(considering_dropout, .pred_Yes) |> pull(.estimate)
```

An AUC of `r round(auc_test, 2)` means the model ranks an at-risk student above a not-at-risk student about `r round(100 * auc_test)`% of the time: a useful model. Chapter 11 looks at classification measures in much more detail, including how to choose the threshold at which a student is flagged.

## Overfitting

Why all this care about separate data? Because models can **overfit**: learn the training data so closely, noise and all, that they predict new data badly. A dramatic example is **k-nearest neighbours** (k-NN) with $k = 1$. It predicts each student's outcome by finding the single most similar student in the training data and copying their answer. On the training data, every student's nearest neighbour is themselves, so it can never be wrong:

```{r}
knn1_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(nearest_neighbor(neighbors = 1) |> set_mode("classification"))
knn1_fit <- fit(knn1_wf, data = dropout_train)

augment(knn1_fit, new_data = dropout_train) |> roc_auc(considering_dropout, .pred_Yes)
augment(knn1_fit, new_data = dropout_test) |> roc_auc(considering_dropout, .pred_Yes)
```

A perfect score on the training data, and a much poorer one on new students. The training score was pure illusion: the model had memorised the training students, not learned anything that carries over. This is why a model must never be judged on the data it was trained on, and why every model in this book is tested on data it has not seen.

## Cross-validation

The test set is used only once, at the end. But during model building, you often need to compare models or settings, and you cannot use the training data to judge them (that is overfitting again) or the test set (that would use it up). **Cross-validation** solves this by reusing the training data cleverly:

1. Split the training data into, say, 10 equal parts, called **folds**.
2. Fit the model on 9 folds, and measure its performance on the fold left out.
3. Repeat 10 times, leaving out each fold in turn.
4. Average the 10 performance measures.

Every student is used for evaluation exactly once, always by a model that did not see them during fitting. @fig-folds shows the idea with 5 folds.

```{r}
#| label: fig-folds
#| fig-cap: "Five-fold cross-validation: in each round, the model is fitted on four folds (grey) and evaluated on the fifth (dark)."
#| fig-alt: "A grid of five rows, one per round, each split into five blocks. In each row, a different block is dark, marking the fold used for evaluation; the other four are grey."
#| fig-width: 6.5
#| fig-height: 2.8
#| echo: false
expand.grid(fold = 1:5, round = 1:5) |>
  mutate(role = if_else(fold == round, "Evaluate", "Fit")) |>
  ggplot(aes(x = fold, y = factor(round, levels = 5:1), fill = role)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  scale_fill_manual(values = c(Evaluate = "#1f6f8b", Fit = "grey80")) +
  labs(x = "Fold", y = "Round", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

`vfold_cv()` creates the folds, and `fit_resamples()` fits and evaluates the workflow on each:

```{r}
set.seed(2026)
dropout_folds <- vfold_cv(dropout_train, v = 10, strata = considering_dropout)

logistic_cv <- fit_resamples(logistic_wf, resamples = dropout_folds,
                             metrics = metric_set(roc_auc, accuracy))
collect_metrics(logistic_cv)
```

The cross-validated AUC, about `r round(collect_metrics(logistic_cv) |> filter(.metric == "roc_auc") |> pull(mean), 2)`, is an honest estimate of how the model will perform on new students, obtained without touching the test set. The standard error (`std_err`) shows how much the estimate varies across folds.

## Tuning a model

Many models have settings that are not learned from the data but must be chosen beforehand, such as $k$ in k-nearest neighbours. They are called **hyperparameters**, to distinguish them from the **parameters**, such as regression coefficients, that the model learns. The best value is found by **tuning**: trying several values and comparing them with cross-validation.

A **decision tree** (explained fully in Chapter 11) predicts by asking a series of yes-or-no questions, such as "is stress above 3.5?". Its depth, the number of questions in a row, is a hyperparameter. A shallow tree is too simple to capture the patterns (**underfitting**); a very deep tree can fit the training data closely, noise included (**overfitting**). In the model specification, `tune()` marks the hyperparameter to be tuned:

```{r}
tree_spec <- decision_tree(tree_depth = tune(), min_n = 10, cost_complexity = 0) |>
  set_mode("classification")

tree_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(tree_spec)
```

`tune_grid()` tries each value in a grid, using the same cross-validation folds:

```{r}
set.seed(2026)
tree_tuning <- tune_grid(tree_wf, resamples = dropout_folds,
                         grid = tibble(tree_depth = 1:10),
                         metrics = metric_set(roc_auc))
```

```{r}
#| label: fig-tuning
#| fig-cap: "Cross-validated ROC AUC of decision trees of different depths."
#| fig-alt: "Line chart of AUC against tree depth from 1 to 10. The AUC rises steeply from 0.5 at depth 1 to a peak around depth 4, then levels off and dips slightly for deeper trees."
#| fig-width: 6
#| fig-height: 3.6
autoplot(tree_tuning) + theme_minimal(base_size = 12)
```

The pattern is the classic one: a tree of depth 1 is no better than chance, the AUC rises as the tree grows, and beyond a depth of about 4 it levels off and dips slightly as the deeper trees start to overfit. `select_best()` picks the best depth:

```{r}
best_depth <- select_best(tree_tuning, metric = "roc_auc")
best_depth
```

## The final test

Only now, with the model chosen, is the test set used. `finalize_workflow()` plugs the best depth into the workflow, and `last_fit()` fits it on the whole training set and evaluates it once on the test set:

```{r}
final_tree <- tree_wf |>
  finalize_workflow(best_depth) |>
  last_fit(dropout_split, metrics = metric_set(roc_auc, accuracy))
collect_metrics(final_tree)
```

And for comparison, the same for logistic regression:

```{r}
final_logistic <- last_fit(logistic_wf, dropout_split, metrics = metric_set(roc_auc, accuracy))
collect_metrics(final_logistic)
```

```{r}
#| include: false
auc_tree <- collect_metrics(final_tree) |> filter(.metric == "roc_auc") |> pull(.estimate)
auc_lr <- collect_metrics(final_logistic) |> filter(.metric == "roc_auc") |> pull(.estimate)
```

The tuned tree reaches an AUC of `r round(auc_tree, 2)`; plain logistic regression, `r round(auc_lr, 2)`. The more flexible model is not the better one. For Elaf's data, where the risk of dropout rises steadily with stress and falls steadily with support, a simple smooth model captures the pattern well, and a tree, which splits the data into boxes, captures it less well. Trying a simple model first, and making a complex one earn its place, is one of the most useful habits in machine learning.

::: {.callout-warning}
## Predictions about people

A model that flags students "at risk" affects real people. Before such a model is used, ask: What happens to a student who is flagged by mistake, or missed? Does the model work equally well for every group of students, for example women and men, or part-time and full-time students? Is it used to offer help, or to judge? A model with a good AUC can still be unfair to a particular group, and in Elaf's case, the right use is to offer support earlier, never to penalise.
:::

::: {.callout-note icon="false"}
## In your field: business and finance

Banks use exactly these methods to predict which loan applicants will repay. The `credit_data` dataset from the **modeldata** package (installed with tidymodels) records real loan applications, with whether each loan turned out well or badly. The same workflow applies:

```{r}
data(credit_data, package = "modeldata")

set.seed(1)
credit_split <- initial_split(credit_data, strata = Status)

credit_wf <- workflow() |>
  add_recipe(
    recipe(Status ~ ., data = training(credit_split)) |>
      step_impute_median(all_numeric_predictors()) |>
      step_impute_mode(all_nominal_predictors()) |>
      step_dummy(all_nominal_predictors())
  ) |>
  add_model(logistic_reg())

last_fit(credit_wf, credit_split) |> collect_metrics()
```

`step_impute_mode()` fills missing categories with the most common one. Only the data and the outcome have changed; every step is the same. (The output also shows the **Brier score**, `brier_class`, another measure of how good the predicted probabilities are: 0 is perfect, and lower is better.)
:::

## Chapter review

### Summary

- Supervised learning predicts an outcome (classification for categories, regression for numbers); unsupervised learning finds structure. Prediction is judged by performance on new data.
- tidymodels splits data (rsample), prepares it (recipes), specifies models (parsnip), combines them (workflows), tunes them (tune), and measures performance (yardstick).
- Split the data into training and test sets, stratified by the outcome, and use the test set only once, at the end.
- A recipe prepares data (imputation, dummy variables, normalisation) using the training data only, so no information leaks from the test set.
- Accuracy misleads for imbalanced outcomes; the ROC AUC is a better summary.
- Overfitting means learning the noise in the training data. Cross-validation estimates performance honestly without using the test set.
- Hyperparameters are chosen by tuning with cross-validation. Simple models often predict as well as complex ones; try them first.
- Predictions about people raise questions of fairness and use.

### Key terms

Machine learning, supervised learning, unsupervised learning, classification, regression, prediction, explanation, imbalanced outcome, training set, test set, stratified split, recipe, imputation, dummy variable, normalisation, data leakage, model specification, workflow, accuracy, ROC AUC, overfitting, underfitting, k-nearest neighbours, cross-validation, fold, hyperparameter, parameter, tuning, decision tree.

## Exercises

The [playground](../playground/chapter10.html) has these and more, with hints and solutions.

1. Remove the questionnaire scores (`stress`, `burnout`, `support`, `satisfaction`) from `dropout_data` and fit the logistic regression workflow again. How much does the cross-validated AUC fall? What does that say about the questionnaire?
2. Split the data 80/20 instead of 75/25 with a different seed. Does the test AUC of the logistic model change much? Why might it?
3. Tune k-nearest neighbours over `neighbors = c(5, 11, 21, 41)` with cross-validation. Which value is best, and how does its AUC compare with logistic regression?
4. Add `step_zv(all_predictors())` to the recipe (it removes predictors with only one value). Look up what "zero variance" means, and explain when this step is useful.
5. In two or three sentences, explain to a university administrator why a model with 85% accuracy might still be useless for finding students at risk.

## Further reading

- *Tidy Modeling with R* [@kuhn2022], the book by the authors of tidymodels, is the definitive guide to the framework.
- *An Introduction to Statistical Learning* [@james2021islr] explains the ideas behind overfitting, cross-validation, and the methods of the next chapters.
