# Classification Models {#sec-ch11}

```{r}
#| label: setup
#| include: false
library(data2thesis)
```

Elaf's first predictive model, a logistic regression, ranks students by their risk of considering dropout quite well. When she shows it to the student counselling service, two questions come back at once. "Aren't there more powerful methods? We hear about random forests and support vector machines." And, more practically: "If we contact every student your model flags, how many of them are really at risk? And how many students at risk will we miss?"

This chapter answers both. It introduces four widely used families of classification models, decision trees, random forests, k-nearest neighbours, and support vector machines, and compares them fairly on Elaf's data. Then it looks beyond the AUC at what happens when a model is actually used to make decisions: the confusion matrix, precision and recall, choosing the threshold at which a student is flagged, and what to do when one outcome is rare.

::: {.callout-tip appearance="simple"}
## By the end of this chapter you will be able to

- Explain how decision trees, random forests, k-nearest neighbours, and support vector machines classify.
- Fit each of them with tidymodels, and compare several models fairly with cross-validation.
- Read a confusion matrix, and calculate sensitivity (recall), specificity, precision, and the F1 score.
- Choose a classification threshold that suits the purpose of the model.
- Explain the options for an imbalanced outcome, and what resampling does and does not change.
:::

## Setting up

The data, split, recipe, and folds are exactly those of Chapter 10, so every model in this chapter is trained and tested on the same students:

```{r}
library(tidymodels)
tidymodels_prefer()

scores <- questionnaire |>
  mutate(
    stress_4     = 6 - stress_4,
    stress       = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
    burnout      = rowMeans(pick(burnout_1:burnout_6), na.rm = TRUE),
    support      = rowMeans(pick(support_1:support_6), na.rm = TRUE),
    satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)
  ) |>
  select(student_id, stress, burnout, support, satisfaction)

dropout_data <- students |>
  left_join(scores, join_by(student_id)) |>
  left_join(semesters |> filter(semester == 1) |> select(-semester), join_by(student_id)) |>
  mutate(considering_dropout = factor(considering_dropout, levels = c("Yes", "No"))) |>
  select(-student_id, -supervisor_id, -workshop, -workshop_sessions)

set.seed(2026)
dropout_split <- initial_split(dropout_data, prop = 0.75, strata = considering_dropout)
dropout_train <- training(dropout_split)
dropout_test  <- testing(dropout_split)

dropout_recipe <- recipe(considering_dropout ~ ., data = dropout_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

set.seed(2026)
dropout_folds <- vfold_cv(dropout_train, v = 10, strata = considering_dropout)
```

## Decision trees

A **decision tree** classifies by asking a series of yes-or-no questions, like a flowchart. To see how a tree chooses its questions, take twelve imaginary students:

```{r}
tiny <- tibble(
  stress  = c(1.5, 2.0, 2.2, 2.8, 3.0, 3.1, 3.4, 3.6, 3.8, 4.0, 4.2, 4.5),
  support = c(4.5, 3.0, 4.0, 2.0, 4.2, 3.8, 2.1, 4.4, 1.8, 2.5, 3.9, 1.5),
  dropout = factor(c("No", "No", "No", "No", "No", "No",
                     "Yes", "No", "Yes", "Yes", "No", "Yes"))
)
```

Four of the twelve consider dropping out. The tree looks for the single question that best separates the "Yes" students from the "No" students. "Is support below 2.75?" puts seven students, all "No", on one side, and five students, four "Yes" and one "No", on the other. The one "No" student among the low-support group has the lowest stress, so a second question, "Is stress 3.1 or more?", separates the rest perfectly:

```{r}
library(rpart)
tiny_tree <- rpart(dropout ~ stress + support, data = tiny, method = "class",
                   control = rpart.control(minsplit = 2, cp = 0))
tiny_tree
```

Each line is a **node**: the question that led to it, the number of students, the number misclassified, and the predicted class. The final nodes, marked `*`, are the **leaves**. To classify a new student, you follow the questions from the top until you reach a leaf. To choose each question, the tree tries every predictor and every possible cut-off and picks the one that makes the two groups most "pure", usually measured by the **Gini impurity** (0 when a group contains only one class).

Decision trees have attractive properties: they are easy to explain, they need no dummy variables or normalisation (a question such as "is stress 3.1 or more?" works the same on any scale), and they capture interactions automatically (stress matters here only for low-support students). For Elaf's data, a recipe that only fills in missing values is enough:

```{r}
tree_recipe <- recipe(considering_dropout ~ ., data = dropout_train) |>
  step_impute_median(all_numeric_predictors())

tree_wf <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(decision_tree(tree_depth = 3, min_n = 10) |> set_mode("classification"))

tree_fit <- fit(tree_wf, data = dropout_train)
```

A depth of 3 keeps the tree small enough to read. @fig-tree draws it with `plot()` and `text()` from rpart; the **rpart.plot** package draws nicer trees if you install it.

```{r}
#| label: fig-tree
#| fig-cap: "A decision tree of depth 3 for considering dropout. At each split, students who meet the condition go left."
#| fig-alt: "A tree diagram. The top split is on support. Further splits on stress and other scale scores lead to leaves labelled Yes or No, each with the number of Yes and No students in it."
#| fig-width: 8
#| fig-height: 5
tree_engine <- extract_fit_engine(tree_fit)
plot(tree_engine, uniform = TRUE, margin = 0.08)
text(tree_engine, use.n = TRUE, cex = 0.75)
```

```{r}
#| include: false
top_split <- as.character(tree_engine$frame$var[1])
```

The first question is about `r top_split`, the strongest single predictor. The numbers under each leaf are the training students in it (Yes/No). The tree is easy to read, but a single tree has a serious weakness: it is unstable. A slightly different sample of students can produce a completely different tree, and, as Chapter 10 showed, a deep tree overfits. The next method turns this weakness into a strength.

## Random forests

A **random forest** grows hundreds of trees, each on a slightly different version of the data, and lets them vote. Two sources of randomness make the trees different:

1. Each tree is grown on a **bootstrap sample** of the training data (Chapter 6): students drawn at random with replacement.
2. At each split, the tree may only choose from a random subset of the predictors. The size of this subset, `mtry`, is a hyperparameter.

Each individual tree overfits in its own way, but because the trees differ, many of their errors cancel out when the trees vote. The predicted probability of "Yes" is the share of trees that vote "Yes". Averaging many models in this way is called an **ensemble**; a random forest is an ensemble of trees [@breiman2001].

The **ranger** package fits random forests quickly. `importance = "permutation"` asks it to also measure how much each predictor matters:

```{r}
forest_spec <- rand_forest(trees = 500) |>
  set_engine("ranger", importance = "permutation") |>
  set_mode("classification")

forest_wf <- workflow() |>
  add_recipe(tree_recipe) |>
  add_model(forest_spec)

set.seed(2026)
forest_fit <- fit(forest_wf, data = dropout_train)
```

A forest of 500 trees cannot be drawn, so it is a "black box". **Permutation importance** opens it a little: for each predictor in turn, the values are shuffled randomly, and the drop in the model's accuracy is recorded. Shuffling an important predictor hurts the predictions; shuffling an unimportant one does not. @fig-importance shows the result.

```{r}
#| label: fig-importance
#| fig-cap: "Permutation importance of each predictor in the random forest."
#| fig-alt: "Horizontal bar chart of predictor importance. Support and stress have the longest bars, followed by burnout; the demographic variables have bars near zero."
#| fig-width: 6.5
#| fig-height: 4.5
importance <- extract_fit_engine(forest_fit)$variable.importance

tibble(predictor = names(importance), importance = importance) |>
  ggplot(aes(x = importance, y = reorder(predictor, importance))) +
  geom_col(fill = "#1f6f8b") +
  labs(x = "Permutation importance (drop in accuracy)", y = NULL) +
  theme_minimal(base_size = 12)
```

The forest relies most on support, stress, and burnout, and hardly at all on the students' background, which agrees with what Elaf found with logistic regression in Chapter 7. Importance says which predictors the model *uses*, not what *causes* dropout; everything Chapter 7 said about confounding still applies.

## k-nearest neighbours

**k-nearest neighbours** (k-NN), met briefly in Chapter 10, classifies a new student by finding the $k$ most similar students in the training data and letting them vote. "Similar" means close together when the predictors are drawn as a map. Suppose a new student in the tiny example has a stress score of 3.5 and a support score of 3.0. The **distance** to each training student is the straight-line distance on the stress-support map:

```{r}
tiny |>
  mutate(distance = sqrt((stress - 3.5)^2 + (support - 3.0)^2)) |>
  arrange(distance) |>
  head(3)
```

With $k = 3$, two of the three nearest students considered dropping out, so the predicted probability of "Yes" is 2/3. @fig-knn shows the picture.

```{r}
#| label: fig-knn
#| fig-cap: "k-nearest neighbours with k = 3. The new student (cross) is classified by the three closest students (circled)."
#| fig-alt: "Scatter plot of support against stress for twelve students, coloured Yes or No. A cross marks a new student at stress 3.5 and support 3. A circle around the cross encloses three students: two Yes and one No."
#| fig-width: 5.5
#| fig-height: 4.2
#| echo: false
circle <- tibble(angle = seq(0, 2 * pi, length.out = 100),
                 stress = 3.5 + 0.91 * cos(angle), support = 3.0 + 0.91 * sin(angle))
ggplot(tiny, aes(x = stress, y = support)) +
  geom_path(data = circle, linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = dropout), size = 3.5) +
  annotate("point", x = 3.5, y = 3.0, shape = 4, size = 5, stroke = 1.5) +
  scale_colour_manual(values = c(No = "#1f6f8b", Yes = "#e07b39")) +
  coord_equal() +
  labs(x = "Stress", y = "Support", colour = "Considering\ndropout") +
  theme_minimal(base_size = 12)
```

Because k-NN is based on distances, the predictors must be on the same scale. Otherwise a variable measured in large numbers, such as caffeine in milligrams, would dominate the distance, and a variable measured on a 1-5 scale would hardly count. That is why the recipe normalises every predictor.

The number of neighbours, $k$, is a hyperparameter. With $k = 1$, the model copies the nearest student and overfits badly (Chapter 10); with a large $k$, it averages over many students and becomes smoother. Tuning shows which works best for Elaf's data:

```{r}
knn_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(nearest_neighbor(neighbors = tune()) |> set_mode("classification"))

set.seed(2026)
knn_tuning <- tune_grid(knn_wf, resamples = dropout_folds,
                        grid = tibble(neighbors = c(1, 5, 11, 21, 41, 81)),
                        metrics = metric_set(roc_auc))
collect_metrics(knn_tuning) |> select(neighbors, mean, std_err)
```

The AUC keeps rising up to the largest $k$ tried. When the best value lies at the edge of the grid, the usual advice is to extend the grid, but here the result is informative in itself: the more neighbours k-NN averages over, the smoother its predictions become, and the better it does. Elaf's data has a smooth pattern (risk rises steadily with stress and falls steadily with support), which a smooth model such as logistic regression captures directly. k-NN is at its best when the pattern is irregular and there is a lot of data.

## Support vector machines

A **support vector machine** (SVM) separates the two classes with a boundary that is as far as possible from the students on either side. Think of drawing a line between the "Yes" and "No" points on the stress-support map: of all the lines that separate them, the SVM picks the one with the widest empty "street" around it, the **margin**. Only the students closest to the boundary, the **support vectors**, determine where it goes. When the classes overlap, as they always do in real data, some students are allowed inside the margin or on the wrong side, at a cost set by a hyperparameter.

A straight boundary is a **linear SVM**. With a **kernel**, the SVM can draw curved boundaries: the popular **radial basis function** (RBF) kernel lets the boundary bend around groups of points. Both are available in tidymodels through the **kernlab** package:

```{r}
svm_linear_spec <- svm_linear() |>
  set_engine("kernlab") |>
  set_mode("classification")

svm_rbf_spec <- svm_rbf() |>
  set_mode("classification")
```

Like k-NN, SVMs are based on distances, so they need normalised predictors. @fig-boundaries shows how differently the four families draw their boundaries, using just stress and support so the result can be drawn as a map.

```{r}
#| label: fig-boundaries
#| fig-cap: "Predicted probability of considering dropout from stress and support, according to four models fitted to Elaf's training data. Darker orange means higher risk; the points are the training students."
#| fig-alt: "Four panels, each a map of stress (x) against support (y), shaded from blue (low risk) to orange (high risk). Logistic regression gives a smooth diagonal gradient; the decision tree gives three rectangular blocks; k-nearest neighbours gives a patchy, irregular surface; the RBF support vector machine gives smooth, rounded regions. In all four, risk is highest at high stress and low support."
#| fig-width: 8
#| fig-height: 6.5
#| echo: false
two_recipe <- recipe(considering_dropout ~ stress + support, data = dropout_train) |>
  step_normalize(all_numeric_predictors())
two_models <- list(
  "Logistic regression" = logistic_reg(),
  "Decision tree (depth 4)" = decision_tree(tree_depth = 4, min_n = 10) |> set_mode("classification"),
  "k-nearest neighbours (k = 21)" = nearest_neighbor(neighbors = 21) |> set_mode("classification"),
  "SVM (RBF kernel)" = svm_rbf() |> set_mode("classification")
)
grid <- expand_grid(stress = seq(1, 5, length.out = 80), support = seq(1, 5, length.out = 80))
set.seed(2026)
surfaces <- purrr::imap(two_models, \(spec, name) {
  fitted <- fit(workflow(two_recipe, spec), data = dropout_train)
  predict(fitted, grid, type = "prob") |> bind_cols(grid) |> mutate(model = name)
}) |> list_rbind() |>
  mutate(model = factor(model, levels = names(two_models)))
ggplot(surfaces, aes(x = stress, y = support)) +
  geom_raster(aes(fill = .pred_Yes)) +
  geom_point(data = dropout_train, aes(shape = considering_dropout), size = 1, alpha = 0.6) +
  scale_fill_gradient(low = "#dbe9f0", high = "#e07b39", limits = c(0, 1)) +
  scale_shape_manual(values = c(Yes = 17, No = 1)) +
  facet_wrap(~ model) +
  coord_equal() +
  labs(x = "Stress", y = "Support", fill = "P(Yes)", shape = "Considering\ndropout") +
  theme_minimal(base_size = 11)
```

## Comparing models fairly

Which model is best? The fair way to compare them is cross-validation on the same folds, and the **workflowsets** package (part of tidymodels) does it for several models at once. `workflow_set()` combines each recipe with each model, and `workflow_map()` runs cross-validation for every combination:

```{r}
dropout_models <- workflow_set(
  preproc = list(tree_data = tree_recipe),
  models  = list(
    tree   = decision_tree(tree_depth = 4, min_n = 10) |> set_mode("classification"),
    forest = rand_forest(trees = 500) |> set_engine("ranger") |> set_mode("classification")
  )
) |>
  bind_rows(workflow_set(
    preproc = list(normalised = dropout_recipe),
    models  = list(
      logistic   = logistic_reg(),
      knn        = nearest_neighbor(neighbors = 41) |> set_mode("classification"),
      svm_linear = svm_linear_spec,
      svm_rbf    = svm_rbf_spec
    )
  ))

dropout_comparison <- workflow_map(dropout_models, "fit_resamples",
                                   resamples = dropout_folds,
                                   metrics = metric_set(roc_auc), seed = 2026)

rank_results(dropout_comparison, rank_metric = "roc_auc") |>
  select(wflow_id, mean, std_err)
```

The trees and forest use the simple recipe, and the others the normalised one. `bind_rows()` joins the two sets of workflows into one.

```{r}
#| include: false
ranked <- rank_results(dropout_comparison, rank_metric = "roc_auc")
best_id <- ranked$wflow_id[1]
auc_of <- function(id) sprintf("%.2f", ranked$mean[ranked$wflow_id == id])
```

The winner, with a cross-validated AUC of `r auc_of("normalised_logistic")`, is plain logistic regression. The linear SVM (`r auc_of("normalised_svm_linear")`), which also draws a straight boundary, comes close, followed by the random forest (`r auc_of("tree_data_forest")`). The standard errors, around 0.03, show that the smaller differences at the top could be due to chance, but none of the flexible models beats the simple one. This is not a failure of the methods: it tells Elaf something about her data. The risk of considering dropout rises smoothly with stress and falls smoothly with support, with no sharp thresholds or complicated interactions for a tree or a forest to find. On data with such structure, and with larger samples, forests and SVMs often do win. The lesson is the one from Chapter 10: compare, and let the simpler model stand unless a complex one clearly does better.

Elaf keeps logistic regression. It predicts as well as any model here, and it can be explained to the counselling service in one sentence.

## The confusion matrix

The AUC measures how well a model *ranks* students. But the counselling service needs a *decision*: contact this student or not. A model makes a decision by comparing each predicted probability with a **threshold**; by default, a student is classified "Yes" when the probability of "Yes" is above 0.5. The **confusion matrix** counts how those decisions turn out.

Take a tiny example first. A screening questionnaire is given to 100 students, 10 of whom are truly at risk. It flags 12 students, 8 of whom are among the 10 at risk:

| | Truly at risk | Not at risk | Total |
|---|---|---|---|
| **Flagged** | 8 (true positives) | 4 (false positives) | 12 |
| **Not flagged** | 2 (false negatives) | 86 (true negatives) | 88 |
| **Total** | 10 | 90 | 100 |

: The four outcomes of a classification {#tbl-confusion}

Every classification ends in one of four cells. A **true positive** is a student at risk who is flagged; a **false negative** is a student at risk who is missed; a **false positive** is a student flagged by mistake; a **true negative** is a student correctly left alone. From the four counts come the key measures:

- **Sensitivity**, also called **recall**: the share of students at risk who are flagged, 8/10 = 80%. *How many at-risk students do we find?*
- **Specificity**: the share of students not at risk who are left alone, 86/90 = 96%.
- **Precision**: the share of flagged students who are truly at risk, 8/12 = 67%. *When we contact a student, how often is it justified?*
- **F1 score**: a single number that balances precision and recall (their harmonic mean), here 0.73. It is high only when both are high.
- **Accuracy**: the share classified correctly, (8 + 86)/100 = 94%. As Chapter 10 showed, it is dominated by the large "No" group.

Now Elaf's model. `conf_mat()` produces the confusion matrix for the test students:

```{r}
logistic_wf <- workflow() |>
  add_recipe(dropout_recipe) |>
  add_model(logistic_reg())
logistic_fit <- fit(logistic_wf, data = dropout_train)

test_results <- augment(logistic_fit, new_data = dropout_test)
test_results |> conf_mat(truth = considering_dropout, estimate = .pred_class)
```

yardstick calculates the measures. `metric_set()` bundles several, and because "Yes" is the first level of the outcome, they treat "Yes" as the event of interest:

```{r}
class_metrics <- metric_set(accuracy, sensitivity, specificity, precision, f_meas)
test_results |> class_metrics(truth = considering_dropout, estimate = .pred_class)
```

```{r}
#| include: false
cm <- test_results |> count(considering_dropout, .pred_class)
tp <- cm$n[cm$considering_dropout == "Yes" & cm$.pred_class == "Yes"]
n_yes <- sum(test_results$considering_dropout == "Yes")
n_flag <- sum(test_results$.pred_class == "Yes")
```

This is sobering. Of the `r n_yes` test students who considered dropping out, the model flags only `r tp`: a sensitivity of `r round(100 * tp / n_yes)`%. The high accuracy and high specificity come almost entirely from correctly leaving alone the large majority who were never at risk. The AUC showed the model ranks students well, so the problem is not the model but the threshold: with only 15% of students at risk, few students ever get a predicted probability above 0.5.

## Choosing the threshold

The threshold of 0.5 is not a law. Lowering it flags more students: more of those at risk are found (higher recall), at the price of more false alarms (lower precision). Which balance is right depends on what the decision costs. Here, a flagged student is offered a conversation with a counsellor: cheap, and harmless if unnecessary. Missing a student who then leaves is costly. So a threshold well below 0.5 makes sense. If the decision were costly or stigmatising, a high threshold would be needed.

The threshold is a choice made while building the model, so it must be chosen without the test set. Cross-validation can provide predictions for every training student from a model that did not see them; `save_pred = TRUE` keeps them:

```{r}
set.seed(2026)
logistic_cv <- fit_resamples(logistic_wf, resamples = dropout_folds,
                             control = control_resamples(save_pred = TRUE))
cv_predictions <- collect_predictions(logistic_cv)
```

The **ROC curve** shows every possible threshold at once. For each threshold, it plots the sensitivity against the false-positive rate (1 − specificity). A useless model follows the diagonal; a good one bends towards the top-left corner, and the AUC is the area under the curve:

```{r}
#| label: fig-roc
#| fig-cap: "ROC curve of the logistic regression model, from cross-validated predictions on the training students. The points mark the thresholds 0.5 and 0.2."
#| fig-alt: "An ROC curve rising steeply from the bottom-left corner and bending towards the top-left, well above the diagonal dashed line. A point at threshold 0.5 sits low on the curve, with sensitivity about 0.3; a point at threshold 0.2 sits higher, with sensitivity about 0.6 and a false positive rate of about 0.2."
#| fig-width: 5
#| fig-height: 5
roc_points <- cv_predictions |>
  roc_curve(truth = considering_dropout, .pred_Yes)

marked <- roc_points |>
  slice(sapply(c(0.5, 0.2), \(t) which.min(abs(.threshold - t))))

ggplot(roc_points, aes(x = 1 - specificity, y = sensitivity)) +
  geom_path(colour = "#1f6f8b", linewidth = 1) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  geom_point(data = marked, size = 3, colour = "#e07b39") +
  geom_text(data = marked, aes(label = paste("threshold", round(.threshold, 1))),
            hjust = -0.15, vjust = 1.2) +
  coord_equal() +
  labs(x = "False positive rate (1 - specificity)", y = "Sensitivity (recall)") +
  theme_minimal(base_size = 12)
```

A table makes the trade-off concrete. For several thresholds, it shows the share of students flagged, and the resulting recall, precision, and specificity:

```{r}
threshold_table <- map(c(0.5, 0.4, 0.3, 0.2, 0.15, 0.1), \(t) {
  cv_predictions |>
    mutate(flag = factor(if_else(.pred_Yes >= t, "Yes", "No"), levels = c("Yes", "No"))) |>
    summarise(
      threshold   = t,
      flagged     = mean(flag == "Yes"),
      recall      = sensitivity_vec(considering_dropout, flag),
      precision   = precision_vec(considering_dropout, flag),
      specificity = specificity_vec(considering_dropout, flag)
    )
}) |>
  list_rbind()

threshold_table
```

`map()` runs the same calculation for each threshold, and `list_rbind()` stacks the results into one table. The `_vec()` versions of the yardstick functions take two vectors instead of a data frame.

```{r}
#| include: false
row20 <- threshold_table |> filter(threshold == 0.2)
```

Elaf discusses the table with the counselling service. They can offer conversations to about a quarter of the new students, so they settle on a threshold of 0.2: the model then flags about `r round(100 * row20$flagged)`% of students, finds about `r round(100 * row20$recall)`% of those at risk, and about `r round(100 * row20$precision)`% of the students it flags are truly at risk. Only now is the chosen threshold applied to the test students:

```{r}
test_results <- test_results |>
  mutate(flag = factor(if_else(.pred_Yes >= 0.2, "Yes", "No"), levels = c("Yes", "No")))

test_results |> conf_mat(truth = considering_dropout, estimate = flag)
test_results |> class_metrics(truth = considering_dropout, estimate = flag)
```

```{r}
#| include: false
tp20 <- sum(test_results$flag == "Yes" & test_results$considering_dropout == "Yes")
found_by_gender <- test_results |>
  filter(considering_dropout == "Yes") |>
  summarise(at_risk = n(), found = sum(flag == "Yes"), .by = gender)
flag20 <- sum(test_results$flag == "Yes")
```

On new students, the model now finds `r tp20` of the `r n_yes` at risk instead of `r tp`, while flagging `r flag20` students in all. About one flagged student in `r round(flag20 / tp20)` is truly at risk. The test numbers are a little lower than the cross-validated ones, as expected with only `r n_yes` at-risk students in the test set. Accuracy has fallen, and it does not matter: accuracy was never the goal.

::: {.callout-important}
## Report the threshold

Sensitivity, specificity, precision, and F1 all depend on the threshold, so a paper that reports them must say which threshold was used, and why. The AUC does not depend on the threshold, which is why it is the standard measure for comparing models, and the threshold-dependent measures are the ones for describing how a model will be used.
:::

## Imbalanced outcomes

When one outcome is rare, as considering dropout is, models tend to predict the common one. Besides moving the threshold, there are two other common remedies:

- **Resampling** the training data so that the classes are balanced: **upsampling** repeats rare-class students, **downsampling** drops common-class students, and **SMOTE** creates new, artificial rare-class students between existing ones. The **themis** package adds all three to a recipe.
- **Class weights**, which make mistakes on the rare class count more when the model is fitted. Some model engines support them directly.

Here is upsampling with themis. `over_ratio = 1` repeats "Yes" students until there are as many as "No" students:

```{r}
library(themis)

upsample_recipe <- dropout_recipe |>
  step_upsample(considering_dropout, over_ratio = 1)

upsample_wf <- workflow() |>
  add_recipe(upsample_recipe) |>
  add_model(logistic_reg())

set.seed(2026)
upsample_cv <- fit_resamples(upsample_wf, resamples = dropout_folds,
                             metrics = metric_set(roc_auc, sensitivity, precision))
collect_metrics(upsample_cv) |> select(.metric, mean)
```

Resampling steps are applied only to the training data; when the model predicts for new students, the step is skipped automatically, so the test set keeps its real balance.

```{r}
#| include: false
up <- collect_metrics(upsample_cv)
up_auc <- up$mean[up$.metric == "roc_auc"]
up_sens <- up$mean[up$.metric == "sensitivity"]
```

With the usual 0.5 threshold, the upsampled model has a sensitivity of about `r round(100 * up_sens)`%, far higher than before. But its AUC, `r sprintf("%.2f", up_auc)`, has not improved. Upsampling did not make the model better at telling students apart; it pushed all the predicted probabilities upwards, which has much the same effect as lowering the threshold. That is often how it works out, and for a model that produces probabilities, choosing the threshold directly is simpler and keeps the probabilities meaningful (after upsampling, a "probability" of 0.6 no longer means a 60% chance). Resampling is most useful for models that do not produce good probabilities, and when the rare class is very rare.

::: {.callout-warning}
## Does the model work equally well for everyone?

Among the test students, the model with a threshold of 0.2 finds `r found_by_gender$found[found_by_gender$gender == "Female"]` of the `r found_by_gender$at_risk[found_by_gender$gender == "Female"]` at-risk women and `r found_by_gender$found[found_by_gender$gender == "Male"]` of the `r found_by_gender$at_risk[found_by_gender$gender == "Male"]` at-risk men, similar shares; but with only `r n_yes` at-risk students in the test set, any difference would be hard to detect. Before a model like this is used, its sensitivity and precision should be checked for each group that matters (gender, faculty, full- and part-time study) on as much data as possible, as Chapter 10 recommended. A threshold that works on average can still miss one group of students much more often than another.
:::

::: {.callout-note icon="false"}
## In your field: ecology

Ecologists classify species from measurements. The `penguins` data in the **modeldata** package (installed with tidymodels) records the bill length, bill depth, flipper length, and body mass of `r nrow(modeldata::penguins)` penguins of three species, measured on islands near Palmer Station in Antarctica [@horst2022]. A random forest classifies the species:

```{r}
data(penguins, package = "modeldata")
penguins <- penguins |>
  drop_na(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

set.seed(1)
penguin_split <- initial_split(penguins, strata = species)

penguin_wf <- workflow() |>
  add_formula(species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g) |>
  add_model(rand_forest(trees = 500) |> set_engine("ranger") |> set_mode("classification"))

penguin_fit <- last_fit(penguin_wf, penguin_split)
collect_predictions(penguin_fit) |> conf_mat(truth = species, estimate = .pred_class)
```

With three classes, the confusion matrix has a row and a column for each species. Correct classifications lie on the diagonal, and any mistake would appear off it, showing exactly which species are confused; here, every penguin in the test set is classified correctly. `add_formula()` replaces a recipe when no preparation is needed. Unlike Elaf's data, the classes here are well separated by the measurements, and flexible models work very well.
:::

## Chapter review

### Summary

- A decision tree classifies by a series of yes-or-no questions. Trees are easy to explain and need no normalisation, but a single tree is unstable and overfits easily.
- A random forest averages hundreds of trees grown on bootstrap samples with random subsets of predictors. Permutation importance shows which predictors it relies on.
- k-nearest neighbours classifies by the vote of the $k$ most similar training cases; it needs normalised predictors, and $k$ must be tuned.
- A support vector machine separates the classes with the widest possible margin; kernels allow curved boundaries.
- Compare models with cross-validation on the same folds (`workflow_set()`). On Elaf's data, with its smooth pattern, logistic regression predicts as well as any flexible model.
- The confusion matrix counts true and false positives and negatives. Sensitivity (recall) is the share of positives found; precision is the share of flagged cases that are positive; F1 balances the two.
- The threshold turns probabilities into decisions. Choose it for the purpose of the model, using cross-validated predictions, and report it.
- For imbalanced outcomes, moving the threshold, resampling, and class weights all shift the balance between recall and precision; resampling rarely improves the AUC.

### Key terms

Decision tree, node, leaf, Gini impurity, random forest, bootstrap sample, ensemble, `mtry`, permutation importance, k-nearest neighbours, distance, support vector machine, margin, support vector, kernel, radial basis function, workflow set, threshold, confusion matrix, true positive, false positive, true negative, false negative, sensitivity, recall, specificity, precision, F1 score, ROC curve, upsampling, downsampling, SMOTE, class weights.

## Exercises

The [playground](../playground/chapter11.html) has these and more, with hints and solutions.

1. Draw the decision tree for Elaf's data with `tree_depth = 2`. Which predictors does it use? Explain the tree in plain words, as you would to a counsellor.
2. Tune the random forest's `mtry` over `c(2, 5, 10)` with cross-validation. Does the best value beat the default?
3. Extend the k-NN grid to `neighbors = c(81, 161, 301)`. Does the AUC keep rising? What happens to a k-NN model as $k$ approaches the number of training students?
4. Using the tiny example's confusion matrix (@tbl-confusion), calculate the F1 score yourself from precision and recall ($F_1 = 2 \times \text{precision} \times \text{recall} / (\text{precision} + \text{recall})$).
5. Suppose the counselling service can only contact 10% of new students. Which threshold would you recommend, and what recall and precision would they get? Use `threshold_table`.
6. Replace `step_upsample()` with `step_downsample()`. How do the sensitivity and AUC compare with upsampling?

## Further reading

- *Tidy Modeling with R* [@kuhn2022] covers workflow sets, model comparison, and class imbalance in depth.
- *An Introduction to Statistical Learning* [@james2021islr] explains trees, random forests, and support vector machines, with clear illustrations of how each draws its boundaries.
- Breiman's original paper on random forests [@breiman2001] is still readable and explains why averaging many trees works.
