# Advanced Clustering {#sec-ch13}

```{r}
#| label: setup
#| include: false
library(data2thesis)
```

In Chapter 8, Elaf used k-means to divide her students into four lifestyle profiles, and hierarchical clustering broadly agreed. But she was left uneasy. The clusters overlapped, many students switched groups between methods, and every student, however unusual, was forced into one of the four boxes. Her supervisor asks the natural follow-up questions: "How sure can you be about which profile a student belongs to? Is four really the right number? And are there students who do not fit *any* profile?"

This chapter answers them with two methods that relax the assumptions of k-means. **Gaussian mixture models** give each student a *probability* of belonging to each profile, and choose the number of profiles with a statistical criterion. **DBSCAN** defines clusters as dense regions of data, and labels students in sparse regions as **noise**: people who fit no group. The chapter ends with the question behind all clustering: how to judge whether a clustering is any good.

::: {.callout-tip appearance="simple"}
## By the end of this chapter you will be able to

- Explain the limitations of k-means: hard assignments, round clusters, and no room for outliers.
- Fit a Gaussian mixture model with mclust, choose the number of clusters with BIC, and interpret membership probabilities.
- Find dense clusters and noise points with DBSCAN, and choose its settings.
- Evaluate a clustering with the silhouette, agreement between methods (the adjusted Rand index), and interpretability.
:::

## The profile data again

The data is exactly that of Chapter 8: each student's average sleep, study hours, caffeine, and exercise across the semesters, and their stress, support, and satisfaction scores, scaled to z-scores.

```{r}
library(dplyr)
library(ggplot2)

q <- questionnaire
q$stress_4 <- 6 - q$stress_4

scores <- q |>
  mutate(
    stress       = rowMeans(pick(stress_1:stress_6), na.rm = TRUE),
    support      = rowMeans(pick(support_1:support_6), na.rm = TRUE),
    satisfaction = rowMeans(pick(satisfaction_1:satisfaction_4), na.rm = TRUE)
  ) |>
  select(student_id, stress, support, satisfaction)

profiles <- semesters |>
  summarise(across(c(sleep_hours, study_hours, caffeine_mg, exercise_days),
                   ~ mean(.x, na.rm = TRUE)),
            .by = student_id) |>
  left_join(scores, join_by(student_id)) |>
  na.omit()

profile_data <- scale(profiles |> select(-student_id))

set.seed(123)
kmeans_clusters <- kmeans(profile_data, centers = 4, nstart = 25)
```

The last two lines repeat Chapter 8's k-means solution, for comparison.

## What k-means assumes

k-means is simple and fast, but it makes three strong assumptions:

1. **Every case belongs to exactly one cluster**, with complete certainty. A student halfway between two profiles is assigned to one of them as firmly as a student at the centre.
2. **Clusters are round and similar in size.** Because each case goes to the nearest centre, k-means draws straight boundaries halfway between centres. Elongated or unequal clusters are cut up wrongly.
3. **Every case belongs to some cluster.** There is no way to say "this student fits nowhere".

Real groups of people rarely satisfy these assumptions. The two methods of this chapter each relax some of them.

## Gaussian mixture models

A **Gaussian mixture model** (GMM) assumes that the data comes from a mix of several groups, each with its own normal (Gaussian) distribution, and estimates each group's centre, spread, and size from the data.

### A one-variable example

The idea is easiest to see with one variable. R's `faithful` data records the waiting time, in minutes, between 272 eruptions of the Old Faithful geyser in Yellowstone National Park. The histogram in @fig-faithful has two humps: short waits and long waits. The **mclust** package fits a mixture model with `Mclust()`:

```{r}
library(mclust)
geyser <- Mclust(faithful$waiting)
summary(geyser, parameters = TRUE)
```

mclust tried mixtures of one to nine groups and chose two. The output describes them: the **mixing probabilities** are the sizes of the groups (about 36% and 64% of eruptions), the **means** are their centres (about 55 and 80 minutes), and the **variances** their spreads. @fig-faithful draws the two fitted normal curves over the data.

```{r}
#| label: fig-faithful
#| fig-cap: "Waiting times between eruptions of Old Faithful, with the two normal distributions of the fitted mixture model."
#| fig-alt: "Histogram of waiting times from about 43 to 96 minutes, with two humps, at about 55 and 80 minutes. Two bell curves, one smaller and one larger, are drawn over the two humps."
#| fig-width: 6
#| fig-height: 3.5
#| echo: false
p <- geyser$parameters
curves <- tibble(waiting = seq(40, 100, length.out = 300)) |>
  mutate(`Short waits` = p$pro[1] * dnorm(waiting, p$mean[1], sqrt(p$variance$sigmasq)),
         `Long waits`  = p$pro[2] * dnorm(waiting, p$mean[2], sqrt(p$variance$sigmasq))) |>
  tidyr::pivot_longer(-waiting, names_to = "group", values_to = "density")
ggplot(faithful, aes(x = waiting)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2, fill = "grey80", colour = "white") +
  geom_line(data = curves, aes(y = density, colour = group), linewidth = 1.2) +
  scale_colour_manual(values = c(`Short waits` = "#e07b39", `Long waits` = "#1f6f8b")) +
  labs(x = "Waiting time (minutes)", y = "Density", colour = NULL) +
  theme_minimal(base_size = 12)
```

Now the key difference from k-means. A waiting time of 67 minutes lies between the two groups. Instead of forcing it into one, the model gives the probability that it came from each:

```{r}
predict(geyser, newdata = c(50, 67, 85))$z |> round(2)
```

A wait of 50 minutes almost certainly belongs to the short group, and 85 minutes to the long group, but 67 minutes is uncertain. These **membership probabilities** (also called soft assignments) are the main advantage of mixture models: they say not just which group a case is in, but how sure that is.

### Choosing the number of clusters with BIC

With more variables, each group is described by a centre and a **covariance matrix**, which sets its shape: round or elongated, and tilted in any direction. Groups can be allowed to differ in size (volume), shape, and orientation, or forced to be the same. mclust names each combination with three letters, such as `EEE` (all equal) or `VVV` (all variable), and fits them all for each number of groups.

To choose among them, it uses the **Bayesian information criterion** (BIC). BIC rewards a model for fitting the data well and penalises it for every parameter it needs, so a more complex model must earn its extra parameters. In mclust, *higher* BIC is better. For Elaf's profile data:

```{r}
set.seed(123)
profile_gmm <- Mclust(profile_data, G = 1:8)
summary(profile_gmm)
```

```{r}
#| label: fig-bic
#| fig-cap: "BIC for Gaussian mixture models with one to eight clusters. Each line is one covariance structure; higher is better."
#| fig-alt: "Line chart of BIC against number of clusters from 1 to 8, with one line per model type. BIC rises steeply from 1 to 3 clusters and is highest at 3, then levels off or falls."
#| fig-width: 7
#| fig-height: 4
library(factoextra)
fviz_mclust(profile_gmm, what = "BIC")
```

```{r}
#| include: false
bic_gap <- round(max(profile_gmm$BIC, na.rm = TRUE) - max(profile_gmm$BIC["4", ], na.rm = TRUE), 1)
```

BIC chooses `r profile_gmm$G` clusters with the `r profile_gmm$modelName` structure (ellipsoidal clusters of equal size and orientation, but different shapes). The best four-cluster model is about `r bic_gap` BIC points behind. A common rule of thumb reads a BIC difference above 6 as strong evidence and above 10 as very strong, so the data favours three profiles, although four remain a reasonable alternative (exercise 1 explores them).

### Describing the profiles

As in Chapter 8, a cluster means something only once it is described:

```{r}
profiles <- profiles |>
  mutate(gmm_cluster = profile_gmm$classification)

profiles |>
  summarise(students = n(), across(sleep_hours:satisfaction, ~ round(mean(.x), 1)),
            .by = gmm_cluster) |>
  arrange(gmm_cluster)
```

The three profiles are clear. One is **balanced**: the most sleep and exercise, the least stress, and the highest support and satisfaction. One is **overloaded**: long study weeks, short sleep, a lot of caffeine, little exercise, and the most stress. The third, the largest, combines few study hours with low support and low satisfaction: students who seem **disengaged or isolated**. (The cluster numbers are arbitrary labels.) How does this compare with k-means?

```{r}
table(gmm = profile_gmm$classification, kmeans = kmeans_clusters$cluster)
```

The mixture model's three profiles correspond closely to Chapter 8's four k-means clusters: the balanced and disengaged groups largely match, and the two k-means "overloaded" clusters, one of them with very high caffeine, are joined into one. The mixture model, with its more flexible cluster shapes, did not need to split the overloaded students in two.

### How sure is each assignment?

`profile_gmm$z` holds each student's membership probabilities, one column per cluster, and `profile_gmm$uncertainty` is 1 minus the largest of them.

```{r}
head(round(profile_gmm$z, 2))

certainty <- apply(profile_gmm$z, 1, max)
sum(certainty < 0.8)
```

`apply(..., 1, max)` takes the maximum of each row. Most students belong clearly to one profile, but `r sum(certainty < 0.8)` have less than an 80% probability for their most likely profile. @fig-uncertainty shows where they are.

```{r}
#| label: fig-uncertainty
#| fig-cap: "The three mixture model clusters on the first two principal components. Larger points are students whose profile is more uncertain."
#| fig-alt: "Scatter plot of students on two principal components, coloured by three clusters. Most points are small; the larger points, marking uncertain students, lie along the boundaries where the clusters meet."
#| fig-width: 7
#| fig-height: 5
fviz_mclust(profile_gmm, what = "uncertainty")
```

The uncertain students lie where the profiles meet. For Elaf's thesis, this is honest and useful: instead of claiming that every student has one profile, she can report the share of students who clearly fit a profile, and treat the rest as mixtures. In a later analysis, she could use the probabilities themselves, for example as weights, rather than the hard labels.

## DBSCAN: clusters as dense regions

**DBSCAN** (density-based spatial clustering of applications with noise) takes a different view. A cluster is a region where cases are packed closely together, separated from other clusters by sparser regions. Cases in sparse regions belong to no cluster; they are **noise**. DBSCAN does not need the number of clusters in advance, can find clusters of any shape, and can say "this case fits nowhere".

It needs two settings:

- `eps` (epsilon): the radius of the neighbourhood around each case.
- `minPts`: the number of cases a neighbourhood must contain for the case to be at the heart of a cluster.

A case with at least `minPts` cases within `eps` is a **core point**. Core points within `eps` of each other are joined into the same cluster, and cases near a core point are added to its cluster as **border points**. Everything else is noise.

### A tiny example

Here are twelve points: two tight groups of five, and two isolated points.

```{r}
tiny <- tibble(
  x = c(1.0, 1.2, 1.1, 0.9, 1.3,   4.0, 4.2, 3.9, 4.1, 4.3,   2.5, 5.5),
  y = c(1.0, 1.1, 1.3, 1.2, 0.9,   3.0, 3.2, 3.1, 2.8, 3.0,   4.5, 0.5)
)

library(dbscan)
tiny_db <- dbscan(tiny, eps = 0.5, minPts = 3)
tiny_db$cluster
```

The two tight groups become clusters 1 and 2, and the two isolated points are labelled 0: noise. k-means with two clusters would have had to put the isolated points into one of the groups.

### Choosing eps

The result depends heavily on `eps`. A common guide is the **k-nearest-neighbour distance plot**: for each case, the distance to its $k$-th nearest neighbour, sorted from smallest to largest. Most cases have close neighbours; the curve bends sharply upwards where the isolated cases begin, and that bend suggests a value for `eps`. With `minPts` set to 8, a common choice for data with seven variables (about the number of variables plus one or more), the plot uses $k = 7$:

```{r}
#| label: fig-knndist
#| fig-cap: "Distance from each student to their seventh-nearest neighbour, sorted. The dashed line marks eps = 2."
#| fig-alt: "A curve that rises slowly from about 0.9 to about 2 over most of the 599 students, then bends sharply upwards to about 4 for the last few students. A horizontal dashed line at 2 crosses the curve where it starts to bend."
#| fig-width: 6
#| fig-height: 3.5
kNNdistplot(profile_data, k = 7)
abline(h = 2, lty = "dashed")
```

The curve bends at a distance of about 2, so Elaf uses `eps = 2`:

```{r}
profile_db <- dbscan(profile_data, eps = 2, minPts = 8)
profile_db
```

```{r}
#| include: false
n_noise <- sum(profile_db$cluster == 0)
n_db_clusters <- max(profile_db$cluster)
```

DBSCAN finds `r n_db_clusters` cluster and `r n_noise` noise points. At first this looks like a failure: no profiles at all. But it is an informative result. DBSCAN looks for dense regions separated by gaps, and Elaf's students form one continuous cloud, in which the profiles found by k-means and the mixture model overlap without gaps between them. (Smaller values of `eps` break the cloud into fragments and label hundreds of students as noise; try it in the exercises.) The profiles are real differences in *where* students sit in the cloud, not separate islands.

### Students who fit no profile

The noise points answer the supervisor's last question. Who are they?

```{r}
profiles |>
  filter(profile_db$cluster == 0) |>
  select(sleep_hours:satisfaction) |>
  round(1)
```

```{r}
#| include: false
noise_rows <- profiles |> filter(profile_db$cluster == 0)
n_extreme_work <- sum(noise_rows$study_hours > 55)
n_extreme_rest <- sum(noise_rows$study_hours < 10)
```

Two kinds of unusual student stand out. `r n_extreme_work` study around 60 hours a week, sleep about 4 hours a night or less, drink over 700 mg of caffeine a day (about seven cups of coffee), and hardly exercise: more extreme than anyone in the overloaded profile. `r n_extreme_rest` others are the opposite: close to 10 hours of sleep, very few study hours, very little caffeine, and exercise almost every day. The remaining few are extreme versions of the overloaded profile. In Chapter 5, extreme values were checked one variable at a time; DBSCAN finds students who are unusual in their *combination* of values.

What should Elaf do with them? Not delete them: they are real students, and the extreme workers are exactly the students a wellbeing service would want to know about. She reports them separately, as students who do not fit the profiles, and checks that the profiles do not change when they are left out.

## Evaluating a clustering

In supervised learning, a model is judged against the true answers. In clustering there are usually no true answers, so evaluation rests on several kinds of evidence, none decisive on its own.

**Internal measures** judge how compact and well separated the clusters are, using the data alone. The **silhouette** of Chapter 8 is the most common, and BIC plays this role for mixture models. The `silhouette()` function from the cluster package calculates it for any clustering:

```{r}
library(cluster)
profile_distances <- dist(profile_data)

mean(silhouette(kmeans_clusters$cluster, profile_distances)[, "sil_width"])
mean(silhouette(profile_gmm$classification, profile_distances)[, "sil_width"])
```

Both averages are low (the silhouette runs from −1 to 1), which confirms what DBSCAN suggested: the profiles overlap, and no clustering of this data will be crisp.

**Agreement between methods** asks whether different reasonable methods find similar groups. The **adjusted Rand index** (ARI) measures the agreement between two clusterings. It counts the pairs of cases that both clusterings put together or both put apart, and adjusts for the agreement expected by chance: 1 means identical groupings (whatever the labels), and 0 means no more agreement than chance. A tiny example shows that the labels themselves do not matter:

```{r}
first  <- c(1, 1, 1, 2, 2, 2)
second <- c("B", "B", "B", "A", "A", "A")
third  <- c(1, 1, 2, 2, 3, 3)
adjustedRandIndex(first, second)
adjustedRandIndex(first, third)
```

`first` and `second` group the six people identically, so their ARI is 1, even though the labels differ. `third` splits them differently, so its ARI is much lower. For Elaf's two methods:

```{r}
adjustedRandIndex(kmeans_clusters$cluster, profile_gmm$classification)
```

```{r}
#| include: false
ari_methods <- adjustedRandIndex(kmeans_clusters$cluster, profile_gmm$classification)
```

An ARI of `r sprintf("%.2f", ari_methods)`: substantial agreement, especially given that one solution has four clusters and the other three.

**External validation** compares clusters with known categories, also using the ARI. It is only possible when such categories exist, for example when clustering flowers of known species to test a method. For real profiles there is no answer key, which is exactly why they are being sought.

**Stability** asks whether the clusters survive small changes: a different random start, a bootstrap sample of the students, or leaving out the unusual students. Clusters that appear only under one setting should not be reported.

**Interpretability and usefulness** matter most in the end. Do the clusters make sense in the light of theory? Do they differ on outcomes they were not built from, such as GPA or considering dropout? Elaf checks the last point:

```{r}
profiles |>
  left_join(students |> select(student_id, considering_dropout), join_by(student_id)) |>
  summarise(students = n(),
            considering_dropout = round(mean(considering_dropout == "Yes"), 2),
            .by = gmm_cluster) |>
  arrange(gmm_cluster)
```

The profiles differ clearly in how many students consider dropping out, although dropout played no part in finding them. That is good evidence that they capture something real about students' situations.

::: {.callout-important}
## Report the choices, not just the clusters

A clustering depends on choices: which variables, how they are scaled, which method, how many clusters, and settings such as `eps`. Different reasonable choices can give different answers. In a thesis, report each choice and why it was made, show the evidence for the number of clusters (BIC, silhouette), say how clearly students fit (for example, the share with a membership probability above 0.8), and describe how robust the profiles are to other reasonable choices.
:::

::: {.callout-note icon="false"}
## In your field: earth sciences

DBSCAN was designed for spatial data, where clusters can have any shape. R's `quakes` data records the location of 1,000 earthquakes near Fiji since 1964. Plotted on a map, they follow the long, curved lines of two ocean trenches: clusters that are neither round nor of equal size.

```{r}
quake_db <- dbscan(quakes[, c("long", "lat")], eps = 1.5, minPts = 10)
table(quake_db$cluster)
```

```{r}
#| label: fig-quakes
#| fig-cap: "Earthquakes near Fiji, clustered by DBSCAN. Grey crosses are noise."
#| fig-alt: "Map-like scatter plot of earthquake longitude and latitude. A long curved band of points running north to south forms one cluster, a smaller band to the west forms another, and a few scattered points away from both are grey crosses."
#| fig-width: 5
#| fig-height: 5
quakes |>
  mutate(cluster = factor(quake_db$cluster)) |>
  ggplot(aes(x = long, y = lat, colour = cluster, shape = cluster == "0")) +
  geom_point(alpha = 0.7) +
  scale_colour_manual(values = c("0" = "grey60", "1" = "#1f6f8b", "2" = "#e07b39")) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 4), guide = "none") +
  coord_quickmap() +
  labs(x = "Longitude", y = "Latitude", colour = "Cluster") +
  theme_minimal(base_size = 12)
```

DBSCAN follows the two curved trenches and leaves the scattered earthquakes between and beyond them as noise. Here, unlike in Elaf's data, there really are dense regions separated by empty space, which is exactly when DBSCAN works best. (`coord_quickmap()` gives longitude and latitude the right proportions for a map.)
:::

## Chapter review

### Summary

- k-means assigns every case to exactly one round cluster, with no room for uncertainty or outliers.
- A Gaussian mixture model describes the data as a mix of normal distributions of different sizes and shapes. It gives each case a probability of belonging to each cluster, and BIC chooses the number of clusters and their shape (higher is better in mclust).
- On Elaf's data, the mixture model finds three profiles (balanced, overloaded, and disengaged) that closely match Chapter 8's k-means clusters, and shows which students fit no profile clearly.
- DBSCAN defines clusters as dense regions and labels cases in sparse regions as noise. It needs `eps` and `minPts`; the k-nearest-neighbour distance plot helps choose `eps`. It works best when clusters are separated by gaps.
- On Elaf's data, DBSCAN finds one continuous cloud of students and flags a few extreme students who fit no profile.
- Evaluate clusterings with internal measures (silhouette, BIC), agreement between methods (adjusted Rand index), stability, and above all interpretability and relevance to outcomes. Report every choice.

### Key terms

Gaussian mixture model, mixture component, mixing probability, membership probability, soft assignment, covariance matrix, Bayesian information criterion (BIC), uncertainty, DBSCAN, density, eps, minPts, core point, border point, noise, k-nearest-neighbour distance plot, internal validation, silhouette, adjusted Rand index, external validation, stability.

## Exercises

The [playground](../playground/chapter13.html) has these and more, with hints and solutions.

1. Fit `Mclust(profile_data, G = 4)` and describe the four profiles. Which of the three-cluster profiles has been split, and how does the new solution compare with Chapter 8's k-means?
2. How many students have a membership probability above 0.95 for their profile? What share of the sample is that?
3. Run DBSCAN on the profile data with `eps` of 1.2, 1.6, and 2.4. How do the number of clusters and noise points change? Why does a small `eps` label so many students as noise?
4. Leave out the DBSCAN noise students and fit the mixture model again. Do the profiles change?
5. Use `adjustedRandIndex()` to compare the three-cluster mixture model with hierarchical clustering (Ward's method, cut into three clusters) from Chapter 8.
6. Compare the three profiles on final GPA. Is the pattern what you would expect from their descriptions?

## Further reading

- *An Introduction to Statistical Learning* [@james2021islr] introduces k-means and hierarchical clustering; its chapter on unsupervised learning is a good companion to Chapters 8 and 13.
- The mclust paper [@scrucca2016] explains Gaussian mixture models, the covariance structures, and BIC, with many examples in R.
- The paper describing the dbscan package [@hahsler2019] explains DBSCAN, its settings, and its relatives, such as HDBSCAN and OPTICS.
