# Using AI with R {#sec-ch17}

```{r}
#| label: setup
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(stringr)
library(yardstick)

# The language model's results are produced once, by data-raw/run_ai_coding.R,
# and read from a file, so the book renders without an API key.
ai_file <- Sys.getenv("AI_RESULTS", "../data-raw/ai_coding_results.csv")
ai_meta_file <- sub("results", "meta", ai_file)
has_ai <- file.exists(ai_file)
```

At the end of the programme, Elaf's questionnaire asked one open question: "What has been your biggest challenge during your studies?" `r nrow(open_responses)` students answered, in their own words. Her eleventh research question (RQ11) asks what those challenges are. The traditional method is **qualitative coding**: reading every answer and assigning it to a theme. Elaf coded `r nrow(open_responses_coded)` answers by hand, which took her most of two weeks. Could an AI tool code the rest, and could she trust it?

Artificial intelligence (AI) tools now appear everywhere in research: assistants that write and explain code, and language models that read, summarise, and classify text. This chapter looks at both uses. It explains what these tools are and why they make mistakes, shows how to use an AI assistant to write R code and how to check what it produces, and then calls a language model from R with the **ellmer** package to code Elaf's answers, testing the results against her hand coding exactly as Chapter 11 tested a classifier. It ends with the rules for using AI responsibly in research.

::: {.callout-tip appearance="simple"}
## By the end of this chapter you will be able to

- Explain in plain terms how large language models work, and why they make confident mistakes.
- Use an AI assistant to write, explain, and debug R code, and check its answers.
- Call a language model from R with ellmer, and get structured answers for many texts.
- Validate AI coding of text against human coding with accuracy and Cohen's kappa.
- Use AI responsibly: protect data, keep work reproducible, and disclose its use.
:::

## How AI assistants work

The AI assistants that researchers use, such as ChatGPT, Claude, Gemini, and Copilot, are built on **large language models** (LLMs). A language model is a neural network (Chapter 14), with billions of weights, trained on an enormous amount of text: books, websites, and code. During training it learns to do one thing extremely well: predict the next word (strictly, the next **token**, a word or part of a word) given the text so far. Answering a question means generating a reply one token at a time, each time choosing a likely continuation. Further training on examples of helpful answers turns this into an assistant that follows instructions.

This explains both their strengths and their weaknesses:

- They are fluent in R, because a huge amount of R code and documentation was in their training text. They can write, explain, and fix code well, especially for common tasks.
- They produce what is *plausible*, not what is *verified*. A made-up function name that looks like a real one, a wrong argument, or an outdated way of doing something can be delivered with complete confidence. Such invented content is called a **hallucination**.
- Their knowledge stops at the date their training data was collected. Packages that changed since then, as tidymodels and ggplot2 have, may be described as they used to be.
- They know nothing about your data unless you tell them, and they cannot run your code unless the tool is designed to do so.

The practical rule follows directly: **use AI to draft, and check everything it produces.** The checking is your job, and you remain responsible for the result.

## AI as a coding assistant

### Asking good questions

An assistant can only be as specific as the question. A good request for R code includes:

- **The goal**, in words: "For each faculty, I want the percentage of students who considered dropping out."
- **The data**: the names and types of the relevant columns. Pasting the output of `glimpse(students)` (Chapter 3) is ideal, but never paste real, identifiable data.
- **The tools**: "Use dplyr with the native pipe `|>`", so the answer matches the style you know.
- **For errors**: the exact error message and the code that produced it.

### Always check the answer

Suppose Elaf asks for the percentage of students in each faculty who considered dropping out, and an assistant suggests:

```{r}
students |>
  summarise(percent = 100 * sum(considering_dropout == "Yes") / nrow(students),
            .by = faculty)
```

The code runs without an error and the numbers look reasonable. But they add up to only `r round(100 * mean(students$considering_dropout == "Yes"))`, the percentage of *all* students who considered dropping out: `nrow(students)` counts every student, not the students in each faculty. The question asked for the share *within* each faculty:

```{r}
students |>
  summarise(percent = 100 * mean(considering_dropout == "Yes"),
            .by = faculty)
```

The mean of a TRUE/FALSE condition is the proportion of TRUE values (Chapter 5), calculated here within each faculty. The first answer was not a hallucination; it was a plausible misunderstanding, the most dangerous kind of error, because nothing looks wrong. Useful checks:

- **Run it on a tiny example** where you know the answer (every chapter of this book starts with one).
- **Check a number by another route**: here, `students |> count(faculty, considering_dropout)` gives the counts to check against.
- **Read the help** for any function you do not know (`?summarise`), to confirm it exists and does what the assistant says.
- **Ask the assistant to explain** each line, and see whether the explanation matches what you wanted.

Assistants are also excellent teachers: "explain this code line by line", "why does this give NA?", and "what does this error mean?" are among their most useful questions.

### AI inside the editor

Assistants can also work inside RStudio and Positron. **GitHub Copilot**, which can be switched on in RStudio's options, suggests code as you type, completing a line or a whole block from a comment such as `# plot wellbeing by semester`. **Positron Assistant** and similar tools add a chat that can see your open files and, with permission, your data's structure. Such suggestions arrive quickly and look authoritative, so the same rule applies: read each suggestion before accepting it. These tools change fast; the ideas in this chapter apply to whichever one you use.

## Coding Elaf's open-ended answers

Here are a few of the answers:

```{r}
open_responses |>
  slice(c(5, 9, 15, 55)) |>
  pull(biggest_challenge)
```

They vary in length and style, often mention more than one challenge, and sometimes contain spelling mistakes, as real answers do. Elaf's codebook has seven themes, and she coded a random sample of `r nrow(open_responses_coded)` answers into the theme that each student presents as their main challenge:

```{r}
open_responses_coded |>
  count(theme, sort = TRUE)
```

Her hand-coded sample is the benchmark for any automatic method, exactly like the test set of Chapter 10.

### A first attempt without AI: keywords

The simplest automatic method is a **dictionary**: a list of keywords for each theme. Each answer is assigned to the theme whose keywords it contains most often, and to "Other" if it contains none. `str_count()` from stringr counts the matches of a **regular expression**, a text pattern in which `|` means "or" and `\\b` marks the start of a word:

```{r}
keywords <- c(
  Supervision = "\\b(supervis|feedback|guidance|meeting)",
  Workload    = "\\b(time|deadline|workload|too much|assignment|reading|busy)",
  Finances    = "\\b(money|fee|scholarship|rent|afford|salary|income|funding|stipend|loan)",
  Family      = "\\b(famil|child|kid|son\\b|daughter|baby|parent|mother|father|husband|wife)",
  Health      = "\\b(sleep|tired|stress|anxi|health|ill\\b|burnout|headache|coffee|exhaust)",
  Isolation   = "\\b(lonel|alone|friends|outsider|miss (home|my)|far from|know no one)",
  Other       = "\\b(ethic|participant|statistic|library|procedure|english|power cut|laptop|internet|software)"
)

code_by_keywords <- function(answer) {
  hits <- sapply(keywords, \(pattern) str_count(str_to_lower(answer), pattern))
  if (all(hits == 0)) "Other" else names(keywords)[which.max(hits)]
}

code_by_keywords("I miss my family and I cannot afford the fees.")
```

The example answer matches one Family keyword ("family"), one Isolation phrase ("miss my"), and two Finances keywords ("afford" and "fees"), so it is coded as Finances, although a human reader might well code it differently. Applying the function to the hand-coded answers:

```{r}
themes <- names(keywords)

keyword_check <- open_responses_coded |>
  mutate(keyword_theme = sapply(biggest_challenge, code_by_keywords),
         theme = factor(theme, levels = themes),
         keyword_theme = factor(keyword_theme, levels = themes))

keyword_check |> accuracy(truth = theme, estimate = keyword_theme)
keyword_check |> kap(truth = theme, estimate = keyword_theme)
```

```{r}
#| include: false
kw_acc <- accuracy(keyword_check, theme, keyword_theme)$.estimate
kw_kap <- kap(keyword_check, theme, keyword_theme)$.estimate
```

The keywords agree with Elaf's coding for `r round(100 * kw_acc)`% of the answers. **Cohen's kappa** (`kap()`) is the usual measure of agreement between two coders: it corrects the raw agreement for the agreement expected by chance, so 0 means no better than guessing and 1 means perfect agreement [@cohen1960]. A common rule of thumb calls 0.41–0.60 moderate, 0.61–0.80 substantial, and above 0.80 almost perfect agreement [@landis1977]. The keywords reach `r sprintf("%.2f", kw_kap)`: moderate. They fail because meaning is not in single words: "my supervisor" in "I feel I am letting down both my family and my supervisor" is about family, and "I never have enough time" can be about workload or about children. Reading for meaning is exactly what language models are good at.

### Calling a language model from R

The **ellmer** package connects R to language models from many providers: `chat_anthropic()` for Claude, `chat_openai()` for ChatGPT's models, `chat_google_gemini()` for Gemini, and `chat_ollama()` for models that run on your own computer. Using a provider's models from code requires an **API key**, a secret password linked to your account, and each request costs a small amount. Store the key in your personal `.Renviron` file (`usethis::edit_r_environ()` opens it), never in a script that others may see:

```bash
ANTHROPIC_API_KEY=your-key-here
```

A chat then works much like the website, from R:

```r
library(ellmer)
chat <- chat_anthropic(model = "claude-sonnet-5")
chat$chat("In one sentence, what does Cohen's kappa measure?")
```

The `model` argument names the exact model. Always set it explicitly and record it: providers update and retire models, and different models give different answers.

### Structured answers for many texts

For coding, Elaf needs two things that a chat on a website does not give: the same instructions for every answer, and a reply in a fixed format that R can use. The **system prompt** holds the instructions: the task and her codebook, with a definition of each theme. A **type** describes the required answer: here, an object with one field, `theme`, which must be one of the seven themes:

```r
codebook <- "
You are helping a researcher code open-ended survey answers from graduate
students, who were asked: 'What has been your biggest challenge during your
studies?' Assign each answer to exactly ONE theme: the main challenge the
student describes. If several challenges are mentioned, choose the one the
student presents first or as most important.

Themes:
- Supervision: the supervisor; feedback, guidance, meetings, disagreements.
- Workload: too much work or too little time; deadlines, coursework, reading.
- Finances: money, fees, scholarships, living costs, paid work to pay for study.
- Family: children, partners, parents, caring duties, studying at home.
- Health: physical or mental health, sleep, stress, anxiety, burnout, illness.
- Isolation: loneliness, being far from home, no colleagues or friends.
- Other: anything else, such as ethics approval, statistics, academic English.
"

theme_type <- type_object(
  theme = type_enum(themes, "The single main theme of the answer.")
)

chat <- chat_anthropic(system_prompt = codebook, model = "claude-sonnet-5",
                       params = params(temperature = 0))

ai_codes <- parallel_chat_structured(
  chat,
  prompts = as.list(open_responses$biggest_challenge),
  type = theme_type
)
```

`type_enum()` restricts the answer to the listed themes, so the model cannot invent a new one or reply with a paragraph. `temperature = 0` asks the model to choose its most likely answer every time, rather than sampling more freely, which makes the results as repeatable as possible. `parallel_chat_structured()` sends one request per answer, several at a time, and returns a data frame with one row per answer and a `theme` column.

Elaf's full script, `data-raw/run_ai_coding.R` in the book's repository, also records the model, the date, the number of tokens, and the cost, and codes the `r nrow(open_responses_coded)` hand-coded answers a second time to check that the model gives the same answers when asked again. It was run once, and the results were saved to a file, so that this chapter can analyse them without calling the model again.

### How well does it agree with Elaf?

```{r}
#| child: !expr if (has_ai) "_17-ai-results.qmd" else "_17-ai-pending.qmd"
```

## Using AI responsibly

### Check, and stay responsible

Everything an AI tool produces is a draft. Check code by running it on known cases, check facts and references against their sources (assistants are known to invent plausible references that do not exist), and check coding against human coding, as Elaf did. Whatever appears in your thesis is your responsibility, whoever or whatever drafted it.

### Protect your data

Text sent to an online AI service leaves your computer and is processed, and possibly stored, by the provider. Before sending research data:

- Check your **ethics approval and consent forms**. Participants who agreed to have their answers read by the research team did not necessarily agree to have them sent to a company.
- Check the provider's terms. Many offer research or enterprise agreements under which data is not stored or used for training; free accounts often do not.
- **Remove identifying information** before sending anything: names, places, and details that could identify a person (Chapter 16).
- For sensitive data, use a **local model** that runs on your own computer, through Ollama and `chat_ollama()`. The data never leaves the machine. Local models are smaller and usually less accurate, so validate them in the same way.

Elaf's answers are anonymous, and her consent form allowed processing by external services; your study may differ.

### Keep it reproducible

Language models can give different answers to the same question on different days, and models are updated and retired. To keep AI-assisted work reproducible:

- Record the provider, the exact model name, the date, and the settings (such as the temperature).
- Save the prompts (the system prompt and codebook) with your code.
- Save the model's outputs, as Elaf did, and analyse the saved outputs rather than calling the model again.
- Check consistency by coding a sample twice.

### Disclose it

Journals and universities increasingly require authors to disclose how they used AI. The consensus of publishers and of the Committee on Publication Ethics (COPE) is that an AI tool cannot be an author, because it cannot take responsibility for the work, and that its use must be described. A methods section might say: "Answers were coded into seven themes by a large language model (Claude Sonnet 5, Anthropic, accessed through the ellmer R package, temperature 0) using the codebook in Appendix X. Agreement with the author's hand coding of a random sample of 200 answers was assessed with Cohen's kappa." For help with code or language, a sentence in the acknowledgements or methods is usually enough; check your university's and journal's policy.

### Be aware of bias

Language models learn from human text, and they can reproduce its biases: in how they describe groups of people, in which answers they find typical, and in how well they understand non-standard English or answers written by non-native speakers. Check whether the model's accuracy differs between groups, as Chapter 10 recommended for any model that makes decisions about people.

::: {.callout-note icon="false"}
## In your field: medicine and public health

Clinical notes, discharge letters, and incident reports hold information that is written as free text but needed as data: diagnoses, medications, doses, and dates. Language models can extract it into a fixed structure. With ellmer, the type describes the fields to extract:

```r
medication_type <- type_object(
  drug = type_string("Name of the medication"),
  dose_mg = type_number("Dose in milligrams"),
  times_per_day = type_integer("How many times a day it is taken")
)

chat <- chat_anthropic(model = "claude-sonnet-5")
chat$chat_structured(
  "Patient to continue metformin 500 mg twice daily with meals.",
  type = medication_type
)
```

The result is a list with `drug`, `dose_mg`, and `times_per_day`, ready to become a row of a data frame. Studies in this area validate the extraction against records checked by clinicians, report agreement for each field, and use local models or secure institutional services, because patient data must not be sent to public AI services.
:::

## Chapter review

### Summary

- Large language models generate text by predicting the next token; they are fluent in R but produce what is plausible, not what is verified. They hallucinate, they may be out of date, and they know nothing about your data unless told.
- Ask assistants specific questions (goal, data structure, tools, exact error), and check every answer: run it on a tiny example, verify numbers by another route, and read the help.
- A dictionary of keywords is a simple, transparent baseline for coding text, but it misses meaning.
- ellmer calls language models from R. Store the API key in `.Renviron`; set the model explicitly; use a system prompt with a codebook, a type to fix the format of the answer, and `parallel_chat_structured()` for many texts.
- Validate AI coding against human coding with accuracy, a confusion matrix, and Cohen's kappa, and examine the disagreements.
- Use AI responsibly: you are responsible for the result; protect participants' data (consent, anonymisation, local models); record models, prompts, and outputs; disclose AI use; and watch for bias.

### Key terms

Artificial intelligence, large language model, token, hallucination, training cut-off, prompt, system prompt, AI coding assistant, qualitative coding, codebook, dictionary method, regular expression, Cohen's kappa, API, API key, structured output, temperature, local model, disclosure.

## Exercises

The [playground](../playground/chapter17.html) has these and more, with hints and solutions.

1. Ask an AI assistant to write code that calculates the average sleep in each faculty and semester. Check the answer with a tiny example and by another route. Did it handle missing values?
2. Add three keywords to the dictionary that you think would fix some of its mistakes. Does the kappa improve? Are you now fitting the dictionary to these 200 answers (Chapter 10's overfitting)?
3. Read ten answers where the keyword coding disagrees with Elaf's coding. Would you agree with Elaf in every case?
4. If you have an API key, code 20 of the hand-coded answers with a language model, using the codebook in this chapter. How many agree with Elaf? Then change one theme definition and code them again.
5. Write the paragraph for your own methods section describing how you would use and validate AI coding in a study of your own.

## Further reading

- The ellmer documentation at ellmer.tidyverse.org covers providers, structured data, and tools, with many examples.
- Gilardi and colleagues compared language models with human coders for text annotation [@gilardi2023], a study that prompted much of the current interest in using them for research.
- Cohen's original paper on kappa [@cohen1960] and Landis and Koch's benchmarks [@landis1977] are the standard references for agreement between coders.
