```{r}
#| include: false
ai <- read.csv(ai_file)
ai_meta <- read.csv(ai_meta_file)

local_run <- grepl("Ollama", ai_meta$provider)
where_run <- if (local_run) "run locally through Ollama" else paste("from", ai_meta$provider)
cost_text <- if (ai_meta$cost_usd == 0) "at no cost" else sprintf("at a cost of about $%.2f", ai_meta$cost_usd)

kappa_label <- function(k) {
  cut(k, c(-Inf, 0.2, 0.4, 0.6, 0.8, Inf),
      labels = c("slight", "fair", "moderate", "substantial", "almost perfect")) |>
    as.character()
}
```

The script used the model `r ai_meta$model`, `r where_run`, on `r format(as.Date(ai_meta$date), "%d %B %Y")`. The `r ai_meta$requests` requests took about `r round(ai_meta$minutes)` minutes, `r cost_text`, and `r if (ai_meta$failed == 0) "every reply was one of the seven themes" else paste(ai_meta$failed, "replies were not one of the seven themes")`. The results are joined to the hand-coded answers:

```{r}
ai <- read.csv(ai_file)

ai_check <- open_responses_coded |>
  left_join(ai, join_by(student_id)) |>
  mutate(across(c(theme, ai_theme, ai_theme_rerun), \(x) factor(x, levels = themes)))

ai_check |> accuracy(truth = theme, estimate = ai_theme)
ai_check |> kap(truth = theme, estimate = ai_theme)
```

```{r}
#| include: false
ai_acc <- accuracy(ai_check, theme, ai_theme)$.estimate
ai_kap <- kap(ai_check, theme, ai_theme)$.estimate
rerun_agree <- mean(ai_check$ai_theme == ai_check$ai_theme_rerun, na.rm = TRUE)
rerun_kap <- kap(ai_check, ai_theme, ai_theme_rerun)$.estimate
```

The model agrees with the hand coding on `r round(100 * ai_acc)`% of the answers, with a kappa of `r sprintf("%.2f", ai_kap)` (`r kappa_label(ai_kap)` agreement), against `r round(100 * kw_acc)`% and `r sprintf("%.2f", kw_kap)` for the keywords. The confusion matrix of Chapter 12 shows where the two disagree:

```{r}
ai_check |> conf_mat(truth = theme, estimate = ai_theme)
```

The diagonal holds the answers on which the hand coding and the model agree. The disagreements are worth reading one by one, because they show whether the model is wrong, the codebook is unclear, or the answer is genuinely ambiguous:

```{r}
ai_check |>
  filter(theme != ai_theme) |>
  select(Answer = biggest_challenge, `Hand coding` = theme, Model = ai_theme) |>
  head(5) |>
  knitr::kable()
```

Look especially at answers that mention two challenges. If the model and the human coder chose different ones as the main challenge, that is a question for the codebook, not only for the model: a clearer rule ("code the challenge mentioned first") helps both human and machine coders.

`r if (rerun_agree == 1) "Asked a second time, the model gave exactly the same theme for every answer: with a temperature of 0, this local model is repeatable." else paste0("Asked a second time, the model gave the same theme for ", round(100 * rerun_agree), "% of the answers (kappa ", sprintf("%.2f", rerun_kap), ").")` That is not guaranteed for every model or setting (online models, in particular, can be updated between runs), which is why the outputs are saved and analysed from the file.

### What the students said

With its agreement checked, the model's coding is used for all `r nrow(ai)` answers:

```{r}
#| label: fig-ai-themes
#| fig-cap: "Main challenge described by the students, coded by a language model."
#| fig-alt: "Horizontal bar chart of the share of answers in each of seven themes, sorted from most to least common."
#| fig-width: 6.5
#| fig-height: 3.5
ai |>
  count(ai_theme) |>
  mutate(share = n / sum(n)) |>
  ggplot(aes(x = share, y = reorder(ai_theme, share))) +
  geom_col(fill = "#2f6793") +
  scale_x_continuous(labels = scales::label_percent()) +
  labs(x = "Share of answers", y = NULL) +
  theme_minimal(base_size = 12)
```

Because the themes are now data, they can be linked to the rest of the study. Students whose main challenge is supervision should report less supervisor support in the questionnaire:

```{r}
q_support <- questionnaire |>
  mutate(support = rowMeans(pick(support_1:support_6), na.rm = TRUE)) |>
  select(student_id, support)

ai |>
  left_join(q_support, join_by(student_id)) |>
  summarise(students = n(), support = round(mean(support, na.rm = TRUE), 2),
            .by = ai_theme) |>
  arrange(support)
```

This is a check of **validity**: the text and the numbers tell the same story. It also answers RQ11 in a way neither source could alone: what students struggle with, in their own words, and how that relates to their situation.
