# R Packages {#sec-app-packages}

```{r}
#| label: setup
#| include: false
# The table is built from packages.txt, next to this file. Chapters are found
# automatically in each chapter's text; the version is the one installed when
# the book was built.
chapter_files <- sort(list.files(pattern = "^[0-9]{2}-.*\\.qmd$|^_17-.*\\.qmd$"))
chapter_number <- as.integer(sub("^_?([0-9]{2}).*", "\\1", chapter_files))
chapter_text <- lapply(chapter_files, function(f) paste(readLines(f, encoding = "UTF-8", warn = FALSE), collapse = "\n"))

lines <- readLines("packages.txt", encoding = "UTF-8", warn = FALSE)
lines <- lines[nzchar(trimws(lines)) & !startsWith(lines, "#")]
fields <- lapply(strsplit(lines, "|", fixed = TRUE), trimws)
packages <- data.frame(
  group = vapply(fields, `[`, character(1), 1),
  package = vapply(fields, `[`, character(1), 2),
  purpose = vapply(fields, `[`, character(1), 3),
  extra = vapply(fields, function(f) if (length(f) >= 4) f[4] else "", character(1))
)

find_chapters <- function(pkg, extra) {
  p <- gsub(".", "\\.", pkg, fixed = TRUE)
  patterns <- c(paste0("library\\(", p, "\\)"), paste0("(^|[^A-Za-z0-9_.])", p, "::"),
                paste0("set_engine\\(\"", p, "\""), paste0("package = \"", p, "\""))
  hit <- vapply(chapter_text, function(x) any(vapply(patterns, grepl, logical(1), x = x)), logical(1))
  ch <- chapter_number[hit]
  if (nzchar(extra)) ch <- c(ch, as.integer(strsplit(extra, ",")[[1]]))
  ch <- sort(unique(ch))
  if (length(ch) > 6) paste0(paste(ch[1:5], collapse = ", "), ", and later") else paste(ch, collapse = ", ")
}
packages$chapters <- mapply(find_chapters, packages$package, packages$extra)
packages$version <- vapply(packages$package, function(p) {
  if (requireNamespace(p, quietly = TRUE)) as.character(packageVersion(p)) else "–"
}, character(1))
```

This appendix lists every R package used in the book: what it is for, the chapters that use it, and the version used when the book was built. Appendix A shows how to install them all with one command. A dash in the Version column marks a package that is recommended to readers but was not needed to build the book itself.

A package is a collection of functions, data, and documentation that someone has written and shared. R comes with a set of **base packages**, such as stats (`t.test()`, `lm()`) and utils (`read.csv()`), which are always available; everything else is installed once with `install.packages()` and loaded in each session with `library()` (Chapter 1).

## The packages in this book

```{r}
#| label: tbl-packages-all
#| tbl-cap: "The R packages used in this book."
#| echo: false
knitr::kable(
  packages[, c("group", "package", "purpose", "chapters", "version")],
  col.names = c("Area", "Package", "What it is for", "Chapters", "Version"),
  row.names = FALSE
)
```

Two entries are collections rather than single packages. **tidyverse** installs and loads dplyr, tidyr, stringr, readr, ggplot2, purrr, and a few others; **tidymodels** does the same for the modelling packages of Chapters 10 to 14 (rsample, recipes, parsnip, workflows, tune, and yardstick, among others). Several model packages (ranger, kknn, kernlab, glmnet, xgboost, nnet) are rarely loaded by name: tidymodels calls them as **engines**, as in `set_engine("ranger")`, but they must be installed.

## Finding and choosing packages

CRAN, R's official package archive, holds more than 20,000 packages, and many more are shared on GitHub. A few ways to find the right one:

- **CRAN Task Views** ([cran.r-project.org/web/views](https://cran.r-project.org/web/views/)) are curated lists of packages by topic, such as psychometrics, survival analysis, or time series, maintained by experts in each field.
- A package's **vignettes**, long-form tutorials that come with it, are the best introduction: `browseVignettes("dplyr")` lists them. Many packages also have a website with examples.
- Methods papers in journals such as the *Journal of Statistical Software* and *The R Journal* describe many packages in depth.

Before relying on a package for your thesis, a few signs show whether it is trustworthy: it is on CRAN (which checks that packages install and run); it was updated in the last year or two; it has documentation and examples; it is described in a published paper or widely used in your field; and its authors are known in the area. The packages in this book meet all or most of these.

## Citing packages

Package authors are researchers too, and citing their work is how they receive credit. `citation()` gives the recommended reference for R itself, and `citation("package")` for a package:

```{r}
citation("psych")
```

Cite the packages that did substantial work in your analysis, and give their version numbers (Chapter 16). `packageVersion("psych")` shows the version installed on your computer.
