# Common Errors and How to Fix Them {#sec-app-errors}

```{r}
#| label: setup
#| include: false
library(data2thesis)
library(dplyr)
library(ggplot2)
library(lme4)
```

Everyone who writes R code sees error messages every day, experts included. An error is not a sign that you are doing badly; it is R telling you, as precisely as it can, what it could not do. This appendix collects the messages that readers of this book are most likely to meet, what each one means, and how to fix it. The messages below are real: each example was run when the book was built, so the wording is what your R will show (it can differ slightly between versions).

## Reading a message

R produces three kinds of messages:

- An **error** stops the code: nothing after it runs. It starts with `Error`.
- A **warning** lets the code finish but tells you something may be wrong. It starts with `Warning`. Never ignore a warning without understanding it: several below mean your results contain missing values or were calculated differently from what you intended.
- A **message** is information only, such as a package telling you it has loaded.

When a long message appears, read it from the end: the last lines usually say what went wrong, and lines starting with `ℹ` point to where. Newer packages (dplyr, ggplot2, tidymodels) write especially helpful messages, often with a suggestion (`Did you mean ...?`).

## Starting out

### object '...' not found

```{r}
#| error: true
mean(sleep_hours)
```

R does not know the name. The usual causes: a typing mistake (names are case-sensitive: `Sleep_hours` is not `sleep_hours`); the object was never created, because the line that creates it was not run; or the name is a column inside a data frame, which must be reached through the data frame, as in `mean(semesters$sleep_hours)` or inside a dplyr verb. If a document fails to render with this error although the code works in the Console, the object was created in the Console but not in the document (Chapter 16).

### could not find function "..."

```{r}
#| error: true
Mean(c(6.5, 7, 5.5))
```

Either the function name is misspelled (here, `mean` with a capital M), or it belongs to a package that is not loaded. Load the package with `library()`, or write `package::function()`. The pipe `%>%` from older code gives the same error until dplyr or magrittr is loaded; the native pipe `|>` needs no package.

### there is no package called '...'

```{r}
#| error: true
library(ggplot)
```

The package is not installed, or its name is misspelled (the package is ggplot2). Install it once with `install.packages("ggplot2")`, then load it with `library(ggplot2)`. Appendix A lists every package the book uses.

### unexpected symbol, unexpected ')', unexpected string constant

```{r}
#| error: true
mean(c(6.5, 7) na.rm = TRUE)
```

A **syntax error**: R cannot read the line at all. The `^` marks where R got lost; the mistake is usually just before it. Here a comma is missing before `na.rm`. Other common causes are an extra or missing bracket (`unexpected ')'`), a missing comma between two pieces of text (`unexpected string constant`), and a missing quotation mark. RStudio highlights matching brackets and marks syntax errors with a red cross in the margin before you even run the code.

### The Console shows `+` and nothing happens

R is waiting for the rest of an unfinished command, usually because a bracket or quotation mark is not closed. Press **Esc** to cancel, fix the line, and run it again.

### argument "..." is missing, with no default

```{r}
#| error: true
rnorm()
```

A required argument was not given. The help page (`?rnorm`) lists the arguments; those without a default value (here `n`, the number of values) must be supplied.

### A misspelled argument, with no error at all

```{r}
mean(c(6.5, NA, 7), na_rm = TRUE)
```

Not every mistake produces a message. The argument is `na.rm`, not `na_rm`, and because `mean()` accepts extra arguments, the misspelled one is silently ignored: the missing value is not removed, and the answer is `NA`. When a result looks wrong, check the spelling of every argument against the help page.

## Files and folders

### cannot open file ...: No such file or directory

```{r}
#| error: true
#| warning: true
read.csv("studnets.csv")
```

R looked for the file in the **working directory** and did not find it. Check the spelling of the file name (here `studnets`), check that the file is in the project folder, and build paths with `here::here()` inside an RStudio Project (Chapter 1), so they work on any computer. `list.files()` shows the files R can see.

### cannot change working directory

```{r}
#| error: true
setwd("C:/Users/Elaf/Documents/thesis")
```

The folder does not exist on this computer, which is exactly why `setwd()` with a full path breaks as soon as a script is shared or moved. Use an RStudio Project and relative paths instead (Chapter 1).

## Data

### non-numeric argument to binary operator

```{r}
#| error: true
"7" + 1
```

Arithmetic on text. The quotation marks make `"7"` a piece of text, not a number. In real data, this happens when a numeric column was imported as text, often because of one stray value such as `"7 hrs"` or a missing-value code. Check with `str()` or `glimpse()`, and convert with `as.numeric()` or `readr::parse_number()` (Chapters 2 and 3).

### argument is not numeric or logical: returning NA

```{r}
#| warning: true
mean(c("6.5", "7"))
```

A **warning**, not an error, and the result is `NA`: the same problem as above, a numeric column stored as text. Convert the column first.

### NAs introduced by coercion

```{r}
#| warning: true
as.numeric(c("6.5", "7 hrs", "six"))
```

Values that could not be converted to numbers became `NA`. Look at which ones (`"7 hrs"` and `"six"` here) before going on: `parse_number()` can rescue `"7 hrs"`, but `"six"` needs a decision (Chapter 3). Losing data silently through this warning is one of the most common problems in real analyses.

### The result is NA

```{r}
mean(c(6.5, NA, 7))
```

Not an error: any calculation that includes a missing value gives `NA`, because the true answer is unknown. Add `na.rm = TRUE` to calculate from the available values, and report how many were missing (Chapter 5).

### object '...' not found, inside dplyr

```{r}
#| error: true
students |> filter(facultty == "Education")
```

A column name is misspelled. dplyr says in which argument the problem is (`ℹ In argument: ...`) and which name it could not find. `names(students)` lists the correct names.

### arguments imply differing number of rows; replacement has ... rows

```{r}
#| error: true
data.frame(student = 1:3, sleep = c(6.5, 7))
```

Every column of a data frame must have the same length. The data behind a new column has more or fewer values than the data frame has rows; check the lengths with `length()` and `nrow()`.

### $ operator is invalid for atomic vectors

```{r}
#| error: true
x <- c(sleep = 6.5, study = 30)
x$sleep
```

`$` works on data frames and lists, but `x` is a vector. Use `x["sleep"]` for a vector, and check what kind of object you have with `class()` or `str()`.

### subscript out of bounds

```{r}
#| error: true
results <- list(6.5, 7)
results[[3]]
```

You asked for an element that does not exist: the third element of a list with two. Check the length with `length()`, and the names with `names()`.

### invalid factor level, NA generated

```{r}
#| warning: true
answer <- factor(c("Yes", "No"))
answer[1] <- "Maybe"
answer
```

A factor accepts only its existing levels (Chapter 2); any other value becomes `NA`. Add the level first with `levels()`, or work with the column as text and convert it to a factor at the end.

### Detected an unexpected many-to-many relationship

```{r}
#| warning: true
students_small <- data.frame(id = c(1, 1), group = c("A", "B"))
scores_small <- data.frame(id = c(1, 1), score = c(3.2, 4.1))
left_join(students_small, scores_small, join_by(id))
```

A **join** (Chapter 3) found identifiers that appear more than once in *both* tables, so every copy was matched with every other, multiplying rows. Usually one of the tables should have had one row per identifier: check for duplicates with `count(id) |> filter(n > 1)`, and remove them or join by more columns.

### Can't combine ... <integer> and ... <character>

```{r}
#| error: true
tidyr::pivot_longer(students, cols = c(age, gender))
```

`pivot_longer()` puts the chosen columns into one column, so they must be of the same type: here a number and a text column. Reshape only columns of the same type, or convert them first.

## Plots

### Cannot use `+` with a single argument

```{r}
#| error: true
plot <- ggplot(students, aes(x = age))
+ geom_histogram()
```

In ggplot2, the `+` must come at the *end* of a line, not the start of the next one. Otherwise R thinks the first line is complete, and the second line starts with a stray `+`. The message even asks: "Did you accidentally put `+` on a new line?"

### `mapping` must be created by `aes()` ... Did you use `%>%` or `|>` instead of `+`?

```{r}
#| error: true
ggplot(students, aes(x = age)) |> geom_histogram()
```

Layers of a ggplot are added with `+`, not with the pipe. The pipe passes data *into* `ggplot()`; after that, it is `+` all the way.

### object '...' not found, in a plot

```{r}
#| error: true
ggplot(students) + geom_point(x = age, y = financial_worry)
```

Columns must be mapped inside `aes()`: `geom_point(aes(x = age, y = financial_worry))`. Outside `aes()`, R looks for objects called `age` and `financial_worry` and does not find them (Chapter 4 on mapping versus setting).

### `stat_count()` must only have an x or y aesthetic

```{r}
#| error: true
ggplot(students, aes(x = faculty, y = age)) + geom_bar()
```

`geom_bar()` counts the rows in each category, so it takes only `x`. To plot a value you have calculated, such as a mean per faculty, use `geom_col()`.

## Statistical tests and models

### grouping factor must have exactly 2 levels

```{r}
#| error: true
t.test(age ~ faculty, data = students)
```

A two-sample t-test compares exactly two groups, but `faculty` has five. Use ANOVA for more than two groups (Chapter 7), or filter the data to the two groups you want to compare.

### not enough 'x' observations

```{r}
#| error: true
t.test(c(6.5))
```

The test needs more data than it was given, often because filtering or missing values left too few cases. Check the number of cases with `nrow()` or `sum(!is.na(x))`.

### contrasts can be applied only to factors with 2 or more levels

```{r}
#| error: true
lm(age ~ gender, data = filter(students, gender == "Female"))
```

A categorical predictor has only one value in the data used, here because the data was filtered to women only, so there is nothing to compare. Remove the predictor, or check the filtering.

### Chi-squared approximation may be incorrect

```{r}
#| warning: true
chisq.test(matrix(c(3, 1, 2, 4), nrow = 2))
```

Some expected counts are below 5, so the chi-square test's p-value may be inaccurate (Chapter 6). Use Fisher's exact test (`fisher.test()`), or combine small categories.

### glm.fit: fitted probabilities numerically 0 or 1 occurred; algorithm did not converge

```{r}
#| warning: true
#| results: hide
glm(passed ~ hours, family = binomial,
    data = data.frame(hours = 1:10, passed = c(0, 0, 0, 0, 0, 1, 1, 1, 1, 1)))
```

In a logistic regression, a predictor separates the outcomes perfectly (here, everyone above 5 hours passed), so the coefficients become enormous and meaningless, and the fitting algorithm may also report that it did not converge. With real data, this usually means a very small group or a predictor that is almost the outcome itself. Check the cross-table of the outcome and the predictor, and consider simplifying the model.

### boundary (singular) fit: see help('isSingular')

```{r}
#| message: true
#| results: hide
lmer(wellbeing ~ semester + (semester | supervisor_id),
     data = left_join(semesters, students, join_by(student_id)))
```

A mixed-effects model estimated some random-effect variance as zero, or a correlation as exactly ±1: the random effects part is too complex for the data (Chapter 9). Simplify it, for example by removing the random slope: `(1 | supervisor_id)`.

### Observations deleted due to missingness

Not an error, but a line in the output of `summary()` for `lm()` and `glm()`: rows with a missing value in any variable of the model are left out. Check how many with `nobs(model)`, report the number used, and think about whether the missing rows differ from the others (Chapters 5 and 7).

## Machine learning

### For a classification model, the outcome should be a <factor>

```{r}
#| error: true
library(tidymodels)
fit(logistic_reg(), considering_dropout ~ age, data = students)
```

tidymodels needs a categorical outcome to be a factor. Convert it first, and set the event of interest as the first level (Chapter 10): `mutate(considering_dropout = factor(considering_dropout, levels = c("Yes", "No")))`.

### Can't select columns that don't exist ... `.pred_Yes`

```{r}
#| error: true
tibble(truth = factor(c("Yes", "No")), probability = c(0.2, 0.8)) |>
  roc_auc(truth, .pred_Yes)
```

A column name does not exist in the data given to a yardstick function. The prediction columns created by `augment()` are named after the outcome's levels (`.pred_Yes`, `.pred_No`), so check the names with `names()`, and check that the predictions were added to the data.

## When you are stuck

1. **Read the message**, from the end, and look at the line it points to.
2. **Restart R and run the script from the top** (*Session > Restart R*). Many mysterious errors come from objects left over from earlier work.
3. **Check your data** with `str()`, `glimpse()`, or `summary()`: wrong types and unexpected missing values cause most errors in real analyses.
4. **Make the problem small.** Reproduce it with a few rows of data or a tiny example, like those at the start of every chapter. Often the cause becomes obvious; if not, a small example is what others need to help you. The reprex package formats such an example for sharing.
5. **Search for the message** in quotation marks, or ask an AI assistant, including the code, the full message, and a description of your data (Chapter 17). Check the answer before trusting it.
6. **Ask a person**: a colleague, your supervisor, the Posit Community forum, or Stack Overflow. A clear question with a small example usually gets a quick answer.
